### 2020-01-27 cactus_maple.tex
initial

\documentclass{standalone}
\title{}
\date{}
\author{}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{algorithm}
\usepackage[noend]{algorithmic}
\usepackage{tikz,pgfplotstable,pgfplots}
\usetikzlibrary{arrows}
\begin{document}

\begin{tikzpicture}
\begin{axis}[
	xlabel={Instances Solved},
	ylabel={Run Time (sec)},width=18cm,height=10cm,legend pos=north west,xmin=0,ymin=0,xmax=600,ymax=5500
]

\addplot table [x=index, y=original_time, col sep=comma] {results_maplelcmdist_data.csv};
\addplot table [x=index, y=original_time, col sep=comma] {results_maplelcmdist_chrono_data.csv};
\addplot table [x=index, y=original_time, col sep=comma] {results_maplelcmdist_trail_noforce_data.csv};
\addplot table [x=index, y=original_time, col sep=comma] {results_maplelcmdist_chrono-dl_data.csv};
\addplot table [x=index, y=original_time, col sep=comma] {results_maplelcmdist_chrono-dl_ncte-true-opt_data.csv};

\legend{maplelcmdist,maplelcmdist-chrono,maplelcmdist-trail,maplelcmdist-chrono-dl,maplelcmdist-trail-dl}
\end{axis}
\end{tikzpicture}

\end{document}

### 2020-01-27 report.tex
initial

\documentclass{article}
\usepackage{times}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{algorithmic}
\usepackage[ruled,linesnumbered,noend,oldcommands]{algorithm2e}
\usepackage[T1,hyphens]{url}
\usepackage{xspace}
\usepackage{graphicx}
\usepackage{listings}
\lstset { %
    language=C++,
    %backgroundcolor=\color{black!5}, % set backgroundcolor
    basicstyle=\footnotesize,% basic font setting
}

\begin{document}

\begin{abstract}
\end{abstract}

\section{Introduction}
The vast majority of modern SAT solvers used in industrial applications are based on the conflict-driven clause learning (CDCL) algorithm. In a CDCL SAT solver, backtracking occurs after every conflict, where all literals from one or more decision levels become unassigned, before the solver resumes making decisions and performing unit propagations. It turns out that many of the literals that are unassigned during a backtrack will often be re-assigned again in roughly the same order when the solver continues. This results in many of the exact same propagation steps being repeated over and over again during the execution of the solver, and avoiding these repetitions could result in a significant speedup.

Chronological backtracking \cite{DBLP:conf/lpar/2013, DBLP:conf/sat/MohleB19, DBLP:conf/sat/NadelR18} addresses this inefficiency by heuristically backtracking fewer levels up the trail after learning a clause. However, chronological backtracking breaks some traditional invariants of the SAT solver, namely that literals are supposed to appear on the trail in monotonically increasing order by decision level, and can be complex to implement. Additionally, it does not allow the search to have as much flexibility in deciding the order of literals to assign, since leaving a large portion of the trail intact "in a sense" forces decisions originally used to construct that portion to be made. Some empirical evidence also suggests that it is only beneficial when used sparingly (e.g., when the number of levels to be backtracked over is greater than 100).

We propose an alternative method, "trail saving", where we cache the part of the trail that is unassigned during a backtrack and then attempt to restore implied parts of it as each new decision is made. Trail saving preserves the traditional invariants of the SAT solver and its basic version is very simple to implement. It also allows the search to choose the order of decisions, but helps make propagation faster. We explore the theoretical speedup that trail saving provides, develop some enhancements to make the idea more effective, and demonstrate experimentally that it speeds up state-of-the-art SAT solvers by a non-trivial factor.

\section{Background}
We will assume the reader is familiar with SAT solving and the basics of the CDCL algorithm \cite{DBLP:series/faia/SilvaLM09}. We will use the notion of a trail to represent a sequence of literals corresponding to the current partial assignment. The literals appear on the trail in the order in which they were assigned. Associated with every literal on the trail is a decision level. Decision level 0 represents all literals which are implied by the formula itself and subsequent decisions start the beginning of a new decision level. When a literal is made true by unit propagation, the literal is given its decision level and a reason clause, where the reason clause is the clause that was made unit by propagation in order to imply this literal. Most modern solvers use the two-literal watch scheme to make unit propagation more efficient.

\section{Trail Saving}
It can be observed that a SAT solver often repeats the same propagations over and over again during execution. Although no two descents down the trail will ever be identical because of clause learning, the solver will often propagate many of the same literals with the same reasons in a similar order to what it had done previously, especially since the order of the decisions usually remains similar. We plan to cache the part of the trail that we are backtracking over and use it to restore some of those literals without propagating them from scratch.

\subsection{Storing the trail cache}
The first step that is needed to use "trail saving" is to store the literals which are being unassigned during a backtrack, along with their respective reason clauses, in a "trail cache". Decisions should be marked with a null reason. The literals should appear in the same order on the trail cache as they had appeared on the trail. We will adopt the notation that the "top" of the trail cache is initially set to be the literal that appeared the earliest on the trail and will get updated as the trail cache gets used up. Note that initially the top of the trail cache is always a decison (i.e., the start of a new decision level). As the literals from the trail cache get restored (i.e., assigned), the top of the trail cache points to the next literal not yet restored.

\subsection{Restoring the trail cache}
We will base trail saving on a convenient observation. If the top of the trail cache becomes true at any point, every literal after the top of the trail cache up until the next decision literal (i.e., the next literal with a null reason) is implied by the current trail. The top of the trail cache can become true either by decision or unit propagation, and this observation still holds.\newline

\textit{Proof} Let $L_0$ be the set of literals on the trail after backtracking, $L_1$ be the set of literals on the current trail, and $L_c[i]$ denote the set of the first i literals on the trail cache. The $i$th literal on the trail cache is implied by $L_0 \cup L_c[i-1]$, as long as the $i$th literal was not a decision. $L_0 \subseteq L_1$, since $L_1$ starts out as $L_0$ and adds non-contradictory literals (by decision or unit propagation). Therefore, the $i$th literal on the trail cache is implied by $L_1 \cup L_c[i-1]$. Assume the top of the trail cache is true. Then, from above, every literal up until the next decision would be implied. $\blacksquare$\newline

Once the backtrack has been completed and the solver resumes with decisions and propagations, we will attempt to restore parts of the trail cache automatically during propagation to make it faster. When a literal is to be propagated, we first check the top of the trail cache to determine its truth value. If that literal's truth value is true, we assign all of the literals that come after it on the trail cache to be true up until (and excluding) the next the next literal with a null reason. If at any time the top of the trail cache is a literal with a non-null reason whose truth value is already false, we determine a conflict has occurred with the conflict clause being that literal's reason. At the beginning of a new backtrack, the trail cache is cleared and a new trail cache is generated.

A potentially quadratic speedup can be realized during the propagation of these literals since they are being propagated as a set, rather than one at a time. We will illustrate this with an example.\newline

\textit{Example 1} Suppose the literals $x_1, ... x_n$ are on the trail cache, with none of them having a null reason, and $x_1$ becomes true (e.g., the solver makes a decision for $x_1$ to be true). Then, $x_2, ... x_n$ are implied by the current trail. Now we will examine the propagation of a clause c which consists of the literals $l, \lnot x_1, \lnot x_2, ..., \lnot x_n$ with and without trail saving. Assume $l$ and $\lnot x_1$ are the current watched literals.

Without trail saving, $\lnot x_1$ becomes false and a new watcher is searched for and found 1 literal later, $\lnot x_2$, and then $\lnot x_1$ and $\lnot x_2$ are swapped. Next $\lnot x_2$ becomes false, a new watcher is searched for and found 2 literals later, $\lnot x_3$, and then $\lnot x_2$ and $\lnot x_3$ are swapped. This process will continue until all $x_n$ literals are false, c becomes unit, and $l$ becomes forced to be true. This propagation took $1 + 2 + ... + n$ searches through clause c, for a total of $O(n^2)$ searches.

With trail saving, $x2, ... x_n$ are all assigned to be true as soon as $x_1$ is made true, so immediately the values of literals $\lnot x_1, ..., \lnot x_n$ in c are false. During propagation, a new watcher is searched for to replace $\lnot x_2$, n literals are saerched over without finding any, thus c is detected as unit and $l$ becomes forced to be true. This propagation took $O(n)$ searches through clause c. $\blacksquare$\newline

Another potential speed up occurs when trail saving detects conflicts early without performing any of the pending propagations. This happens when the top of the trail cache is a literal with a non-null reason and the truth value of this literal is already false, and we can use that literal's reason clause as the conflict clause.

\subsection{Enhancements}

The first enhancement to trail saving is to keep the trail cache intact instead of clearing it before each backtrack, and to add the literals that will be backtracked over to the already existing trail cache. In order to preserve soundness, we must add the literals from the current backtrack to the beginning of the trail cache, such that the trail cache has as its prefix the literals which are being backtracked over in the same order they appear on the current trail, and as its suffix the previous trail cache. In order for the suffix to be of any use, one must actually not store the last level (i.e. the conflicting level) to the trail cache at all, since the learned clause will prevent the conflicting level from being duplicated in the future. Because the trail cache can now grow indefinitely, occasionally it needs to be trimmed by removing duplicate literals. We decided to trim the trail cache whenever its size exceeded the total number of variables.

The next enhancement to trail saving is to scan the trail cache to check if there is a literal that is already assigned false within some limit. The limit we found worked best is to scan the trail until the third null reason is found (i.e. "looking ahead 2 decisions"). If a literal that is already assigned false is found, we know that we are guaranteed to have a conflict in the near future, and we force the SAT solver's subsequent decisions to come from the trail cache instead of using the regular variable selection heuristic (i.e. EVSIDS).

\section{Related Work}
\subsection{Trail Saving for Assumption-Based SAT}
Trail saving was briefly introduced in the context of assumption-based SAT \cite{DBLP:conf/sat/HickeyB19} in order to speed up the propagation of the assumptions each time the solver backtracks past the assumption levels. When the assumption-based SAT solver backtracks past the assumption levels (i.e., during a restart or whenever a unit clause is learned), the solver will have to repeat much of the work it had previously done when it propagates through the assumptions again. In fact, the set of literals whose truth value is determined by the level 0 implicants and the assumption literals can only grow throughout the remainder of the solving episode, unless a conflict occurs.

In order to take advantage of this fact, we use "trail saving" to store all of the literals assigned during the assumption levels in a cache along with their reason clauses. Before propagating the assumption levels again, this cache is accessed in the same order as it had previously appeared on the trail to assign all of the previous propagants automatically. One must be careful to check that the truth value of the literal is not already assigned before assigning it; if the truth value of the literal is already true then there is no point re-assigning it, and if the truth value of the literal is already false then there is a conflict.

\subsection{Comparison to Chronological Backtracking}
As mentioned earlier, chronological backtracking also achieves the theoretical speedups that trail saving achieves. One advantage of chronological backtracking is that it does not require the caching of the trail or accessing this cache to "restore". It simply leaves a vast portion of the trail in tact rather than backtracking all the way to the conflict level. Chronological backtracking will also speed up propagations slightly more, because it leaves more literals assigned at once which leads to clauses being detected as unit more quickly. One disadvantage that chronological backtracking has compared to trail saving is that it doesn't allow the search as much flexibility. Because chronological backtracking leaves a large portion of the trail assigned, it will behave similarly to how the solver would behave if it was to force the decisions that do occur within this large portion of literals, whereas we do not force any decisions during trail saving. Another disadvantage is that, because propagation is not performed at every decision level, literals that are already assigned do not have a chance to move their decision levels up and this could potentially effect the lbd score or length of the learnt clause that gets generated at the next conflict. An example of this is when the solver was supposed to backtrack over a literal that would end up being unit propagated at level 0, but stays assigned at its current decision level when chronological backtracking is used. Additionally, chronological backtracking breaks the invarant that the all literals appear on the trail in monotonically increasing order by decision level, which makes its implementation slightly more complicated.

\section{Experiments and Results}
We implemented our techniques in two different SAT solvers, MapleSAT and Cadical, both of which have finished at or near the top of SAT competitions for the past several years. Each of these solvers was run as is without any trail saving or chronological backtracking in one version, with chronological backtracking in another version, and with the various types of trail saving we described in the remaining versions.

We then ran each solver on the 400 benchmarks used in the 2018 SAT Competition and the 400 benchmarks used in the SAT 2019 Competition. The experiments were run on 2.7 GHz Intel cores with each process given 7 GB of memory and 5000 seconds of CPU time. We chose not to output or verify the proofs generated by any of the solvers. The results are reported in Figure 1.

When comparing regular cadical with non-chronological backtracking against cadical with trail saving, the average time taken to solve those instances that were solved by both solvers decreased and the number of propagations per second increased. Although the total number of solved instances decreased for both versions of trail saving, the Par-2 score increased for both of them. The trail saving versions also solved more instances and had better Par-2 scores than cadical with chronological backtracking (default parameters).

When comparing regular MapleSAT with non-chronological backtracking against either of the MapleSAT trail saving versions, the total number of instances solved increased significantly, the average time taken to solve those instances which were solved by both solvers decreased, and the number of propagations per second increased. When comparing the trail saving versions to MapleSAT with chronological backtracking (default parameters), the trail saving versions perform slightly better in both total instances solved and PAR-2 scores.

\begin{figure}
\begin{tabular}{|l|l|l|l|l|}
\hline
                 & cadical                     & cadical\_chrono & cadical\_trail & cadical\_trail\_enhanced \\ \hline
Instances Solved & 527                         & 515             & 520            & 524                      \\ \hline
Par-2 (x $10^6$s)& 3.220                  & 3.269      & 3.248     & 3.216               \\ \hline
                 & maple                       & maple\_chrono   & maple\_trail   & maple\_trail\_enhanced          \\ \hline
Instance Solved  & 458                         & 470             & 472            & 473                      \\ \hline
Par-2 (x $10^6$s)& 3.797 & 3.690      & 3.694     & 3.667               \\ \hline
\end{tabular}
\caption{Table of results}
\end{figure}
\begin{figure}\includegraphics[scale=0.8]{cactus_cadical.pdf}\caption{\small{Comparison of run times for versions of cadical. cadical is without chronological backtracking, cadical-chrono is with chronological backtracking, cadical-trail is with the first version of trail saving, cadical-trail-2levels-ahead is the trail saving with all enhancements added.}}\end{figure}
\begin{figure}\includegraphics[scale=0.8]{cactus_maple2.pdf}\caption{\small{Comparison of run times for versions of MapleSAT. maplelcmdist is from the 2017 SAT Competition, maplelecmdist-chrono is with chronological backtracking (2018 SAT Competition), maplelcmdist-trail is with the first version of trail saving, maplelcmdist-trail-enhanced is with trail saving and all enhancements added.}}\end{figure}
\iffalse
\begin{figure}\includegraphics[scale=0.8]{cactus_maple.pdf}\caption{\small{Comparison of run times for versions of MapleSAT. maplelcmdist is from the 2017 SAT Competition, maplelecmdist-chrono is with chronological backtracking (2018 SAT Competition), maplelcmdist-trail is with the first version of trail saving, maplelcmdist-chrono-dl is with chronological backtracking and duplicate learnts (2019 SAT Competition), maplelcmdist-trail-dl is with trail saving and duplicate learnts}}\end{figure}
\fi
\clearpage

\iffalse
\begin{figure}[t]
\begin{center}
    \fss{9pt}{10pt}
	\setlength\tabcolsep{1pt}
	\begin{tabular}{| p{4.2cm} | c | c | c | c | c | c | c | c |}
	  \hline
	  & \multicolumn{4}{|c|}{Total} & \multicolumn{2}{|c|}{Unweighted} & \multicolumn{2}{|c|}{Weighted} \\ \hline
	  & MaxHS & +/- & \hspace{1pt} RC2 \hspace{1pt} & +/- & MaxHS & \hspace{1pt} RC2 \hspace{1pt} & MaxHS & \hspace{1pt} RC2 \hspace{1pt} \\ \hline
	  original & 6052 & 0/0 & 6030 & 0/0 & 3940 & 3993 & 2112 & 2037 \\ \hline
	  enqueue assumptions as set & 6131 & 114/35 & 6081 & \hspace{0.5pt} 99/48 \hspace{0.5pt} & 3995 & 4032 & 2136 & 2049 \\ \hline
	  enqueue assumptions as set + \newline save literals after learnt units & 6136 & 120/36 & 6079 & \hspace{0.5pt} 95/46 \hspace{0.5pt} & 3991 & 4030 & 2145 & 2049 \\ \hline
	  enqueue assumptions as set + \newline save literals after learnt units + \newline save literals from last invocation & 6138 & 115/29 & 6080 & \hspace{0.5pt} 92/42 \hspace{0.5pt} & 3989 & 4030 & 2149 & 2050 \\ \hline
	\end{tabular}
\end{center}
\caption{Number of \maxsat instances solved by MaxHS and RC2 using
  different extensions of the underlying \sat solver. The +/- column shows
  the number of instances gained/lost vs the original.}
\label{fig:nsolved}
\end{figure}
\fi

\section{Conclusion}

\bibliography{ref}{}
\bibliographystyle{splncs04}
\end{document}

### 2020-01-27 Solver.cc

@@ -1,0 +1,2426 @@
+/***************************************************************************************[Solver.cc]
+ MiniSat -- Copyright (c) 2003-2006, Niklas Een, Niklas Sorensson
+ Copyright (c) 2007-2010, Niklas Sorensson
+ 
+ Chanseok Oh's MiniSat Patch Series -- Copyright (c) 2015, Chanseok Oh
+ 
+ Maple_LCM, Based on MapleCOMSPS_DRUP -- Copyright (c) 2017, Mao Luo, Chu-Min LI, Fan Xiao: implementing a learnt clause minimisation approach
+ Reference: M. Luo, C.-M. Li, F. Xiao, F. Manya, and Z. L. , “An effective learnt clause minimization approach for cdcl sat solvers,” in IJCAI-2017, 2017, pp. to–appear.
+ 
+ Maple_LCM_Dist, Based on Maple_LCM -- Copyright (c) 2017, Fan Xiao, Chu-Min LI, Mao Luo: using a new branching heuristic called Distance at the beginning of search
+ 
+ MapleLCMDistChronoBT, based on Maple_LCM_Dist -- Copyright (c), Alexander Nadel, Vadim Ryvchin: "Chronological Backtracking" in SAT-2018, pp. 111-121.
+ 
+ MapleLCMDistChronoBT-DL, based on MapleLCMDistChronoBT -- Copyright (c), Stepan Kochemazov, Oleg Zaikin, Victor Kondratiev, Alexander Semenov: The solver was augmented with heuristic that moves duplicate learnt clauses into the core/tier2 tiers depending on a number of parameters.
+ 
+ Permission is hereby granted, free of charge, to any person obtaining a copy of this software and
+ associated documentation files (the "Software"), to deal in the Software without restriction,
+ including without limitation the rights to use, copy, modify, merge, publish, distribute,
+ sublicense, and/or sell copies of the Software, and to permit persons to whom the Software is
+ furnished to do so, subject to the following conditions:
+ 
+ The above copyright notice and this permission notice shall be included in all copies or
+ substantial portions of the Software.
+ 
+ THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR IMPLIED, INCLUDING BUT
+ NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND
[…]

### 2020-01-27 Solver_1.cc

@@ -1,0 +1,2577 @@
+/***************************************************************************************[Solver.cc]
+MiniSat -- Copyright (c) 2003-2006, Niklas Een, Niklas Sorensson
+           Copyright (c) 2007-2010, Niklas Sorensson
+ 
+Chanseok Oh's MiniSat Patch Series -- Copyright (c) 2015, Chanseok Oh
+ 
+Maple_LCM, Based on MapleCOMSPS_DRUP -- Copyright (c) 2017, Mao Luo, Chu-Min LI, Fan Xiao: implementing a learnt clause minimisation approach
+Reference: M. Luo, C.-M. Li, F. Xiao, F. Manya, and Z. L. , “An effective learnt clause minimization approach for cdcl sat solvers,” in IJCAI-2017, 2017, pp. to–appear.
+
+Maple_LCM_Dist, Based on Maple_LCM -- Copyright (c) 2017, Fan Xiao, Chu-Min LI, Mao Luo: using a new branching heuristic called Distance at the beginning of search
+
+MapleLCMDistChronoBT, based on Maple_LCM_Dist -- Copyright (c), Alexander Nadel, Vadim Ryvchin: "Chronological Backtracking" in SAT-2018, pp. 111-121.
+
+MapleLCMDistChronoBT-DL, based on MapleLCMDistChronoBT -- Copyright (c), Stepan Kochemazov, Oleg Zaikin, Victor Kondratiev, Alexander Semenov: The solver was augmented with heuristic that moves duplicate learnt clauses into the core/tier2 tiers depending on a number of parameters.
+
+Permission is hereby granted, free of charge, to any person obtaining a copy of this software and
+associated documentation files (the "Software"), to deal in the Software without restriction,
+including without limitation the rights to use, copy, modify, merge, publish, distribute,
+sublicense, and/or sell copies of the Software, and to permit persons to whom the Software is
+furnished to do so, subject to the following conditions:
+
+The above copyright notice and this permission notice shall be included in all copies or
+substantial portions of the Software.
+
+THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR IMPLIED, INCLUDING BUT
+NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND
+NONINFRINGEMENT. IN NO EVENT SHALL THE AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM,
[…]

### 2020-01-27 parse.py

@@ -1,0 +1,42 @@
+import os
+import math
+import sys
+
+f = open(sys.argv[1], 'r')
+os.system('rm ' + sys.argv[2])
+f0 = open(sys.argv[2], 'w')
+f1 = open('2' + sys.argv[2], 'w')
+lines = f.read().split('\r')
+inter = {}
+incr = 0.2
+extreme_neg = 0
+extreme_pos = 0
+intervals = [-2.0, -1.8, -1.6, -1.4, -1.2, -1.0, -0.8, -0.6, -0.4, -0.2, 0.0, 0.2, 0.4, 0.6, 0.8, 1.0, 1.2, 1.4, 1.6, 1.8, 2.0]
+total = 0
+for i in intervals:
+	inter[i] = 0
+
+for l in lines[1:]:
+	a, b = (float)(l.split('\t')[0]), (float)(l.split('\t')[1])
+	data = math.log((a+1e-8) / (b+1e-8), 2.0)
+	if data >= intervals[0] - incr / 2.0 and data < intervals[len(intervals)-1] + incr / 2.0:
+		for i in intervals:
+			if data >= i - incr / 2.0 and data < i + incr / 2.0:
+				inter[i] += 1
+				total += 1
+				break
+	elif data < intervals[0] - incr / 2.0:
+		extreme_neg += 1
+	elif data >= intervals[len(intervals)-1] + incr / 2.0:
+		extreme_pos += 1
+
+for i in intervals:
+	if i < 0:
+		f0.write((str)(-i) + '\t' + (str)(inter[i]) + '\n')
+	elif i > 0:
+		f1.write((str)(i) + '\t' + (str)(inter[i]) + '\n')
+
+print inter[0], ' ', extreme_neg, ' ', extreme_pos, ' ', total
+f.close()
+f0.close()
+f1.close()

### 2020-01-27 report.tex
initial

@@
 \subsection{Trail Saving for Assumption-Based SAT}
-Trail saving was briefly introduced in the context of assumption-based SAT \cite{DBLP:conf/sat/HickeyB19} in order to speed up the propagation of the assumptions each time the solver backtracks past the assumption levels.
-When the assumption-based SAT solver backtracks past the assumption levels (i.e., during a restart or whenever a unit clause is learned), the solver will have to repeat much of the work it had previously done when it propagates through the assumptions again.
-In fact, the set of literals whose truth value is determined by the level 0 implicants and the assumption literals can only grow throughout the remainder of the solving episode, unless a conflict occurs.
-
-In order to take advantage of this fact, we use "trail saving" to store all of the literals assigned during the assumption levels in a cache along with their reason clauses.
-Before propagating the assumption levels again, this cache is accessed in the same order as it had previously appeared on the trail to assign all of the previous propagants automatically.
-One must be careful to check that the truth value of the literal is not already assigned before assigning it; if the truth value of the literal is already true then there is no point re-assigning it, and if the truth value of the literal is already false then there is a conflict.
+Trail saving was briefly introduced in the context of assumption-based SAT \cite{DBLP:conf/sat/HickeyB19} in order to speed up the propagation of the assumptions each time the solver backtracked past the assumption levels.
+Note that the process is the same, but it only works to restore the propagants and implicants of the assumptions and only after a learnt unit clause.
+Becuase the number of propagants and implicants of the assumptions tends to become dominated by the number of assumptions in the applications it was tested on, the technique did not yield a significant improvement.
+It would be interesting to combine the extended version of trail saving outlined in this paper with trail saving over the assumptions to see if that could produce better results.
 
 \subsection{Comparison to Chronological Backtracking}
-As mentioned earlier, chronological backtracking also achieves the theoretical speedups that trail saving achieves.
-One advantage of chronological backtracking is that it does not require the caching of the trail or accessing this cache to "restore".
-It simply leaves a vast portion of the trail in tact rather than backtracking all the way to the conflict level.
-Chronological backtracking will also speed up propagations slightly more, because it leaves more literals assigned at once which leads to clauses being detected as unit more quickly.
+As mentioned earlier, chronological backtracking also achieves some of the theoretical speedups that trail saving achieves.
+One advantage of chronological backtracking is that it does not require any caching of the trail or accessing such a cache during propagation; it simply leaves a vast portion of the trail intact rather than backtracking all the way to the conflict level.
+Chronological backtracking will also speed up propagations slightly more than trail saving, because it leaves more literals assigned at once which leads to clauses being detected as unit more quickly.
 One disadvantage that chronological backtracking has compared to trail saving is that it doesn't allow the search as much flexibility.
-Because chronological backtracking leaves a large portion of the trail assigned, it will behave similarly to how the solver would behave if it was to force the decisions that do occur within this large portion of literals, whereas we do not force any decisions during trail saving.
+Because chronological backtracking leaves a large portion of the trail assigned, it will behave similarly to how the solver would behave if it was to force the decisions that do occur within this large portion of literals, whereas trail saving does not force any decisions.
 Another disadvantage is that, because propagation is not performed at every decision level, literals that are already assigned do not have a chance to move their decision levels up and this could potentially effect the lbd score or length of the learnt clause that gets generated at the next conflict.
 An example of this is when the solver was supposed to backtrack over a literal that would end up being unit propagated at level 0, but stays assigned at its current decision level when chronological backtracking is used.
-Additionally, chronological backtracking breaks the invarant that the all literals appear on the trail in monotonically increasing order by decision level, which makes its implementation slightly more complicated.
+Additionally, chronological backtracking breaks the invarant that all literals appear on the trail in monotonically increasing order by decision level, which makes its implementation slightly more complicated.
 

### 2020-01-27 report.tex
initial

@@
 However, chronological backtracking breaks some traditional invariants of the SAT solver, namely that literals are supposed to appear on the trail in monotonically increasing order by decision level, and can be complex to implement.
-Additionally, it does not allow the search to have as much flexibility in deciding the order of literals to assign, since leaving a large portion of the trail intact "in a sense" forces decisions originally used to construct that portion to be made.
+Additionally, it does not allow the search to have as much flexibility in deciding the order of literals to assign, since leaving a large portion of the trail intact "in a sense" forces decisions originally used to construct that portion.
 Some empirical evidence also suggests that it is only beneficial when used sparingly (e.g., when the number of levels to be backtracked over is greater than 100).
@@
 Chronological backtracking will also speed up propagations slightly more than trail saving, because it leaves more literals assigned at once which leads to clauses being detected as unit more quickly.
-One disadvantage that chronological backtracking has compared to trail saving is that it doesn't allow the search as much flexibility.
+
+One disadvantage that chronological backtracking has compared to trail saving is that it doesn't allow the search as much flexibility to jump out of local neighbourhoods, as the common variable selection heuristics tend to depend on \cite{DBLP:conf/dac/MoskewiczMZZM01, DBLP:conf/sat/2015, DBLP:conf/sat/LiangGPC16}.
 Because chronological backtracking leaves a large portion of the trail assigned, it will behave similarly to how the solver would behave if it was to force the decisions that do occur within this large portion of literals, whereas trail saving does not force any decisions.
-Another disadvantage is that, because propagation is not performed at every decision level, literals that are already assigned do not have a chance to move their decision levels up and this could potentially effect the lbd score or length of the learnt clause that gets generated at the next conflict.
+As variable selection heuristics become more refined, it is likely that adhering to them will become more crucial to performance.
+
+Another disadvantage is that, because propagation is not performed at every decision level, literals that remain assigned during chronological backtracking (but otherwise would have been unassigned) do not have a chance to move their decision levels up.
 An example of this is when the solver was supposed to backtrack over a literal that would end up being unit propagated at level 0, but stays assigned at its current decision level when chronological backtracking is used.
+This could potentially effect the lbd score or length of the learnt clause that gets generated at the next conflict.
 Additionally, chronological backtracking breaks the invarant that all literals appear on the trail in monotonically increasing order by decision level, which makes its implementation slightly more complicated.

### 2020-01-27 parse.py

@@ -17,7 +17,7 @@ for i in intervals:
 	inter[i] = 0
 
 for l in lines[1:]:
-	a, b = (float)(l.split('\t')[0]), (float)(l.split('\t')[1])
+	a, b = (float)(l.split(',')[0]), (float)(l.split(',')[1])
 	data = math.log((a+1e-8) / (b+1e-8), 2.0)
 	if data >= intervals[0] - incr / 2.0 and data < intervals[len(intervals)-1] + incr / 2.0:
 		for i in intervals:

### 2020-01-27 report.tex
figures
alg

@@
 \usepackage{amssymb}
-\usepackage{algorithmic}
+\usepackage[noend]{algcompatible}
 \usepackage[ruled,linesnumbered,noend,oldcommands]{algorithm2e}
@@
 When a literal is made true by unit propagation, the literal is given its decision level and a reason clause, where the reason clause is the clause that was made unit by propagation in order to imply this literal.
-Most modern solvers use the two-literal watch scheme to make unit propagation more efficient.
+Most modern solvers use the two-literal watch scheme to make unit propagation more efficient, where each clause is watched by two non-false literals.
+When a watcher for a clause becomes false, a replacement is searched for in the clause.
+If no such replacement exists
 
@@
 We plan to cache the part of the trail that we are backtracking over and use it to restore some of those literals without propagating them from scratch.
+
+	\begin{algorithm}
+  	\caption{propagate()}
+	\begin{algorithmic}[1]
+	\WHILE{queue not empty}
+	\STATE $l := $ next literal on queue
+	\\BEGIN CODE TO ADD FOR TRAIL SAVING
+	\IF{trail saving turned on}
+		\IF{val(top of trail cache) $=$ true}
+			\WHILE{trail cache not empty}
+				\STATE{move top of trail cache to next literal}
+				\STATE{let tc $=$ top of trail cache}
+				\IF{reason(tc) is null}
+					\STATE{break}
+				\ELSIF{val(tc) $=$ false}
+					\STATE{return conflict with reason(tc)}
+				\ELSIF{val(tc) $=$ unassigned}
+					\STATE{enqueue tc with reason reason(tc)}
+				\ENDIF
+			\ENDWHILE
+		\ENDIF
+	\ENDIF
+	\\END CODE TO ADD FOR TRAIL SAVING
+  	\FOR{each clause $c \in $ watchlist($\neg l$)}
+		\FOR{each literal $x \in c$}
+			\IF{$val($x$) \neq$ false and x is not already a watcher of c}
+				\STATE let $x$ be a watcher of $c$ and add $c$ to watchlist($\neg x$)
+				\STATE break
+			\ENDIF
+		\ENDFOR
+		\IF{no new watcher was found}
+			\IF{other watcher $x_2$ of $c = $ false}
+				\STATE return c as conflict
+			\ELSIF{other watcher $x_2$ of $c$ is unassigned}
+				\STATE enqueue $x_2$
+			\ENDIF
+		\ENDIF
+	\ENDFOR
+	\ENDWHILE
+	\end{algorithmic}
+	\end{algorithm}
+\clearpage
 
@@
 maplelcmdist is from the 2017 SAT Competition, maplelecmdist-chrono is with chronological backtracking (2018 SAT Competition), maplelcmdist-trail is with the first version of trail saving, maplelcmdist-trail-enhanced is with trail saving and all enhancements added.}}\end{figure}
+\begin{figure}\includegraphics[scale=0.8]{test.pdf}\caption{}\end{figure}
+\begin{figure}\includegraphics[scale=0.8]{test2.pdf}\caption{}\end{figure}
 \iffalse

### 2020-02-10 report.tex
some edits

@@
 \section{Introduction}
-The vast majority of modern SAT solvers used in industrial applications are based on the conflict-driven clause learning (CDCL) algorithm.
+The vast majority of modern SAT solvers that are used to solve real-world problems are based on the conflict-driven clause learning (CDCL) algorithm.
 In a CDCL SAT solver, backtracking occurs after every conflict, where all literals from one or more decision levels become unassigned, before the solver resumes making decisions and performing unit propagations.
-It turns out that many of the literals that are unassigned during a backtrack will often be re-assigned again in roughly the same order when the solver continues.
-This results in many of the exact same propagation steps being repeated over and over again during the execution of the solver, and avoiding these repetitions could result in a significant speedup.
+Traditionally, CDCL solvers would backtrack non-chronologically to the conflict level, which is the second highest decision level remaining in the conflict clause after conflict analysis has resolved away all but one literal from the current decision level \cite{DBLP:conf/dac/MoskewiczMZZM01}.
+Recently, however, it has been shown that backtracking chronologically (i.e.
+backtracking one level only after conflict analysis) \cite{DBLP:conf/lpar/2013, DBLP:conf/sat/NadelR18} can be effective on many instances and was implemented in solvers that won the last two SAT competitions.
+Although chronological backtracking breaks some of the conventional invariants of CDCL solvers, it has been formalized and proven correct \cite{DBLP:conf/sat/MohleB19}.
 
-Chronological backtracking \cite{DBLP:conf/lpar/2013, DBLP:conf/sat/MohleB19, DBLP:conf/sat/NadelR18} addresses this inefficiency by heuristically backtracking fewer levels up the trail after learning a clause.
-However, chronological backtracking breaks some traditional invariants of the SAT solver, namely that literals are supposed to appear on the trail in monotonically increasing order by decision level, and can be complex to implement.
-Additionally, it does not allow the search to have as much flexibility in deciding the order of literals to assign, since leaving a large portion of the trail intact "in a sense" forces decisions originally used to construct that portion.
-Some empirical evidence also suggests that it is only beneficial when used sparingly (e.g., when the number of levels to be backtracked over is greater than 100).
+There are at least two possible side-effects of chronological backtracking which may explain its effectiveness.
+It has been observed that when a solver backtracks non-chronologically after a conflict, many of the literals that are unassigned during the backtrack will be re-assigned again in roughly the same order when the solver continues.
+By using chronological backtracking, the solver keeps a lot more of its partial assignment intact and saves from having to reconstruct a lot of the trail via propagation that it otherwise would have.
+Saving from having to reconstruct portions of the trail after backtracking was first introduced in the context of restarts \cite{DBLP:journals/jsat/TakRH11}.
+The second side-effect of chronological backtracking is that it forces the solver to stay in the same local neighbourhood during search, since it unassigns less of the partial assignment and one could argue that this might lead to finding additional conflicts more quickly or reaching a satisfying assignment (if one exists) more quickly.
+This second side-effect also leads to the search with chronological backtracking being less flexibile to jump out of local neighbourhoods and obey the variable selection heuristic.
+We argue that this second side-effect is not the primary explanation for the effectiveness of chronological backtracking, and aim to to develop a technique which will produce a benefit similar to the first side-effect while maintaining the freedom for the solver to jump out of local neighbourhoods and obey the variable selection heuristic.
 
-We propose an alternative method, "trail saving", where we cache the part of the trail that is unassigned during a backtrack and then attempt to restore implied parts of it as each new decision is made.
+Our alternative method is "trail saving", where we cache the part of the trail that is unassigned during a backtrack and then attempt to restore implied parts of it as each new decision is made.
 Trail saving preserves the traditional invariants of the SAT solver and its basic version is very simple to implement.
@@
 When a watcher for a clause becomes false, a replacement is searched for in the clause.
-If no such replacement exists
+If no such replacement exists and the other watcher is true or unassigned, then the clause is unit.
+If no such replacement exists and the other watcher is also false, then the clause is a conflict and needs to be analyzed to produce a learnt clause.
 
@@
 
+\subsection{Storing the trail cache}
+
+The first step that is needed to use "trail saving" is to store all of the literals which are being unassigned during a backtrack, along with their respective reason clauses, in a "trail cache".
+Decisions should be marked with a null reason.
+The literals should appear in the same order on the trail cache as they had appeared on the trail.
+We will keep a pointer, which we will refer to as the "top" of the trail cache, and initially set it to be the literal which had appeared the earliest on the trail.
+Note that initially the top of the trail cache always has a null reason (i.e., it was the start of a new decision level on the previous trail).
+As the literals from the trail cache get restored (i.e., assigned), the top of the trail cache will be updated to the next literal not yet restored.
+
+\subsection{Restoring the trail cache}
+We will base trail saving on the following critical invariant.
+If the top of the trail cache becomes true at any point, even after an arbitrary number of decisions and propagations have been made since the last backtrack, the set of literals on the trail cache after the top and up to but excluding the next literal with a null reason is still implied by the current trail.
+The top of the trail cache can become true either by decision or unit propagation, and this invariant still holds.\newline
+
+\textit{Proof 1} We will refer to the set of literals that become unassigned during the backtrack after conflict analysis has taken place as $L_u$, and will refer to the rest of the trail that remains assigned as $L_0$.
+We will refer to all of the literals that get placed on the trail after $L_0$ during subsequent propagations and decisions as $L_1$.
+We will refer to the current trail $L_0 \cup L_1$ as $L_T$.
+The trail cache will contain the literals in $L_u$ ordered by their appearance on the old trail.
+Let $L_c[i]$ denote the set of the first i literals on the trail cache.
+Let $L_c[top]$ be the top of the old trail, which gets updated as parts of the trail cache get restored.
+
+The $i$th literal on the trail cache is implied by $L_0 \cup L_c[i-1]$, as long as the $i$th literal has a non-null reason, since $L_0 \cup L_c[i-1]$ had forced $L_c[i]$ on the previous trail.
+$L_0 \subseteq L_T$, since $L_T = L_0 \cup L_1$.
+Therefore, the $i$th literal on the trail cache is also implied by $L_T \cup L_c[i-1]$.
+
+Assume $L_c[top]$ is true, which means that $L_c[0], ..., L_C[top] \subseteq L_1 \subseteq L_T$.
+Then, from above, every literal on the trail cache up to but excluding the next literal with a nun-null reason would be implied.
+$\blacksquare$\newline
+
+Once the backtrack has been completed and the solver resumes with decisions and propagations, we will attempt to restore parts of the trail cache automatically during propagation to make it faster by using the invariant stated above.
+Each time the propagation routine is about to propagate a new literal, we first check the top of the trail cache to determine its truth value.
+If that literal's truth value is already true, we assign all of the literals that come after it on the trail cache to be true up until (and excluding) the next literal with a null reason.
+
+A potentially quadratic speedup can be realized during the propagation of these literals since they are being propagated as a set, rather than one at a time, as argued in \cite{DBLP:conf/sat/HickeyB19, DBLP:journals/jair/Gent13}.
+Restoring literals from the trail cache automatically not only leads to quicker propagation of future literals, but also means that the propagation routine has less work to do as it searches for new watchers in clauses on watchlists.
+All of the reasons attached to these literals that are automatically restored will now be satisfied, thus the propagation routine can skip over them rather than having to access each of them potentially multiple times to find they are unit.
+
+If at any time the top of the trail cache is a literal $l$ with a non-null reason whose truth value is already false, we determine a conflict has occurred with the conflict clause being $l$'s reason from the trail cache.
+This is because we know from Proof 1 that $l$ is implied to be true at this point by the reason on the trail cache (i.e., the reason on the trail cache is unit), but $\lnot l$ was also previously assigned to be true, therefore we have a conflict.
+This leads to a third potential speed up, since a conflict has been detected and none of the pending propagations have to be completed.
+Since trail saving can sometimes save hundreds or thousands of literals at a time, the number of propagations required to detect a conflict could be very costly.
+
+\begin{figure}\includegraphics[scale=0.8]{trail_diagram.pdf}\caption{Storing into a trail cache after backtracking.}\end{figure}
+
 	\begin{algorithm}
-  	\caption{propagate()}
+  	\TitleOfAlgo{propagate()}
 	\begin{algorithmic}[1]
 	\WHILE{queue not empty}
+	\STATE { }
+	\STATE restoreTrailCache()
+	\STATE { }
 	\STATE $l := $ next literal on queue
-	\\BEGIN CODE TO ADD FOR TRAIL SAVING
-	\IF{trail saving turned on}
-		\IF{val(top of trail cache) $=$ true}
-			\WHILE{trail cache not empty}
-				\STATE{move top of trail cache to next literal}
-				\STATE{let tc $=$ top of trail cache}
-				\IF{reason(tc) is null}
-					\STATE{break}
-				\ELSIF{val(tc) $=$ false}
-					\STATE{return conflict with reason(tc)}
-				\ELSIF{val(tc) $=$ unassigned}
-					\STATE{enqueue tc with reason reason(tc)}
-				\ENDIF
-			\ENDWHILE
-		\ENDIF
-	\ENDIF
-	\\END CODE TO ADD FOR TRAIL SAVING
   	\FOR{each clause $c \in $ watchlist($\neg l$)}
@@
 	\end{algorithm}
-\clearpage
 
-\subsection{Storing the trail cache}
-The first step that is needed to use "trail saving" is to store the literals which are being unassigned during a backtrack, along with their respective reason clauses, in a "trail cache".
-Decisions should be marked with a null reason.
-The literals should appear in the same order on the trail cache as they had appeared on the trail.
-We will adopt the notation that the "top" of the trail cache is initially set to be the literal that appeared the earliest on the trail and will get updated as the trail cache gets used up.
-Note that initially the top of the trail cache is always a decison (i.e., the start of a new decision level).
-As the literals from the trail cache get restored (i.e., assigned), the top of the trail cache points to the next literal not yet restored.
-
-\subsection{Restoring the trail cache}
-We will base trail saving on a convenient observation.
-If the top of the trail cache becomes true at any point, every literal after the top of the trail cache up until the next decision literal (i.e., the next literal with a null reason) is implied by the current trail.
-The top of the trail cache can become true either by decision or unit propagation, and this observation still holds.\newline
-
-\textit{Proof} Let $L_0$ be the set of literals on the trail after backtracking, $L_1$ be the set of literals on the current trail, and $L_c[i]$ denote the set of the first i literals on the trail cache.
-The $i$th literal on the trail cache is implied by $L_0 \cup L_c[i-1]$, as long as the $i$th literal was not a decision.
-$L_0 \subseteq L_1$, since $L_1$ starts out as $L_0$ and adds non-contradictory literals (by decision or unit propagation).
-Therefore, the $i$th literal on the trail cache is implied by $L_1 \cup L_c[i-1]$.
-Assume the top of the trail cache is true.
-Then, from above, every literal up until the next decision would be implied.
-$\blacksquare$\newline
-
-Once the backtrack has been completed and the solver resumes with decisions and propagations, we will attempt to restore parts of the trail cache automatically during propagation to make it faster.
-When a literal is to be propagated, we first check the top of the trail cache to determine its truth value.
-If that literal's truth value is true, we assign all of the literals that come after it on the trail cache to be true up until (and excluding) the next the next literal with a null reason.
-If at any time the top of the trail cache is a literal with a non-null reason whose truth value is already false, we determine a conflict has occurred with the conflict clause being that literal's reason.
-At the beginning of a new backtrack, the trail cache is cleared and a new trail cache is generated.
-
-A potentially quadratic speedup can be realized during the propagation of these literals since they are being propagated as a set, rather than one at a time.
-We will illustrate this with an example.\newline
-
-\textit{Example 1} Suppose the literals $x_1, ...
-x_n$ are on the trail cache, with none of them having a null reason, and $x_1$ becomes true (e.g., the solver makes a decision for $x_1$ to be true).
-Then, $x_2, ...
-x_n$ are implied by the current trail.
-Now we will examine the propagation of a clause c which consists of the literals $l, \lnot x_1, \lnot x_2, ..., \lnot x_n$ with and without trail saving.
-Assume $l$ and $\lnot x_1$ are the current watched literals.
-
-Without trail saving, $\lnot x_1$ becomes false and a new watcher is searched for and found 1 literal later, $\lnot x_2$, and then $\lnot x_1$ and $\lnot x_2$ are swapped.
-Next $\lnot x_2$ becomes false, a new watcher is searched for and found 2 literals later, $\lnot x_3$, and then $\lnot x_2$ and $\lnot x_3$ are swapped.
-This process will continue until all $x_n$ literals are false, c becomes unit, and $l$ becomes forced to be true.
-This propagation took $1 + 2 + ...
-+ n$ searches through clause c, for a total of $O(n^2)$ searches.
-
-With trail saving, $x2, ...
-x_n$ are all assigned to be true as soon as $x_1$ is made true, so immediately the values of literals $\lnot x_1, ..., \lnot x_n$ in c are false.
-During propagation, a new watcher is searched for to replace $\lnot x_2$, n literals are saerched over without finding any, thus c is detected as unit and $l$ becomes forced to be true.
-This propagation took $O(n)$ searches through clause c.
-$\blacksquare$\newline
-
-Another potential speed up occurs when trail saving detects conflicts early without performing any of the pending propagations.
-This happens when the top of the trail cache is a literal with a non-null reason and the truth value of this literal is already false, and we can use that literal's reason clause as the conflict clause.
+	\begin{algorithm}
+  	\TitleOfAlgo{restoreTrailCache()}
+	\begin{algorithmic}[1]
+	\IF{trail saving turned on}
+		\STATE{let tc $=$ top of trail cache}
+		\IF{val(top of trail cache) $=$ true}
+			\WHILE{trail cache not empty}
+				\STATE{move top of trail cache (tc) to next literal}
+				\IF{reason(tc) is null}
+					\STATE{break}
+				\ELSIF{val(tc) $=$ false}
+					\STATE{return conflict with reason(tc)}
+				\ELSIF{val(tc) $=$ unassigned}
+					\STATE{enqueue tc with reason reason(tc)}
+				\ENDIF
+			\ENDWHILE
+		\ENDIF
+	\ENDIF
+	\end{algorithmic}
+	\end{algorithm}
 
@@
 
-The first enhancement to trail saving is to keep the trail cache intact instead of clearing it before each backtrack, and to add the literals that will be backtracked over to the already existing trail cache.
-In order to preserve soundness, we must add the literals from the current backtrack to the beginning of the trail cache, such that the trail cache has as its prefix the literals which are being backtracked over in the same order they appear on the current trail, and as its suffix the previous trail cache.
-In order for the suffix to be of any use, one must actually not store the last level (i.e.
-the conflicting level) to the trail cache at all, since the learned clause will prevent the conflicting level from being duplicated in the future.
-Because the trail cache can now grow indefinitely, occasionally it needs to be trimmed by removing duplicate literals.
-We decided to trim the trail cache whenever its size exceeded the total number of variables.
+In the standard version of trail saving outlined above, the trail cache needs to be deleted after each backtrack in order to make room for the new backtracked literals to be cached.
+The first enhancement to trail saving is to keep the trail cache intact instead of clearing it before each backtrack, and to add the literals that will be backtracked over to the beginning of the already existing trail cache.
+In order to preserve soundness, we must add the literals from the current backtrack to the beginning of the trail cache, such that the trail cache has as its prefix the literals which are being backtracked over, and as its suffix the previous trail cache.
+\newline
+
+\textit{Proof 2} We will use all of the same set names as Proof 1 and also refer to the set of literals on our previous trail cache as $L_{pc}$.
+We concatenate the literals in $L_{pc}$ to the end of $L_c$ to get $L_{sc}$.
+We know that $L_{pc}$ is a valid trail cache for $L_0 \cup L_u$.
+If we reach the end of $L_c$ without any conflicts, we know that $L_u \subseteq L_T$.
+Since $L_0 \subseteq L_T$, then $L_u \cup L_0 \subseteq L_T$ and therefore $L_{pc}$ is a valid trail cache for $L_T$.
+$\blacksquare$\newline
+
+In order for the previous trail cache to be of any use, one must actually never store the last level of the current trail to the trail cache at all, since a learnt clause prevents the last level from being duplicated in the future.
+If any part of the previous trail cache was used to construct this last level, it must also be discarded from the previous trail cache before the previous trail cache is concatenated to the end.
+
+When concatenating trail caches together as above, it is important to note that the running trail cache can grow indefinitely.
+The trail cache can be trimmed periodically in the following way.
+Traverse down the trail cache and whenever a literal appears more than once, delete its entry in the trail cache.
+Whenever a literal $x$ appears whose complement $\lnot x$ has already appeared on the trail, $x$ should be deleted and everything remaining until the end of the trail cache should be deleted (in order to keep the trail cache free of conflicts).
+We decided to trim the trail cache whenever its size exceeded twice the total number of variables, as that is what worked best in preliminary experiments.\newline
 
@@
 "looking ahead 2 decisions").
-If a literal that is already assigned false is found, we know that we are guaranteed to have a conflict in the near future, and we force the SAT solver's subsequent decisions to come from the trail cache instead of using the regular variable selection heuristic (i.e.
-EVSIDS).
+If a literal that is already assigned false is found, we know that we are guaranteed to have a conflict in the near future, and we force the SAT solver's subsequent decisions to come from the trail cache instead of using the regular variable selection heuristic.
+
+Sometimes using trail saving to automatically restore literals with their previous reasons will lead to a literal getting a reason that is "worse" (in terms of number of literals, but one could use a different metric) than the reason it would have ended up with under normal propagation (i.e.
+without trail saving).
+The last enhancement we made to trail saving was to stop restoring literals from the old trail as soon as we hit a reason that we suspected might be "bad".
+For this, we kept a running mean and standard deviation of the length of all reasons used in the solver so far, and decided to stop restoring from the trail whenever we hit a reason that was greater in length than two standard deviations above the mean length.
+This way, when normal propagation resumes, we give it a chance to find a better reason than the one that trail saving had stored from the last descent down the trail.
+
+Another potential enhancement one could use is to use the trail cache to try to learn a second clause automatically when the solver reaches a conflict.
+However, we did not find a way to implement this to make it beneficial and the number of levels we have to scan for a second conflict would have to be more than the number of levels we already look ahead on each decision (when the solver is not under conflict).
 
@@
 \begin{figure}
-\begin{tabular}{|l|l|l|l|l|}
+\begin{tabular}{|l|l|l|l|l|l|}
 \hline
-                 & cadical                     & cadical\_chrono & cadical\_trail & cadical\_trail\_enhanced \\ \hline
-Instances Solved & 527                         & 515             & 520            & 524                      \\ \hline
-Par-2 (x $10^6$s)& 3.220                  & 3.269      & 3.248     & 3.216               \\ \hline
-                 & maple                       & maple\_chrono   & maple\_trail   & maple\_trail\_enhanced          \\ \hline
-Instance Solved  & 458                         & 470             & 472            & 473                      \\ \hline
-Par-2 (x $10^6$s)& 3.797 & 3.690      & 3.694     & 3.667               \\ \hline
+                         & Total Solved & SAT & UNSAT & Par-2 (x 10\textasciicircum{}6 s) \\ \hline
+cadical                  & 527          & 315 &  212  & 3.220                             \\ \hline
+cadical\_chrono          & 515          & 305 &  210  & 3.269                             \\ \hline
+cadical\_trail           & 520          & 307 &  213  & 3.248                             \\ \hline
+cadical\_trail\_enhanced & 524          & 315 &  209  & 3.216                             \\ \hline
 \end{tabular}
-\caption{Table of results}
+\caption{Table of results for cadical or "chrono", version published in \cite{DBLP:conf/sat/MohleB19}.}
 \end{figure}
+
+\begin{figure}
+\begin{tabular}{|l|l|l|l|l|l|}
+\hline
+                         & Total Solved & SAT & UNSAT \\ \hline
+cadical                  & 492          & 289 &  203  \\ \hline
+cadical\_chrono          & 498          & 292 &  206  \\ \hline
+cadical\_trail           & 493          & 290 &  203  \\ \hline
+cadical\_trail\_enhanced & 496          & 292 &  204  \\ \hline
+\end{tabular}
+\caption{Table of results for cadical, version pulled from github as of July 30, 2020.}
+\end{figure}
+
+\begin{figure}
+\begin{tabular}{|l|l|l|l|l|l|}
+\hline
+                & Total Solved & SAT & UNSAT & Par-2 (x 10\textasciicircum{}6 s) \\ \hline
+maple           & 458          & 259 &  199  & 3.797                             \\ \hline
+maple\_chrono   & 470          & 271 &  199  & 3.690                             \\ \hline
+maple\_trail    & 472          & 271 &  201  & 3.694                             \\ \hline
+maple\_enhanced & 473          & 277 &  195  & 3.667                             \\ \hline
+\end{tabular}
+\caption{Table of results for MapleLCMDist.}
+\end{figure}
+\clearpage
+
+\section{Future Work}
+
+With more clever data structures the efficiency of restoring the trail could be improved.
+There are a number of parameters which can be tuned, including how often to refresh the trail, how large the cutoff point for rejecting "long" reasons should be, how long to wait to prune the trail cache if one is appending multiple trail caches together, how far to look ahead on the trail cache for conflicts (in terms of number of decisions or total number of literals).
+We only experimented with a few different settings of these parameters that made sense, but more careful fine-tuning of these parameters could further enhance the effectiveness of the technique.
+There are potentially other techniques one could use a trail cache for, including inprocessing steps that involve probing.
+A look down the trail cache is an incomplete but very fast way to do a probing step, as opposed to propagation.
+Trail saving and chronological backtracking could be combined in some way to produce even better results.
+\bibliography{ref}{}
+\bibliographystyle{splncs04}
+\end{document}
+
+\iffalse
+\textit{Example 1} Suppose the literal $y$ can be restored from the trail cache, and its reason clause $cr$ contains the literals ${y \lnot x_1, \lnot x_2, ..., \lnot x_n}$.
+This means that $x1, ..., x_n$ have become true and are currently on the trail.
+Suppose $x_n$ is the most recent literal on the trail and  Without trail saving, the propagation routine will have to access $cr$ from the watchlist of one of the literals  $\blacksquare$\newline
+
+\textit{Example 1} Suppose the literals $x_1, ...
+x_n$ are on the trail cache, with none of them having a null reason, and $x_1$ becomes true (e.g., the solver makes a decision for $x_1$ to be true).
+Then, $x_2, ...
+x_n$ are implied by the current trail.
+Now we will examine the propagation of a clause c which consists of the literals $l, \lnot x_1, \lnot x_2, ..., \lnot x_n$ with and without trail saving.
+Assume $l$ and $\lnot x_1$ are the current watched literals.
+
+Without trail saving, $\lnot x_1$ becomes false and a new watcher is searched for and found 1 literal later, $\lnot x_2$, and then $\lnot x_1$ and $\lnot x_2$ are swapped.
+Next $\lnot x_2$ becomes false, a new watcher is searched for and found 2 literals later, $\lnot x_3$, and then $\lnot x_2$ and $\lnot x_3$ are swapped.
+This process will continue until all $x_n$ literals are false, c becomes unit, and $l$ becomes forced to be true.
+This propagation took $1 + 2 + ...
++ n$ searches through clause c, for a total of $O(n^2)$ searches.
+
+With trail saving, $x2, ...
+x_n$ are all assigned to be true as soon as $x_1$ is made true, so immediately the values of literals $\lnot x_1, ..., \lnot x_n$ in c are false.
+During propagation, a new watcher is searched for to replace $\lnot x_2$, n literals are saerched over without finding any, thus c is detected as unit and $l$ becomes forced to be true.
+This propagation took $O(n)$ searches through clause c.
+$\blacksquare$\newline
+\fi
+
+\iffalse
 \begin{figure}\includegraphics[scale=0.8]{cactus_cadical.pdf}\caption{\small{Comparison of run times for versions of cadical.
@@
 \begin{figure}\includegraphics[scale=0.8]{test2.pdf}\caption{}\end{figure}
+\fi
 \iffalse
@@
 \fi
-\clearpage
 
@@
 \fi
-
-\section{Conclusion}
-
-\bibliography{ref}{}
-\bibliographystyle{splncs04}
-\end{document}

### 2020-02-10 trail_diagram.tex
some edits

\documentclass{standalone}
\title{}
\date{}
\author{}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{algorithm}
\usepackage[noend]{algorithmic}
\usepackage{tikz}
\usetikzlibrary{decorations.pathreplacing,angles,quotes}
\begin{document}

\begin{tikzpicture}
\draw (0, 10) rectangle (1, 11);
\draw (0, 9.5) rectangle (1, 10);
\draw (0, 7) rectangle (1, 9.5);
\draw (0, 6.5) rectangle (1, 7);
\draw (0, 3) rectangle (1, 6.5);
\draw (0, 4.5) rectangle (1, 5);
\draw (0, 2.5) rectangle (1, 3);
\draw (0, 0) rectangle (1, 2.5);
\node[align=center] at (0.5, 10.75) {lvl 0};
\node[align=center] at (0.5, 9.75) {$x_{d_1}$};
\node[align=center] at (0.5, 8.5) {$x_1$ \\ $x_2$ \\ . \\ . \\ .};
\node[align=center] at (0.5, 6.75) {$x_{d_2}$};
\node[align=center] at (0.5, 5.25) {$x_j$};
\node[align=center] (A) at (0.5, 4.75) {$x_{d_3}$};
\node[align=center] at (0.5, 2.75) {$x_{d_4}$};
\node[align=center] (C) at (0.5, 0.25) {$x_n$};

\draw (3, 10) rectangle (4, 11);
\draw (3, 9.5) rectangle (4, 10);
\draw (3, 7) rectangle (4, 9.5);
\draw (3, 6.5) rectangle (4, 7);
\draw (3, 5) rectangle (4, 6.5);
\draw (3, 6.5) rectangle (4, 3.5);
\draw (5, 4.5) rectangle (6, 5);
\draw (5, 0) rectangle (6, 4.5);
%\draw (5, 2.5) rectangle (6, 3);
%\draw (5, 0) rectangle (6, 2.5);
\node[align=center] at (3.5, 10.75) {lvl 0};
\node[align=center] at (3.5, 9.75) {$x_{d_1}$};
\node[align=center] at (3.5, 8.5) {$x_1$ \\ $x_2$ \\ . \\ . \\ .};
\node[align=center] at (3.5, 6.75) {$x_{d_2}$};
\node[align=center] at (3.5, 5.25) {$x_j$};
\node[align=center] (B) at (5.5, 4.75) {$l_1$};
\node[align=center] at (5.5, 3.75) {$l_2$ \\ . \\ . \\ .};
%\node[align=center] at (5.5, 2.75) {$l_k$};
\node[align=center] at (5.5, 0.25) {$l_n$};
\node[align=center] at (6.5, 4.75) {top};

\draw[decoration={brace,raise=20pt},decorate] (0.5, 5) -- node[right=21pt] {$L_u$} (0.5, 0);
\draw[decoration={brace,raise=20pt},decorate] (0.5, 11) -- node[right=21pt] {$L_0$} (0.5, 5);
\draw[decoration={brace,raise=20pt},decorate] (3.5, 11) -- node[right=21pt] {$L_0$} (3.5, 5);
\draw[decoration={brace,raise=20pt},decorate] (3.5, 5) -- node[right=21pt] {$L_1$} (3.5, 3.5);
\draw[decoration={brace,mirror,raise=20pt},decorate] (3.5, 11) -- node[left=21pt] {$L_T$} (3.5, 3.5);

\node[align=center] at (0.5, -1) {Trail Before \\ Backtracking};
\node[align=center] at (3.5, -1) {Trail After \\ Backtracking};
\node[align=center] at (5.5, -1) {Trail Cache};
\draw[->] (A) -- (B);
\draw[->] (C) to [out=150] (A) node[rotate=90] at (-1,3) {backtrack};
\end{tikzpicture}
\end{document}

### 2020-02-10 report.tex
some edits

@@
 It also allows the search to choose the order of decisions, but helps make propagation faster.
-We explore the theoretical speedup that trail saving provides, develop some enhancements to make the idea more effective, and demonstrate experimentally that it speeds up state-of-the-art SAT solvers by a non-trivial factor.
+We explore the theoretical speedup that trail saving provides, develop some enhancements to make the idea more effective, and demonstrate experimentally that it performs similarly well as chronological backtracking without suffering from some of the aforementioned drawbacks.
 

### 2020-02-13 report.tex
some edits

@@
 The second side-effect of chronological backtracking is that it forces the solver to stay in the same local neighbourhood during search, since it unassigns less of the partial assignment and one could argue that this might lead to finding additional conflicts more quickly or reaching a satisfying assignment (if one exists) more quickly.
-This second side-effect also leads to the search with chronological backtracking being less flexibile to jump out of local neighbourhoods and obey the variable selection heuristic.
-We argue that this second side-effect is not the primary explanation for the effectiveness of chronological backtracking, and aim to to develop a technique which will produce a benefit similar to the first side-effect while maintaining the freedom for the solver to jump out of local neighbourhoods and obey the variable selection heuristic.
+However, this second side-effect also leads to less search flexibility, since the solver does not jump out of local neighbourhoods as often and does not obey the variable selection heuristic.
+We will argue that it is the first side-effect that leads to the effectiveness of chronological backtracking and not the second by developing a technique which also saves from having to repeat propagations, but allows the search to remain flexible and obey the variable selection heuristic.
 
@@
 As the literals from the trail cache get restored (i.e., assigned), the top of the trail cache will be updated to the next literal not yet restored.
+
+\begin{figure}\center\includegraphics[scale=0.7]{trail_diagram.pdf}\caption{Storing into a trail cache after backtracking.}\end{figure}
 
@@
 Since trail saving can sometimes save hundreds or thousands of literals at a time, the number of propagations required to detect a conflict could be very costly.
-
-\begin{figure}\includegraphics[scale=0.8]{trail_diagram.pdf}\caption{Storing into a trail cache after backtracking.}\end{figure}
 
@@
 Traverse down the trail cache and whenever a literal appears more than once, delete its entry in the trail cache.
-Whenever a literal $x$ appears whose complement $\lnot x$ has already appeared on the trail, $x$ should be deleted and everything remaining until the end of the trail cache should be deleted (in order to keep the trail cache free of conflicts).
-We decided to trim the trail cache whenever its size exceeded twice the total number of variables, as that is what worked best in preliminary experiments.\newline
+Whenever a literal $x$ appears whose complement $\lnot x$ has already appeared on the trail cache, $x$ should be deleted and everything remaining until the end of the trail cache should be deleted (in order to keep the trail cache free of conflicts).
+We decided to trim the trail cache whenever its size exceeded twice the total number of variables in the SAT instance, as that is what worked best in preliminary experiments.\newline
 
@@
 \hline
-                         & Total Solved & SAT & UNSAT \\ \hline
-cadical                  & 492          & 289 &  203  \\ \hline
-cadical\_chrono          & 498          & 292 &  206  \\ \hline
-cadical\_trail           & 493          & 290 &  203  \\ \hline
-cadical\_trail\_enhanced & 496          & 292 &  204  \\ \hline
+                         & Total Solved & SAT & UNSAT & Par-2 (x 10\textasciicircum{}6 s) \\ \hline
+cadical                  & 492          & 289 &  203  & 3.5020                            \\ \hline
+cadical\_chrono          & 498          & 292 &  206  & 3.4400                            \\ \hline
+cadical\_trail           & 493          & 290 &  203  & 3.5228                            \\ \hline
+cadical\_trail\_enhanced & 496          & 292 &  204  & 3.4847                            \\ \hline
 \end{tabular}
@@
 A look down the trail cache is an incomplete but very fast way to do a probing step, as opposed to propagation.
-Trail saving and chronological backtracking could be combined in some way to produce even better results.
+It is also possible that trail saving and chronological backtracking could be combined in some way to produce even better results.
 \bibliography{ref}{}

### 2020-02-16 figures/cactus_cadical.tex
fb notes for changes

--- 2020-01-27 cactus_maple.tex
@@
 
-\addplot table [x=index, y=original_time, col sep=comma] {results_maplelcmdist_data.csv};
-\addplot table [x=index, y=original_time, col sep=comma] {results_maplelcmdist_chrono_data.csv};
-\addplot table [x=index, y=original_time, col sep=comma] {results_maplelcmdist_trail_noforce_data.csv};
-\addplot table [x=index, y=original_time, col sep=comma] {results_maplelcmdist_chrono-dl_data.csv};
-\addplot table [x=index, y=original_time, col sep=comma] {results_maplelcmdist_chrono-dl_ncte-true-opt_data.csv};
+\addplot table [x=index, y=original_time, col sep=comma] {results_cadical-master_data.csv};
+\addplot table [x=index, y=original_time, col sep=comma] {results_cadical-master-chrono_data.csv};
+\addplot table [x=index, y=original_time, col sep=comma] {results_cadical-master-trail-working-version_data.csv};
+\addplot table [x=index, y=original_time, col sep=comma] {results_cadical-master-trail-1lvl-proper_data.csv};
 
-\legend{maplelcmdist,maplelcmdist-chrono,maplelcmdist-trail,maplelcmdist-chrono-dl,maplelcmdist-trail-dl}
+\legend{cadical,cadical-chrono,cadical-trail,cadical-trail-enhanced}
 \end{axis}

### 2020-02-16 figures/cactus_maple2.tex
fb notes for changes

--- 2020-01-27 cactus_maple.tex
@@
 \addplot table [x=index, y=original_time, col sep=comma] {results_maplelcmdist_trail_noforce_data.csv};
-\addplot table [x=index, y=original_time, col sep=comma] {results_maplelcmdist_chrono-dl_data.csv};
-\addplot table [x=index, y=original_time, col sep=comma] {results_maplelcmdist_chrono-dl_ncte-true-opt_data.csv};
+\addplot table [x=index, y=original_time, col sep=comma] {results_maplelcmdist_chrono-dl_ncte-true-opt-1lvl_data.csv};
 
-\legend{maplelcmdist,maplelcmdist-chrono,maplelcmdist-trail,maplelcmdist-chrono-dl,maplelcmdist-trail-dl}
+\legend{maplelcmdist,maplelcmdist-chrono,maplelcmdist-trail,maplelcmdist-trail-enhanced}
 \end{axis}

### 2020-02-16 figures/test.tex
fb notes for changes

\documentclass{standalone}
\title{}
\date{}
\author{}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{algorithm}
\usepackage[noend]{algorithmic}
\usepackage{tikz,pgfplotstable,pgfplots}
\usetikzlibrary{arrows}
\begin{document}

\begin{tikzpicture}
\begin{axis}[xlabel=Speedup or Slowdown Factor,ylabel=Frequency,width=12cm,height=12cm,
ybar=0pt,title=Cadical Trail Saving Run Time Comparison,
xticklabel style={/pgf/number format/fixed,/pgf/number format/precision=2},
yticklabel style={/pgf/number format/fixed,/pgf/number format/precision=5},]
\addplot table [x, y, col sep=tab] {2testdata};
\addplot table [x, y, col sep=tab] {testdata};
\legend{speedup, slowdown}
\draw [<-] (axis cs:0.1,700)-- +(10pt,10pt) node[right] {5104 instances had speedup or slowdown factor $< 1.07$};
\draw (axis cs:2.1,250) node[left] {34 instances had speedup factor $> 4$};
\draw (axis cs:2.1,200) node[left] {32 instances had slowdown factor $> 4$};
\iffalse
\draw [blue] (rel axis cs:0.54,0) -- (rel axis cs:0.54,1);
$\lvert log_2 \frac{\text{enqueue time}}{\text{trail saving time}} \rvert$
\fi
\end{axis}
\end{tikzpicture}

\end{document}

### 2020-02-16 figures/test2.tex
fb notes for changes

\documentclass{standalone}
\usepackage{tikz,pgfplotstable,pgfplots}
\pgfplotsset{compat=1.13}
\begin{document}
\begin{tikzpicture}
\begin{axis}[title=Trail Saving Running Time Per Instance,axis equal,ymin=0,xmin=0,
width=11cm,height=11cm,xlabel=original time (s),ylabel=trail saving time (s),axis x line=bottom,axis y line=left,xlabel near ticks,ylabel near ticks]
\addplot[only marks,mark size=1pt,mark=*,mark options={fill=black}]
	table[x=original,y=trail]
{test2.txt};
\addplot [red,samples at={0,1}] {x};
\draw [blue,dashed] (rel axis cs:0,0) -- (rel axis cs:1,1);
\end{axis}
\end{tikzpicture}
\end{document}

### 2020-02-16 figures/trail_diagram.tex
fb notes for changes

--- 2020-02-10 trail_diagram.tex
@@
 \node[align=center] at (3.5, 5.25) {$x_j$};
-\node[align=center] (B) at (5.5, 4.75) {$l_1$};
-\node[align=center] at (5.5, 3.75) {$l_2$ \\ . \\ . \\ .};
+\node[align=center] (B) at (5.5, 4.75) {$x_{d_3}$};
+\node[align=center] at (5.5, 3.75) {$x_k$ \\ . \\ . \\ .};
 %\node[align=center] at (5.5, 2.75) {$l_k$};
-\node[align=center] at (5.5, 0.25) {$l_n$};
+\node[align=center] at (5.5, 0.25) {$x_n$};
 \node[align=center] at (6.5, 4.75) {top};
 
-\draw[decoration={brace,raise=20pt},decorate] (0.5, 5) -- node[right=21pt] {$L_u$} (0.5, 0);
-\draw[decoration={brace,raise=20pt},decorate] (0.5, 11) -- node[right=21pt] {$L_0$} (0.5, 5);
-\draw[decoration={brace,raise=20pt},decorate] (3.5, 11) -- node[right=21pt] {$L_0$} (3.5, 5);
-\draw[decoration={brace,raise=20pt},decorate] (3.5, 5) -- node[right=21pt] {$L_1$} (3.5, 3.5);
-\draw[decoration={brace,mirror,raise=20pt},decorate] (3.5, 11) -- node[left=21pt] {$L_T$} (3.5, 3.5);
+\draw[decoration={brace,raise=20pt,amplitude=0.2cm},decorate] (0.5, 5) -- node[right=21pt] {$L_u$} (0.5, 0);
+\draw[decoration={brace,raise=20pt,amplitude=0.2cm},decorate] (0.5, 11) -- node[right=21pt] {$L_0$} (0.5, 5);
+\draw[decoration={brace,raise=20pt,amplitude=0.2cm},decorate] (3.5, 11) -- node[right=21pt] {$L_0$} (3.5, 5);
+\draw[decoration={brace,raise=20pt,amplitude=0.2cm},decorate] (3.5, 5) -- node[right=21pt] {$L_1$} (3.5, 3.5);
+\draw[decoration={brace,mirror,raise=20pt,amplitude=0.2cm},decorate] (3.5, 11) -- node[left=21pt] {$L_T$} (3.5, 3.5);
 

### 2020-02-16 paper.tex
fb notes for changes

--- 2020-02-13 report.tex
@@
-\documentclass{article}
+\documentclass[runningheads]{llncs}
 \usepackage{times}
@@
 \usepackage[ruled,linesnumbered,noend,oldcommands]{algorithm2e}
-\usepackage[T1,hyphens]{url}
+\usepackage[T1]{fontenc}
+\usepackage{hyperref}
 \usepackage{xspace}
@@
 \usepackage{listings}
+\usepackage{xcolor}
 \lstset { %
   language=C++,
-    %backgroundcolor=\color{black!5}, % set backgroundcolor
+  % backgroundcolor=\color{black!5}, % set backgroundcolor
   basicstyle=\footnotesize,% basic font setting
@@
 
+\renewcommand{\note}[1]{{\color{red}{#1}}}
+
+\title{Trail Saving on Backtrack}
+\author{Randy Hickey \and Fahiem Bacchus}
+\institute{Department of Computer Science, University of Toronto, Canada, 
+  \email{rgh000@gmail.com, fbacchus@cs.toronto.edu}}
+
 \begin{document}
-
+\maketitle
 \begin{abstract}
@@
 \section{Introduction}
-The vast majority of modern SAT solvers that are used to solve real-world problems are based on the conflict-driven clause learning (CDCL) algorithm.
-In a CDCL SAT solver, backtracking occurs after every conflict, where all literals from one or more decision levels become unassigned, before the solver resumes making decisions and performing unit propagations.
-Traditionally, CDCL solvers would backtrack non-chronologically to the conflict level, which is the second highest decision level remaining in the conflict clause after conflict analysis has resolved away all but one literal from the current decision level \cite{DBLP:conf/dac/MoskewiczMZZM01}.
-Recently, however, it has been shown that backtracking chronologically (i.e.
-backtracking one level only after conflict analysis) \cite{DBLP:conf/lpar/2013, DBLP:conf/sat/NadelR18} can be effective on many instances and was implemented in solvers that won the last two SAT competitions.
-Although chronological backtracking breaks some of the conventional invariants of CDCL solvers, it has been formalized and proven correct \cite{DBLP:conf/sat/MohleB19}.
+\note{Fix later---also \cite{DBLP:journals/jsat/TakRH11} can't be
+  first use of trail savings?}
 
-There are at least two possible side-effects of chronological backtracking which may explain its effectiveness.
-It has been observed that when a solver backtracks non-chronologically after a conflict, many of the literals that are unassigned during the backtrack will be re-assigned again in roughly the same order when the solver continues.
-By using chronological backtracking, the solver keeps a lot more of its partial assignment intact and saves from having to reconstruct a lot of the trail via propagation that it otherwise would have.
-Saving from having to reconstruct portions of the trail after backtracking was first introduced in the context of restarts \cite{DBLP:journals/jsat/TakRH11}.
-The second side-effect of chronological backtracking is that it forces the solver to stay in the same local neighbourhood during search, since it unassigns less of the partial assignment and one could argue that this might lead to finding additional conflicts more quickly or reaching a satisfying assignment (if one exists) more quickly.
-However, this second side-effect also leads to less search flexibility, since the solver does not jump out of local neighbourhoods as often and does not obey the variable selection heuristic.
-We will argue that it is the first side-effect that leads to the effectiveness of chronological backtracking and not the second by developing a technique which also saves from having to repeat propagations, but allows the search to remain flexible and obey the variable selection heuristic.
+The vast majority of modern SAT solvers that are used to solve
+real-world problems are based on the conflict-driven clause learning
+(CDCL) algorithm. In a CDCL SAT solver, backtracking occurs after
+every conflict, where all literals from one or more decision levels
+become unassigned, before the solver resumes making decisions and
+performing unit propagations. Traditionally, CDCL solvers would
+backtrack non-chronologically to the conflict level, which is the
+second highest decision level remaining in the conflict clause after
+conflict analysis has resolved away all but one literal from the
+current decision level \cite{DBLP:conf/dac/MoskewiczMZZM01}. Recently,
+however, it has been shown that backtracking chronologically
+(i.e. backtracking one level only after conflict analysis)
+\cite{DBLP:conf/lpar/2013,DBLP:conf/sat/NadelR18} can be effective on
+many instances and was implemented in solvers that won the last two
+SAT competitions. Although chronological backtracking breaks some of
+the conventional invariants of CDCL solvers, it has been formalized
+and proven correct \cite{DBLP:conf/sat/MohleB19}.
 
-Our alternative method is "trail saving", where we cache the part of the trail that is unassigned during a backtrack and then attempt to restore implied parts of it as each new decision is made.
-Trail saving preserves the traditional invariants of the SAT solver and its basic version is very simple to implement.
-It also allows the search to choose the order of decisions, but helps make propagation faster.
-We explore the theoretical speedup that trail saving provides, develop some enhancements to make the idea more effective, and demonstrate experimentally that it performs similarly well as chronological backtracking without suffering from some of the aforementioned drawbacks.
+There are at least two possible side-effects of chronological
+backtracking which may explain its effectiveness. It has been observed
+that when a solver backtracks non-chronologically after a conflict,
+many of the literals that are unassigned during the backtrack will be
+re-assigned again in roughly the same order when the solver
+continues. By using chronological backtracking, the solver keeps a lot
+more of its partial assignment intact and saves from having to
+reconstruct a lot of the trail via propagation that it otherwise would
+have. Saving from having to reconstruct portions of the trail after
+backtracking was first introduced in the context of restarts
+\cite{DBLP:journals/jsat/TakRH11}. The second side-effect of
+chronological backtracking is that it forces the solver to stay in the
+same local neighbourhood during search, since it unassigns less of the
+partial assignment and one could argue that this might lead to finding
+additional conflicts more quickly or reaching a satisfying assignment
+(if one exists) more quickly. However, this second side-effect also
+leads to less search flexibility, since the solver does not jump out
+of local neighbourhoods as often and does not obey the variable
+selection heuristic. We will argue that it is the first side-effect
+that leads to the effectiveness of chronological backtracking and not
+the second by developing a technique which also saves from having to
+repeat propagations, but allows the search to remain flexible and obey
+the variable selection heuristic.
+
+Our alternative method is "trail saving", where we cache the part of
+the trail that is unassigned during a backtrack and then attempt to
+restore implied parts of it as each new decision is made. Trail saving
+preserves the traditional invariants of the SAT solver and its basic
+version is very simple to implement. It also allows the search to
+choose the order of decisions, but helps make propagation faster. We
+explore the theoretical speedup that trail saving provides, develop
+some enhancements to make the idea more effective, and demonstrate
+experimentally that it performs similarly well as chronological
+backtracking without suffering from some of the aforementioned
+drawbacks.
 
 \section{Background}
-We will assume the reader is familiar with SAT solving and the basics of the CDCL algorithm \cite{DBLP:series/faia/SilvaLM09}.
-We will use the notion of a trail to represent a sequence of literals corresponding to the current partial assignment.
-The literals appear on the trail in the order in which they were assigned.
-Associated with every literal on the trail is a decision level.
-Decision level 0 represents all literals which are implied by the formula itself and subsequent decisions start the beginning of a new decision level.
-When a literal is made true by unit propagation, the literal is given its decision level and a reason clause, where the reason clause is the clause that was made unit by propagation in order to imply this literal.
-Most modern solvers use the two-literal watch scheme to make unit propagation more efficient, where each clause is watched by two non-false literals.
-When a watcher for a clause becomes false, a replacement is searched for in the clause.
-If no such replacement exists and the other watcher is true or unassigned, then the clause is unit.
-If no such replacement exists and the other watcher is also false, then the clause is a conflict and needs to be analyzed to produce a learnt clause.
+\note{elaborate backtrack after learning a clause. Also describe
+  properties of the trail (a) reason valid every reason clause is made
+  unit by its prefix, (b) non-contradictory, (c) non-conflicting (no
+  clause is falsified), and (d) propagation complete }
+
+We will assume the reader is familiar with SAT solving and the basics
+of the CDCL algorithm \cite{DBLP:series/faia/SilvaLM09}. We will use
+the notion of a trail to represent a sequence of literals
+corresponding to the current partial assignment. The literals appear
+on the trail in the order in which they were assigned. Associated with
+every literal on the trail is a decision level. Decision level 0
+represents all literals which are implied by the formula itself and
+subsequent decisions start the beginning of a new decision level. When
+a literal is made true by unit propagation, the literal is given its
+decision level and a reason clause, where the reason clause is the
+clause that was made unit by propagation in order to imply this
+literal. Most modern solvers use the two-literal watch scheme to make
+unit propagation more efficient, where each clause is watched by two
+non-false literals. When a watcher for a clause becomes false, a
+replacement is searched for in the clause. If no such replacement
+exists and the other watcher is true or unassigned, then the clause is
+unit. If no such replacement exists and the other watcher is also
+false, then the clause is a conflict and needs to be analyzed to
+produce a learnt clause.
 
 \section{Trail Saving}
-It can be observed that a SAT solver often repeats the same propagations over and over again during execution.
-Although no two descents down the trail will ever be identical because of clause learning, the solver will often propagate many of the same literals with the same reasons in a similar order to what it had done previously, especially since the order of the decisions usually remains similar.
-We plan to cache the part of the trail that we are backtracking over and use it to restore some of those literals without propagating them from scratch.
+\note{
+  \begin{enumerate}
+  \item Use simple backtrack and save to motivate and state the
+      invariant. 
+  \item Introduce notation for concatenating two trails.
+  \item Name and state the invariant in text (e.g. a
+      defn)---concenating the cached trail with the current trail
+      yields a trail that is always reason valid, but not necessarily
+      non-contradictory, non-conflicting, nor propagation complete).
+  \item Proof (simple) that the invariant holds for caching on
+      backtrack.
+  \item On backtrack at least the asserted literal and any new
+      propagants at the backtrack level are added.
+  \item Prove that the invariant holds (???do we need an invariant or
+      simply a property of the augmented trail) for trails with extras
+      inbeween.
+      $T_{\mathit{old}} + T_{\mathit{newlyAdded}} +
+      T_{\mathit{saved}}$ also satisfies the invariant
+  \item re-acheving propagation completeness for the augmented trail
+      (a) only add one level of the cached trail at a time, and do
+      unit prop after (b) only add non-contradicted decision from the
+      old trail. Also note that conflicts are valid and can be learnt
+      from even if the augmented trail is not propagation complete. 
+  \item Finally figure out a way to formalize the extra technique of
+      storing on top after repeated backtracking. 
+  \item Explain how freedom for the search is maintained by makeing
+      decisions then consulting the trail.
+  \item 
+  \end{enumerate}
+}
+
+It can be observed that a SAT solver often repeats the same
+propagations over and over again during execution. Although no two
+descents down the trail will ever be identical because of clause
+learning, the solver will often propagate many of the same literals
+with the same reasons in a similar order to what it had done
+previously, especially since the order of the decisions usually
+remains similar. We plan to cache the part of the trail that we are
+backtracking over and use it to restore some of those literals without
+propagating them from scratch.
 
@@
 
-The first step that is needed to use "trail saving" is to store all of the literals which are being unassigned during a backtrack, along with their respective reason clauses, in a "trail cache".
-Decisions should be marked with a null reason.
-The literals should appear in the same order on the trail cache as they had appeared on the trail.
-We will keep a pointer, which we will refer to as the "top" of the trail cache, and initially set it to be the literal which had appeared the earliest on the trail.
-Note that initially the top of the trail cache always has a null reason (i.e., it was the start of a new decision level on the previous trail).
-As the literals from the trail cache get restored (i.e., assigned), the top of the trail cache will be updated to the next literal not yet restored.
+The first step that is needed to use "trail saving" is to store all of
+the literals which are being unassigned during a backtrack, along with
+their respective reason clauses, in a "trail cache". Decisions should
+be marked with a null reason. The literals should appear in the same
+order on the trail cache as they had appeared on the trail. We will
+keep a pointer, which we will refer to as the "top" of the trail
+cache, and initially set it to be the literal which had appeared the
+earliest on the trail. Note that initially the top of the trail cache
+always has a null reason (i.e., it was the start of a new decision
+level on the previous trail). As the literals from the trail cache get
+restored (i.e., assigned), the top of the trail cache will be updated
+to the next literal not yet restored.
 
-\begin{figure}\center\includegraphics[scale=0.7]{trail_diagram.pdf}\caption{Storing into a trail cache after backtracking.}\end{figure}
+\begin{figure}\center\includegraphics[scale=0.7]{figures/trail_diagram.pdf}\caption{Storing into a trail cache after backtracking.}\end{figure}
 
 \subsection{Restoring the trail cache}
-We will base trail saving on the following critical invariant.
-If the top of the trail cache becomes true at any point, even after an arbitrary number of decisions and propagations have been made since the last backtrack, the set of literals on the trail cache after the top and up to but excluding the next literal with a null reason is still implied by the current trail.
-The top of the trail cache can become true either by decision or unit propagation, and this invariant still holds.\newline
+We will base trail saving on the following critical invariant. If the
+top of the trail cache becomes true at any point, even after an
+arbitrary number of decisions and propagations have been made since
+the last backtrack, the set of literals on the trail cache after the
+top and up to but excluding the next literal with a null reason is
+still implied by the current trail. The top of the trail cache can
+become true either by decision or unit propagation, and this invariant
+still holds.\newline
 
-\textit{Proof 1} We will refer to the set of literals that become unassigned during the backtrack after conflict analysis has taken place as $L_u$, and will refer to the rest of the trail that remains assigned as $L_0$.
-We will refer to all of the literals that get placed on the trail after $L_0$ during subsequent propagations and decisions as $L_1$.
-We will refer to the current trail $L_0 \cup L_1$ as $L_T$.
-The trail cache will contain the literals in $L_u$ ordered by their appearance on the old trail.
-Let $L_c[i]$ denote the set of the first i literals on the trail cache.
-Let $L_c[top]$ be the top of the old trail, which gets updated as parts of the trail cache get restored.
+\textit{Proof 1} We will refer to the set of literals that become
+unassigned during the backtrack after conflict analysis has taken
+place as $L_u$, and will refer to the rest of the trail that remains
+assigned as $L_0$. We will refer to all of the literals that get
+placed on the trail after $L_0$ during subsequent propagations and
+decisions as $L_1$. We will refer to the current trail $L_0 \cup L_1$
+as $L_T$. The trail cache will contain the literals in $L_u$ ordered
+by their appearance on the old trail. Let $L_c[i]$ denote the set of
+the first i literals on the trail cache. Let $L_c[top]$ be the top of
+the old trail, which gets updated as parts of the trail cache get
+restored.
 
-The $i$th literal on the trail cache is implied by $L_0 \cup L_c[i-1]$, as long as the $i$th literal has a non-null reason, since $L_0 \cup L_c[i-1]$ had forced $L_c[i]$ on the previous trail.
-$L_0 \subseteq L_T$, since $L_T = L_0 \cup L_1$.
-Therefore, the $i$th literal on the trail cache is also implied by $L_T \cup L_c[i-1]$.
+The $i$th literal on the trail cache is implied by
+$L_0 \cup L_c[i-1]$, as long as the $i$th literal has a non-null
+reason, since $L_0 \cup L_c[i-1]$ had forced $L_c[i]$ on the previous
+trail. $L_0 \subseteq L_T$, since $L_T = L_0 \cup L_1$. Therefore, the
+$i$th literal on the trail cache is also implied by
+$L_T \cup L_c[i-1]$.
 
-Assume $L_c[top]$ is true, which means that $L_c[0], ..., L_C[top] \subseteq L_1 \subseteq L_T$.
-Then, from above, every literal on the trail cache up to but excluding the next literal with a nun-null reason would be implied.
-$\blacksquare$\newline
+Assume $L_c[top]$ is true, which means that
+$L_c[0], ..., L_C[top] \subseteq L_1 \subseteq L_T$. Then, from above,
+every literal on the trail cache up to but excluding the next literal
+with a nun-null reason would be implied. $\blacksquare$\newline
 
-Once the backtrack has been completed and the solver resumes with decisions and propagations, we will attempt to restore parts of the trail cache automatically during propagation to make it faster by using the invariant stated above.
-Each time the propagation routine is about to propagate a new literal, we first check the top of the trail cache to determine its truth value.
-If that literal's truth value is already true, we assign all of the literals that come after it on the trail cache to be true up until (and excluding) the next literal with a null reason.
+Once the backtrack has been completed and the solver resumes with
+decisions and propagations, we will attempt to restore parts of the
+trail cache automatically during propagation to make it faster by
+using the invariant stated above. Each time the propagation routine is
+about to propagate a new literal, we first check the top of the trail
+cache to determine its truth value. If that literal's truth value is
+already true, we assign all of the literals that come after it on the
+trail cache to be true up until (and excluding) the next literal with
+a null reason.
 
-A potentially quadratic speedup can be realized during the propagation of these literals since they are being propagated as a set, rather than one at a time, as argued in \cite{DBLP:conf/sat/HickeyB19, DBLP:journals/jair/Gent13}.
-Restoring literals from the trail cache automatically not only leads to quicker propagation of future literals, but also means that the propagation routine has less work to do as it searches for new watchers in clauses on watchlists.
-All of the reasons attached to these literals that are automatically restored will now be satisfied, thus the propagation routine can skip over them rather than having to access each of them potentially multiple times to find they are unit.
+A potentially quadratic speedup can be realized during the propagation
+of these literals since they are being propagated as a set, rather
+than one at a time, as argued in \cite{DBLP:conf/sat/HickeyB19,DBLP:journals/jair/Gent13}. Restoring literals from the trail cache
+automatically not only leads to quicker propagation of future
+literals, but also means that the propagation routine has less work to
+do as it searches for new watchers in clauses on watchlists. All of
+the reasons attached to these literals that are automatically restored
+will now be satisfied, thus the propagation routine can skip over them
+rather than having to access each of them potentially multiple times
+to find they are unit.
 
-If at any time the top of the trail cache is a literal $l$ with a non-null reason whose truth value is already false, we determine a conflict has occurred with the conflict clause being $l$'s reason from the trail cache.
-This is because we know from Proof 1 that $l$ is implied to be true at this point by the reason on the trail cache (i.e., the reason on the trail cache is unit), but $\lnot l$ was also previously assigned to be true, therefore we have a conflict.
-This leads to a third potential speed up, since a conflict has been detected and none of the pending propagations have to be completed.
-Since trail saving can sometimes save hundreds or thousands of literals at a time, the number of propagations required to detect a conflict could be very costly.
+If at any time the top of the trail cache is a literal $l$ with a
+non-null reason whose truth value is already false, we determine a
+conflict has occurred with the conflict clause being $l$'s reason from
+the trail cache. This is because we know from Proof 1 that $l$ is
+implied to be true at this point by the reason on the trail cache
+(i.e., the reason on the trail cache is unit), but $\lnot l$ was also
+previously assigned to be true, therefore we have a conflict. This
+leads to a third potential speed up, since a conflict has been
+detected and none of the pending propagations have to be
+completed. Since trail saving can sometimes save hundreds or thousands
+of literals at a time, the number of propagations required to detect a
+conflict could be very costly.
 
@@
 
-In the standard version of trail saving outlined above, the trail cache needs to be deleted after each backtrack in order to make room for the new backtracked literals to be cached.
-The first enhancement to trail saving is to keep the trail cache intact instead of clearing it before each backtrack, and to add the literals that will be backtracked over to the beginning of the already existing trail cache.
-In order to preserve soundness, we must add the literals from the current backtrack to the beginning of the trail cache, such that the trail cache has as its prefix the literals which are being backtracked over, and as its suffix the previous trail cache.
-\newline
+In the standard version of trail saving outlined above, the trail
+cache needs to be deleted after each backtrack in order to make room
+for the new backtracked literals to be cached. The first enhancement
+to trail saving is to keep the trail cache intact instead of clearing
+it before each backtrack, and to add the literals that will be
+backtracked over to the beginning of the already existing trail
+cache. In order to preserve soundness, we must add the literals from
+the current backtrack to the beginning of the trail cache, such that
+the trail cache has as its prefix the literals which are being
+backtracked over, and as its suffix the previous trail cache. \newline
 
-\textit{Proof 2} We will use all of the same set names as Proof 1 and also refer to the set of literals on our previous trail cache as $L_{pc}$.
-We concatenate the literals in $L_{pc}$ to the end of $L_c$ to get $L_{sc}$.
-We know that $L_{pc}$ is a valid trail cache for $L_0 \cup L_u$.
-If we reach the end of $L_c$ without any conflicts, we know that $L_u \subseteq L_T$.
-Since $L_0 \subseteq L_T$, then $L_u \cup L_0 \subseteq L_T$ and therefore $L_{pc}$ is a valid trail cache for $L_T$.
-$\blacksquare$\newline
+\textit{Proof 2} We will use all of the same set names as Proof 1 and
+also refer to the set of literals on our previous trail cache as
+$L_{pc}$. We concatenate the literals in $L_{pc}$ to the end of $L_c$
+to get $L_{sc}$. We know that $L_{pc}$ is a valid trail cache for
+$L_0 \cup L_u$. If we reach the end of $L_c$ without any conflicts, we
+know that $L_u \subseteq L_T$. Since $L_0 \subseteq L_T$, then
+$L_u \cup L_0 \subseteq L_T$ and therefore $L_{pc}$ is a valid trail
+cache for $L_T$. $\blacksquare$\newline
 
-In order for the previous trail cache to be of any use, one must actually never store the last level of the current trail to the trail cache at all, since a learnt clause prevents the last level from being duplicated in the future.
-If any part of the previous trail cache was used to construct this last level, it must also be discarded from the previous trail cache before the previous trail cache is concatenated to the end.
+In order for the previous trail cache to be of any use, one must
+actually never store the last level of the current trail to the trail
+cache at all, since a learnt clause prevents the last level from being
+duplicated in the future. If any part of the previous trail cache was
+used to construct this last level, it must also be discarded from the
+previous trail cache before the previous trail cache is concatenated
+to the end.
 
-When concatenating trail caches together as above, it is important to note that the running trail cache can grow indefinitely.
-The trail cache can be trimmed periodically in the following way.
-Traverse down the trail cache and whenever a literal appears more than once, delete its entry in the trail cache.
-Whenever a literal $x$ appears whose complement $\lnot x$ has already appeared on the trail cache, $x$ should be deleted and everything remaining until the end of the trail cache should be deleted (in order to keep the trail cache free of conflicts).
-We decided to trim the trail cache whenever its size exceeded twice the total number of variables in the SAT instance, as that is what worked best in preliminary experiments.\newline
+When concatenating trail caches together as above, it is important to
+note that the running trail cache can grow indefinitely. The trail
+cache can be trimmed periodically in the following way. Traverse down
+the trail cache and whenever a literal appears more than once, delete
+its entry in the trail cache. Whenever a literal $x$ appears whose
+complement $\lnot x$ has already appeared on the trail cache, $x$
+should be deleted and everything remaining until the end of the trail
+cache should be deleted (in order to keep the trail cache free of
+conflicts).  We decided to trim the trail cache whenever its size
+exceeded twice the total number of variables in the SAT instance, as
+that is what worked best in preliminary experiments.\newline
 
-The next enhancement to trail saving is to scan the trail cache to check if there is a literal that is already assigned false within some limit.
-The limit we found worked best is to scan the trail until the third null reason is found (i.e.
-"looking ahead 2 decisions").
-If a literal that is already assigned false is found, we know that we are guaranteed to have a conflict in the near future, and we force the SAT solver's subsequent decisions to come from the trail cache instead of using the regular variable selection heuristic.
+The next enhancement to trail saving is to scan the trail cache to
+check if there is a literal that is already assigned false within some
+limit. The limit we found worked best is to scan the trail until the
+third null reason is found (i.e. "looking ahead 2 decisions"). If a
+literal that is already assigned false is found, we know that we are
+guaranteed to have a conflict in the near future, and we force the SAT
+solver's subsequent decisions to come from the trail cache instead of
+using the regular variable selection heuristic.
 
-Sometimes using trail saving to automatically restore literals with their previous reasons will lead to a literal getting a reason that is "worse" (in terms of number of literals, but one could use a different metric) than the reason it would have ended up with under normal propagation (i.e.
-without trail saving).
-The last enhancement we made to trail saving was to stop restoring literals from the old trail as soon as we hit a reason that we suspected might be "bad".
-For this, we kept a running mean and standard deviation of the length of all reasons used in the solver so far, and decided to stop restoring from the trail whenever we hit a reason that was greater in length than two standard deviations above the mean length.
-This way, when normal propagation resumes, we give it a chance to find a better reason than the one that trail saving had stored from the last descent down the trail.
+Sometimes using trail saving to automatically restore literals with
+their previous reasons will lead to a literal getting a reason that is
+"worse" (in terms of number of literals, but one could use a different
+metric) than the reason it would have ended up with under normal
+propagation (i.e. without trail saving). The last enhancement we made
+to trail saving was to stop restoring literals from the old trail as
+soon as we hit a reason that we suspected might be "bad". For this, we
+kept a running mean and standard deviation of the length of all
+reasons used in the solver so far, and decided to stop restoring from
+the trail whenever we hit a reason that was greater in length than two
+standard deviations above the mean length. This way, when normal
+propagation resumes, we give it a chance to find a better reason than
+the one that trail saving had stored from the last descent down the
+trail.
 
-Another potential enhancement one could use is to use the trail cache to try to learn a second clause automatically when the solver reaches a conflict.
-However, we did not find a way to implement this to make it beneficial and the number of levels we have to scan for a second conflict would have to be more than the number of levels we already look ahead on each decision (when the solver is not under conflict).
+Another potential enhancement one could use is to use the trail cache
+to try to learn a second clause automatically when the solver reaches
+a conflict. However, we did not find a way to implement this to make
+it beneficial and the number of levels we have to scan for a second
+conflict would have to be more than the number of levels we already
+look ahead on each decision (when the solver is not under conflict).
 
@@
 \subsection{Trail Saving for Assumption-Based SAT}
-Trail saving was briefly introduced in the context of assumption-based SAT \cite{DBLP:conf/sat/HickeyB19} in order to speed up the propagation of the assumptions each time the solver backtracked past the assumption levels.
-Note that the process is the same, but it only works to restore the propagants and implicants of the assumptions and only after a learnt unit clause.
-Becuase the number of propagants and implicants of the assumptions tends to become dominated by the number of assumptions in the applications it was tested on, the technique did not yield a significant improvement.
-It would be interesting to combine the extended version of trail saving outlined in this paper with trail saving over the assumptions to see if that could produce better results.
+Trail saving was briefly introduced in the context of assumption-based
+SAT \cite{DBLP:conf/sat/HickeyB19} in order to speed up the
+propagation of the assumptions each time the solver backtracked past
+the assumption levels. Note that the process is the same, but it only
+works to restore the propagants and implicants of the assumptions and
+only after a learnt unit clause. Becuase the number of propagants and
+implicants of the assumptions tends to become dominated by the number
+of assumptions in the applications it was tested on, the technique did
+not yield a significant improvement. It would be interesting to
+combine the extended version of trail saving outlined in this paper
+with trail saving over the assumptions to see if that could produce
+better results.
 
 \subsection{Comparison to Chronological Backtracking}
-As mentioned earlier, chronological backtracking also achieves some of the theoretical speedups that trail saving achieves.
-One advantage of chronological backtracking is that it does not require any caching of the trail or accessing such a cache during propagation; it simply leaves a vast portion of the trail intact rather than backtracking all the way to the conflict level.
-Chronological backtracking will also speed up propagations slightly more than trail saving, because it leaves more literals assigned at once which leads to clauses being detected as unit more quickly.
+As mentioned earlier, chronological backtracking also achieves some of
+the theoretical speedups that trail saving achieves. One advantage of
+chronological backtracking is that it does not require any caching of
+the trail or accessing such a cache during propagation; it simply
+leaves a vast portion of the trail intact rather than backtracking all
+the way to the conflict level. Chronological backtracking will also
+speed up propagations slightly more than trail saving, because it
+leaves more literals assigned at once which leads to clauses being
+detected as unit more quickly.
 
-One disadvantage that chronological backtracking has compared to trail saving is that it doesn't allow the search as much flexibility to jump out of local neighbourhoods, as the common variable selection heuristics tend to depend on \cite{DBLP:conf/dac/MoskewiczMZZM01, DBLP:conf/sat/2015, DBLP:conf/sat/LiangGPC16}.
-Because chronological backtracking leaves a large portion of the trail assigned, it will behave similarly to how the solver would behave if it was to force the decisions that do occur within this large portion of literals, whereas trail saving does not force any decisions.
-As variable selection heuristics become more refined, it is likely that adhering to them will become more crucial to performance.
+One disadvantage that chronological backtracking has compared to trail
+saving is that it doesn't allow the search as much flexibility to jump
+out of local neighbourhoods, as the common variable selection
+heuristics tend to depend on
+\cite{DBLP:conf/dac/MoskewiczMZZM01,DBLP:conf/sat/2015,DBLP:conf/sat/LiangGPC16}. Because
+chronological backtracking leaves a large portion of the trail
+assigned, it will behave similarly to how the solver would behave if
+it was to force the decisions that do occur within this large portion
+of literals, whereas trail saving does not force any decisions. As
+variable selection heuristics become more refined, it is likely that
+adhering to them will become more crucial to performance.
 
-Another disadvantage is that, because propagation is not performed at every decision level, literals that remain assigned during chronological backtracking (but otherwise would have been unassigned) do not have a chance to move their decision levels up.
-An example of this is when the solver was supposed to backtrack over a literal that would end up being unit propagated at level 0, but stays assigned at its current decision level when chronological backtracking is used.
-This could potentially effect the lbd score or length of the learnt clause that gets generated at the next conflict.
-Additionally, chronological backtracking breaks the invarant that all literals appear on the trail in monotonically increasing order by decision level, which makes its implementation slightly more complicated.
+Another disadvantage is that, because propagation is not performed at
+every decision level, literals that remain assigned during
+chronological backtracking (but otherwise would have been unassigned)
+do not have a chance to move their decision levels up. An example of
+this is when the solver was supposed to backtrack over a literal that
+would end up being unit propagated at level 0, but stays assigned at
+its current decision level when chronological backtracking is
+used. This could potentially effect the lbd score or length of the
+learnt clause that gets generated at the next conflict. Additionally,
+chronological backtracking breaks the invarant that all literals
+appear on the trail in monotonically increasing order by decision
+level, which makes its implementation slightly more complicated.
 
 \section{Experiments and Results}
-We implemented our techniques in two different SAT solvers, MapleSAT and Cadical, both of which have finished at or near the top of SAT competitions for the past several years.
-Each of these solvers was run as is without any trail saving or chronological backtracking in one version, with chronological backtracking in another version, and with the various types of trail saving we described in the remaining versions.
+We implemented our techniques in two different SAT solvers, MapleSAT
+and Cadical, both of which have finished at or near the top of SAT
+competitions for the past several years. Each of these solvers was run
+as is without any trail saving or chronological backtracking in one
+version, with chronological backtracking in another version, and with
+the various types of trail saving we described in the remaining
+versions.
 
-We then ran each solver on the 400 benchmarks used in the 2018 SAT Competition and the 400 benchmarks used in the SAT 2019 Competition.
-The experiments were run on 2.7 GHz Intel cores with each process given 7 GB of memory and 5000 seconds of CPU time.
-We chose not to output or verify the proofs generated by any of the solvers.
-The results are reported in Figure 1.
+We then ran each solver on the 400 benchmarks used in the 2018 SAT
+Competition and the 400 benchmarks used in the SAT 2019
+Competition. The experiments were run on 2.7 GHz Intel cores with each
+process given 7 GB of memory and 5000 seconds of CPU time. We chose
+not to output or verify the proofs generated by any of the
+solvers. The results are reported in Figure 1.
 
-When comparing regular cadical with non-chronological backtracking against cadical with trail saving, the average time taken to solve those instances that were solved by both solvers decreased and the number of propagations per second increased.
-Although the total number of solved instances decreased for both versions of trail saving, the Par-2 score increased for both of them.
-The trail saving versions also solved more instances and had better Par-2 scores than cadical with chronological backtracking (default parameters).
+When comparing regular cadical with non-chronological backtracking
+against cadical with trail saving, the average time taken to solve
+those instances that were solved by both solvers decreased and the
+number of propagations per second increased. Although the total number
+of solved instances decreased for both versions of trail saving, the
+Par-2 score increased for both of them. The trail saving versions also
+solved more instances and had better Par-2 scores than cadical with
+chronological backtracking (default parameters).
 
-When comparing regular MapleSAT with non-chronological backtracking against either of the MapleSAT trail saving versions, the total number of instances solved increased significantly, the average time taken to solve those instances which were solved by both solvers decreased, and the number of propagations per second increased.
-When comparing the trail saving versions to MapleSAT with chronological backtracking (default parameters), the trail saving versions perform slightly better in both total instances solved and PAR-2 scores.
+When comparing regular MapleSAT with non-chronological backtracking
+against either of the MapleSAT trail saving versions, the total number
+of instances solved increased significantly, the average time taken to
+solve those instances which were solved by both solvers decreased, and
+the number of propagations per second increased. When comparing the
+trail saving versions to MapleSAT with chronological backtracking
+(default parameters), the trail saving versions perform slightly
+better in both total instances solved and PAR-2 scores.
 
@@
 
-With more clever data structures the efficiency of restoring the trail could be improved.
-There are a number of parameters which can be tuned, including how often to refresh the trail, how large the cutoff point for rejecting "long" reasons should be, how long to wait to prune the trail cache if one is appending multiple trail caches together, how far to look ahead on the trail cache for conflicts (in terms of number of decisions or total number of literals).
-We only experimented with a few different settings of these parameters that made sense, but more careful fine-tuning of these parameters could further enhance the effectiveness of the technique.
-There are potentially other techniques one could use a trail cache for, including inprocessing steps that involve probing.
-A look down the trail cache is an incomplete but very fast way to do a probing step, as opposed to propagation.
-It is also possible that trail saving and chronological backtracking could be combined in some way to produce even better results.
-\bibliography{ref}{}
+With more clever data structures the efficiency of restoring the trail
+could be improved. There are a number of parameters which can be
+tuned, including how often to refresh the trail, how large the cutoff
+point for rejecting "long" reasons should be, how long to wait to
+prune the trail cache if one is appending multiple trail caches
+together, how far to look ahead on the trail cache for conflicts (in
+terms of number of decisions or total number of literals). We only
+experimented with a few different settings of these parameters that
+made sense, but more careful fine-tuning of these parameters could
+further enhance the effectiveness of the technique. There are
+potentially other techniques one could use a trail cache for,
+including inprocessing steps that involve probing. A look down the
+trail cache is an incomplete but very fast way to do a probing step,
+as opposed to propagation. It is also possible that trail saving and
+chronological backtracking could be combined in some way to produce
+even better results.
+
+\bibliography{sat}{}
 \bibliographystyle{splncs04}
@@
 
-\iffalse
-\textit{Example 1} Suppose the literal $y$ can be restored from the trail cache, and its reason clause $cr$ contains the literals ${y \lnot x_1, \lnot x_2, ..., \lnot x_n}$.
-This means that $x1, ..., x_n$ have become true and are currently on the trail.
-Suppose $x_n$ is the most recent literal on the trail and  Without trail saving, the propagation routine will have to access $cr$ from the watchlist of one of the literals  $\blacksquare$\newline
+\iffalse \textit{Example 1} Suppose the literal $y$ can be restored
+from the trail cache, and its reason clause $cr$ contains the literals
+${y \lnot x_1, \lnot x_2, ..., \lnot x_n}$. This means that
+$x1, ..., x_n$ have become true and are currently on the
+trail. Suppose $x_n$ is the most recent literal on the trail and
+Without trail saving, the propagation routine will have to access $cr$
+from the watchlist of one of the literals $\blacksquare$\newline
 
-\textit{Example 1} Suppose the literals $x_1, ...
-x_n$ are on the trail cache, with none of them having a null reason, and $x_1$ becomes true (e.g., the solver makes a decision for $x_1$ to be true).
-Then, $x_2, ...
-x_n$ are implied by the current trail.
-Now we will examine the propagation of a clause c which consists of the literals $l, \lnot x_1, \lnot x_2, ..., \lnot x_n$ with and without trail saving.
-Assume $l$ and $\lnot x_1$ are the current watched literals.
+\textit{Example 1} Suppose the literals $x_1, ... x_n$ are on the
+trail cache, with none of them having a null reason, and $x_1$ becomes
+true (e.g., the solver makes a decision for $x_1$ to be true). Then,
+$x_2, ... x_n$ are implied by the current trail. Now we will examine
+the propagation of a clause c which consists of the literals
+$l, \lnot x_1, \lnot x_2, ..., \lnot x_n$ with and without trail
+saving. Assume $l$ and $\lnot x_1$ are the current watched literals.
 
-Without trail saving, $\lnot x_1$ becomes false and a new watcher is searched for and found 1 literal later, $\lnot x_2$, and then $\lnot x_1$ and $\lnot x_2$ are swapped.
-Next $\lnot x_2$ becomes false, a new watcher is searched for and found 2 literals later, $\lnot x_3$, and then $\lnot x_2$ and $\lnot x_3$ are swapped.
-This process will continue until all $x_n$ literals are false, c becomes unit, and $l$ becomes forced to be true.
-This propagation took $1 + 2 + ...
-+ n$ searches through clause c, for a total of $O(n^2)$ searches.
+Without trail saving, $\lnot x_1$ becomes false and a new watcher is
+searched for and found 1 literal later, $\lnot x_2$, and then
+$\lnot x_1$ and $\lnot x_2$ are swapped. Next $\lnot x_2$ becomes
+false, a new watcher is searched for and found 2 literals later,
+$\lnot x_3$, and then $\lnot x_2$ and $\lnot x_3$ are swapped. This
+process will continue until all $x_n$ literals are false, c becomes
+unit, and $l$ becomes forced to be true. This propagation took
+$1 + 2 + ... + n$ searches through clause c, for a total of $O(n^2)$
+searches.
 
-With trail saving, $x2, ...
-x_n$ are all assigned to be true as soon as $x_1$ is made true, so immediately the values of literals $\lnot x_1, ..., \lnot x_n$ in c are false.
-During propagation, a new watcher is searched for to replace $\lnot x_2$, n literals are saerched over without finding any, thus c is detected as unit and $l$ becomes forced to be true.
-This propagation took $O(n)$ searches through clause c.
-$\blacksquare$\newline
-\fi
+With trail saving, $x2, ... x_n$ are all assigned to be true as soon
+as $x_1$ is made true, so immediately the values of literals
+$\lnot x_1, ..., \lnot x_n$ in c are false. During propagation, a new
+watcher is searched for to replace $\lnot x_2$, n literals are
+saerched over without finding any, thus c is detected as unit and $l$
+becomes forced to be true. This propagation took $O(n)$ searches
+through clause c. $\blacksquare$\newline \fi
 
 \iffalse
-\begin{figure}\includegraphics[scale=0.8]{cactus_cadical.pdf}\caption{\small{Comparison of run times for versions of cadical.
-cadical is without chronological backtracking, cadical-chrono is with chronological backtracking, cadical-trail is with the first version of trail saving, cadical-trail-2levels-ahead is the trail saving with all enhancements added.}}\end{figure}
-\begin{figure}\includegraphics[scale=0.8]{cactus_maple2.pdf}\caption{\small{Comparison of run times for versions of MapleSAT.
-maplelcmdist is from the 2017 SAT Competition, maplelecmdist-chrono is with chronological backtracking (2018 SAT Competition), maplelcmdist-trail is with the first version of trail saving, maplelcmdist-trail-enhanced is with trail saving and all enhancements added.}}\end{figure}
-\begin{figure}\includegraphics[scale=0.8]{test.pdf}\caption{}\end{figure}
-\begin{figure}\includegraphics[scale=0.8]{test2.pdf}\caption{}\end{figure}
+\begin{figure}\includegraphics[scale=0.8]{figures/cactus_cadical.pdf}\caption{\small{Comparison
+        of run times for versions of cadical. cadical is without
+        chronological backtracking, cadical-chrono is with
+        chronological backtracking, cadical-trail is with the first
+        version of trail saving, cadical-trail-2levels-ahead is the
+        trail saving with all enhancements added.}}\end{figure}
+\begin{figure}\includegraphics[scale=0.8]{figures/cactus_maple2.pdf}\caption{\small{Comparison
+        of run times for versions of MapleSAT. maplelcmdist is from
+        the 2017 SAT Competition, maplelecmdist-chrono is with
+        chronological backtracking (2018 SAT Competition),
+        maplelcmdist-trail is with the first version of trail saving,
+        maplelcmdist-trail-enhanced is with trail saving and all
+        enhancements added.}}\end{figure}
+\begin{figure}\includegraphics[scale=0.8]{figures/test.pdf}\caption{}\end{figure}
+\begin{figure}\includegraphics[scale=0.8]{figures/test2.pdf}\caption{}\end{figure}
 \fi
 \iffalse
-\begin{figure}\includegraphics[scale=0.8]{cactus_maple.pdf}\caption{\small{Comparison of run times for versions of MapleSAT.
-maplelcmdist is from the 2017 SAT Competition, maplelecmdist-chrono is with chronological backtracking (2018 SAT Competition), maplelcmdist-trail is with the first version of trail saving, maplelcmdist-chrono-dl is with chronological backtracking and duplicate learnts (2019 SAT Competition), maplelcmdist-trail-dl is with trail saving and duplicate learnts}}\end{figure}
+\begin{figure}\includegraphics[scale=0.8]{figures/cactus_maple.pdf}\caption{\small{Comparison
+        of run times for versions of MapleSAT. maplelcmdist is from
+        the 2017 SAT Competition, maplelecmdist-chrono is with
+        chronological backtracking (2018 SAT Competition),
+        maplelcmdist-trail is with the first version of trail saving,
+        maplelcmdist-chrono-dl is with chronological backtracking and
+        duplicate learnts (2019 SAT Competition),
+        maplelcmdist-trail-dl is with trail saving and duplicate
+        learnts}}\end{figure}
 \fi

### 2020-02-16 Solver_1.cc

@@ -1,2577 +1,0 @@
-/***************************************************************************************[Solver.cc]
-MiniSat -- Copyright (c) 2003-2006, Niklas Een, Niklas Sorensson
-           Copyright (c) 2007-2010, Niklas Sorensson
- 
-Chanseok Oh's MiniSat Patch Series -- Copyright (c) 2015, Chanseok Oh
- 
-Maple_LCM, Based on MapleCOMSPS_DRUP -- Copyright (c) 2017, Mao Luo, Chu-Min LI, Fan Xiao: implementing a learnt clause minimisation approach
-Reference: M. Luo, C.-M. Li, F. Xiao, F. Manya, and Z. L. , “An effective learnt clause minimization approach for cdcl sat solvers,” in IJCAI-2017, 2017, pp. to–appear.
-
-Maple_LCM_Dist, Based on Maple_LCM -- Copyright (c) 2017, Fan Xiao, Chu-Min LI, Mao Luo: using a new branching heuristic called Distance at the beginning of search
-
-MapleLCMDistChronoBT, based on Maple_LCM_Dist -- Copyright (c), Alexander Nadel, Vadim Ryvchin: "Chronological Backtracking" in SAT-2018, pp. 111-121.
-
-MapleLCMDistChronoBT-DL, based on MapleLCMDistChronoBT -- Copyright (c), Stepan Kochemazov, Oleg Zaikin, Victor Kondratiev, Alexander Semenov: The solver was augmented with heuristic that moves duplicate learnt clauses into the core/tier2 tiers depending on a number of parameters.
-
-Permission is hereby granted, free of charge, to any person obtaining a copy of this software and
-associated documentation files (the "Software"), to deal in the Software without restriction,
-including without limitation the rights to use, copy, modify, merge, publish, distribute,
-sublicense, and/or sell copies of the Software, and to permit persons to whom the Software is
-furnished to do so, subject to the following conditions:
-
-The above copyright notice and this permission notice shall be included in all copies or
-substantial portions of the Software.
-
-THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR IMPLIED, INCLUDING BUT
-NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND
-NONINFRINGEMENT. IN NO EVENT SHALL THE AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM,
[…]

### 2020-02-18 paper.tex
results

@@
 \section{Experiments and Results}
-We implemented our techniques in two different SAT solvers, MapleSAT
-and Cadical, both of which have finished at or near the top of SAT
-competitions for the past several years. Each of these solvers was run
-as is without any trail saving or chronological backtracking in one
-version, with chronological backtracking in another version, and with
-the various types of trail saving we described in the remaining
-versions.
+For the version of cadical published in
+\cite{DBLP:conf/sat/MohleB19}chronological backtracking gives solves 6
+more problems and has a lower Par-2 score by a decent margin. When
+adding trail saving and all enhancements to the default
+non-chronological version of cadical, we end up solving 4 more problems
+and the Par-2 score goes decreases by a margin not as good as that
+produced by the non-chronological backtracking version. The trail saving
+versions improve on the standard non-chronological backtracking but do
+slightly worse than the chronological backtracking versions.
 
-We then ran each solver on the 400 benchmarks used in the 2018 SAT
-Competition and the 400 benchmarks used in the SAT 2019
-Competition. The experiments were run on 2.7 GHz Intel cores with each
-process given 7 GB of memory and 5000 seconds of CPU time. We chose
-not to output or verify the proofs generated by any of the
-solvers. The results are reported in Figure 1.
+For the latest version of cadical that we downloaded as of January 1,
+2020, chronological backtracking (set to default parameters)
+interestingly makes the solver perform worse than standard
+non-chronological backtracking does. This shows that not all solvers
+will benefit from chronological backtracking; it depends on the solver's
+mix of heuristics and may require fine-tuning the parameters associated
+with chronological backtracking. However, after implementing trail
+saving and enhancements on top of this version of cadical with
+non-chronological backtracking, the solver performs roughly the same as
+it does with standard non-chronological backtracking, and does not
+suffer a slow down as chronological backtracking appears to.
 
-When comparing regular cadical with non-chronological backtracking
-against cadical with trail saving, the average time taken to solve
-those instances that were solved by both solvers decreased and the
-number of propagations per second increased. Although the total number
-of solved instances decreased for both versions of trail saving, the
-Par-2 score increased for both of them. The trail saving versions also
-solved more instances and had better Par-2 scores than cadical with
-chronological backtracking (default parameters).
-
-When comparing regular MapleSAT with non-chronological backtracking
-against either of the MapleSAT trail saving versions, the total number
-of instances solved increased significantly, the average time taken to
-solve those instances which were solved by both solvers decreased, and
-the number of propagations per second increased. When comparing the
-trail saving versions to MapleSAT with chronological backtracking
-(default parameters), the trail saving versions perform slightly
-better in both total instances solved and PAR-2 scores.
+For MapleSat, chronological backtracking makes a significant improvement
+on MapleLCMDist both in terms of total number of problems solved and
+Par-2 score. Similarly, trail saving implemented on top of MapleLCMDist
+with non-chronological backtracking receives a similarly significant
+improvement in terms of total instances solved and Par-2 score, and
+actually outperforms chronological backtracking on both measures.
 
@@
     \end{tabular}
-    \caption{Table of results for cadical, version pulled from github as of July 30, 2020.}
+    \caption{Table of results for cadical, version pulled from github as of July 30, 2019.}
 \end{figure}

### 2020-02-18 paper.tex
results

@@
       & Total Solved & SAT & UNSAT & Par-2 (x 10\textasciicircum{}6 s) \\ \hline
-      cadical                  & 527          & 315 &  212  & 3.220                             \\ \hline
-      cadical\_chrono          & 515          & 305 &  210  & 3.269                             \\ \hline
-      cadical\_trail           & 520          & 307 &  213  & 3.248                             \\ \hline
-      cadical\_trail\_enhanced & 524          & 315 &  209  & 3.216                             \\ \hline
+      cadical                  & 532          & 314 &  218  & 3.220                             \\ \hline
+      cadical\_chrono          & 525          & 308 &  217  & 3.269                             \\ \hline
+      cadical\_trail           & 529          & 310 &  219  & 3.248                             \\ \hline
+      cadical\_trail\_enhanced & 531          & 313 &  218  & 3.216                             \\ \hline
     \end{tabular}
-    \caption{Table of results for cadical or "chrono", version published in \cite{DBLP:conf/sat/MohleB19}.}
+    \caption{Table of results for cadical, version pulled from github as of January 1, 2020.}
 \end{figure}
@@
     \end{tabular}
-    \caption{Table of results for cadical, version pulled from github as of July 30, 2019.}
+    \caption{Table of results for cadical or "chrono", version published in \cite{DBLP:conf/sat/MohleB19}.}
 \end{figure}

### 2020-02-19 paper.tex
notes and useful papers

@@
 \section{Introduction}
+\textbf{Overall message of the paper.}
+\begin{enumerate}
+\item Formalism for understanding a more general form of trail savings
+    which caches a copy of the trail that is backtracked over, and
+    verifying how it exploited soundly.
+
+\item Potential advantages over chronological backtracking.  Literals
+    remain at the correct level, supports the solver having more
+    freedom in its decisions.
+
+\item Potential advantages of techique (1) increase in efficiency of
+    propagation, (2) the new uses of the cached information.
+
+\item Potential disadvantage. Reasons are recycled, in some circumstances
+    new reasons might be better. We show in a simple experiment that
+    this effect can occur.
+
+\item Techique captures most of the positive effects of
+    non-chronological backtracking in a simpler to implement way. It
+    also preserves the simpler standard structure of CDCL
+    solvers. Already provides some advantages for current SAT
+    solvers. The ideas also have the potential for further
+    exploitation. Lookahead for conflicts seems to be beneficial, and
+    potentially if better discrimatory measures could be developed for
+    reason clauses we could also better determine whether or not to
+    use the cached information.. 
+\end{enumerate}
+
 \note{Fix later---also \cite{DBLP:journals/jsat/TakRH11} can't be
   first use of trail savings?}
+
+\note{Reusing the Assignment Trail in CDCL Solvers: restart to level
+  $i$ where the next decision has a higher score than the decision at
+  level $i$ and a lower score than all decisions at level $j$ for
+  $j< i$.
+
+  They argum changes to reason clauses don't make much difference, but
+  that was with minisat, before metrics for clause quality were in place in sat solvers.
+
+  The savings for Luby-100 (used in minisat) were about 1\%}
 

### 2020-02-21 paper.tex
added abstract

@@
 \usepackage{amssymb}
-\usepackage[noend]{algcompatible}
-\usepackage[ruled,linesnumbered,noend,oldcommands]{algorithm2e}
+\usepackage{algorithm}
+\usepackage[noend]{algpseudocode}
+%\usepackage[noend]{algcompatible}
+%\usepackage[ruled,linesnumbered,noend,oldcommands]{algorithm2e}
 \usepackage[T1]{fontenc}
@@
 \usepackage{xcolor}
+\usepackage{theorem}
+\algrenewcommand\algorithmicprocedure{}
+\algrenewcommand\algorithmicthen{}
+\algrenewcommand\algorithmicdo{}
+
 \lstset { %
@@
 \renewcommand{\note}[1]{{\color{red}{#1}}}
+\newcommand{\fs}[1]{\fontsize{#1}{#1}\selectfont}
+\newcommand{\fss}[2]{\fontsize{#1}{#2}\selectfont}
+\newcommand{\sat}{SAT\xspace}
+\newcommand{\trail}{\ensuremath{\mathcal{T}}}
+\newcommand{\trailIdxT}[2]{\ensuremath{\iota_{#1}(#2)}}
+\newcommand{\trailIdx}[1]{\ensuremath{\iota(#1)}}
+\newcommand{\range}[2]{#1\ldots#2}
+\newcommand{\dlevel}[1]{\ensuremath{\mathit{decLvl}(#1)}}
+\newcommand{\dlevels}{\ensuremath{\mathit{decLvls}}}
+\newcommand{\var}{\text{var}}
+\newcommand{\true}{\textsc{true}\xspace}
+\newcommand{\false}{\textsc{false}\xspace}
+\newcommand{\reason}[1]{\ensuremath{\mathit{reason}(#1)}}
+\newcommand{\reasonsave}[1]{\ensuremath{\mathit{reason_{\mathit{save}}(#1)}}}
+\newcommand{\litsave}{\ensuremath{\mathit{l_{\mathit{save}}}}}
+\newcommand{\formula}{\ensuremath{\mathcal{F}}}
+\renewcommand{\implies}{\rightarrow}
+\newcommand{\uipcls}{C_{\textit{1-UIP}}}
+\newcommand{\deepestLvl}{L_{\textit{deep}}}
+\newcommand{\deepestLit}{\ell_{\textit{deep}}}
+\newcommand{\btL}{L_{\textit{back}}}
+\newcommand{\cbt}{C-bt\xspace}
+\newcommand{\trailsave}{\trail_{\mathit{save}}}
+
+\newtheorem{Obs}{Observation}
+\newtheorem{Cor}{Corollary}
+\newtheorem{defn}{Definition}
+\newtheorem{thm}{Theorem}
+\newtheorem{ex}{Example}
+\newcommand{\whitebox}{\raisebox{.5ex}{\fbox{\hspace*{.2ex}}}}
 
@@
 \begin{abstract}
+    A CDCL \sat solver can backtrack a large distance when it learns a
+    new clause, e.g, when the new learnt clause is a unit clause the
+    solver has to backtrack to level zero. When the length of the
+    backtrack is large the solver can end up reproducing many of the
+    same decisions and propagations when it redescent the search
+    tree. Motivated by this potential redundancy, partial backtracking
+    (chronological backtracking) has been proposed. This technique
+    allows the solver to reduce the length of its backtrack, e.g.,
+    allowing the solver to backtrack to the previous level after a
+    conflict is found. However, this technique has its shortcomings
+    and only tends to work when limits are placed on its
+    application. In this paper we present a new trail saving technique
+    that makes a copy of the part of the trail that was backtracked
+    over. This saved copy can then be used to improve the efficiency
+    of the solver's subsequent redescent. Although our technique does
+    not save as much work as chronological backtracking it also does
+    not have chronological backtracking's shortcomings. Furthermore,
+    the saved trail also provides the solver the ability to look ahead
+    along the previous trail. This can be exploited to improve its
+    efficiency. Our empirical results show that our technique often
+    yields superior performance over chronological backtracking, and
+    that is is able to improve the performance of state of the art
+    solvers.
+
 \end{abstract}
@@
 
-\note{Fix later---also \cite{DBLP:journals/jsat/TakRH11} can't be
-  first use of trail savings?}
-
 \note{Reusing the Assignment Trail in CDCL Solvers: restart to level
@@
 however, it has been shown that backtracking chronologically
-(i.e. backtracking one level only after conflict analysis)
-\cite{DBLP:conf/lpar/2013,DBLP:conf/sat/NadelR18} can be effective on
-many instances and was implemented in solvers that won the last two
+\cite{DBLP:conf/lpar/JiangZ13} (i.e. backtracking one level only after
+conflict analysis)
+\cite{DBLP:conf/sat/NadelR18,DBLP:conf/sat/MohleB19} can be effective
+on many instances and was implemented in solvers that won the last two
 SAT competitions. Although chronological backtracking breaks some of
@@
 have. Saving from having to reconstruct portions of the trail after
-backtracking was first introduced in the context of restarts
-\cite{DBLP:journals/jsat/TakRH11}. The second side-effect of
-chronological backtracking is that it forces the solver to stay in the
-same local neighbourhood during search, since it unassigns less of the
-partial assignment and one could argue that this might lead to finding
-additional conflicts more quickly or reaching a satisfying assignment
-(if one exists) more quickly. However, this second side-effect also
-leads to less search flexibility, since the solver does not jump out
-of local neighbourhoods as often and does not obey the variable
-selection heuristic. We will argue that it is the first side-effect
-that leads to the effectiveness of chronological backtracking and not
-the second by developing a technique which also saves from having to
-repeat propagations, but allows the search to remain flexible and obey
-the variable selection heuristic.
+backtracking was first introduced in the context of restarts by van
+der Tak et al. \cite{DBLP:journals/jsat/TakRH112}. The second
+side-effect of chronological backtracking is that it forces the solver
+to stay in the same local neighbourhood during search, since it
+unassigns less of the partial assignment and one could argue that this
+might lead to finding additional conflicts more quickly or reaching a
+satisfying assignment (if one exists) more quickly. However, this
+second side-effect also leads to less search flexibility, since the
+solver does not jump out of local neighbourhoods as often and does not
+obey the variable selection heuristic. We will argue that it is the
+first side-effect that leads to the effectiveness of chronological
+backtracking and not the second by developing a technique which also
+saves from having to repeat propagations, but allows the search to
+remain flexible and obey the variable selection heuristic.
 
@@
 \section{Background}
-\note{elaborate backtrack after learning a clause. Also describe
-  properties of the trail (a) reason valid every reason clause is made
-  unit by its prefix, (b) non-contradictory, (c) non-conflicting (no
-  clause is falsified), and (d) propagation complete }
+A propositional formula $\formula$ expressed in Conjunctive Normal
+Form (CNF), contains a set of variables $V$. A literal is a variable
+$v\in V$ or its negation $\lnot v$, and for a literal $l$ we let
+$\var(l)$ denote its underlying variable. A CNF consists of a
+conjunction of clauses, each of which is a disjunction of literals. We
+often view a clause as being a set of literals and employ set
+notation, e.g., $\ell\in C$ and $C'\subset C$.
 
-We will assume the reader is familiar with SAT solving and the basics
-of the CDCL algorithm \cite{DBLP:series/faia/SilvaLM09}. We will use
-the notion of a trail to represent a sequence of literals
-corresponding to the current partial assignment. The literals appear
-on the trail in the order in which they were assigned. Associated with
-every literal on the trail is a decision level. Decision level 0
-represents all literals which are implied by the formula itself and
-subsequent decisions start the beginning of a new decision level. When
-a literal is made true by unit propagation, the literal is given its
-decision level and a reason clause, where the reason clause is the
-clause that was made unit by propagation in order to imply this
-literal. Most modern solvers use the two-literal watch scheme to make
-unit propagation more efficient, where each clause is watched by two
-non-false literals. When a watcher for a clause becomes false, a
-replacement is searched for in the clause. If no such replacement
-exists and the other watcher is true or unassigned, then the clause is
-unit. If no such replacement exists and the other watcher is also
-false, then the clause is a conflict and needs to be analyzed to
-produce a learnt clause.
+We assume the reader is familiar with the basic operations of CDCL
+\sat solvers. A good source for this background is
+\cite{DBLP:series/faia/SilvaLM09}.
+
+\paragraph{Trails.}
+CDCL \sat solvers maintain a trail which is the sequence of literals
+that have currently been assigned \true by the solver. During its
+operation a \sat solver will add newly assigned literals to the end of
+the trail, and on backtrack remove literals from the end of the trail
+(unassigning them as they are removed).
+
+A \sat solver's trail satisfies a number of conditions that we will
+discuss below. However, in we will need some additional flexibility in
+our definitions, as we will be sometimes be working with trails that
+would never be constructed by a \sat solver. Hence, we define a
+\emph{trail} to be a sequence of literals each of which is either a
+\emph{decision} literal or an \emph{implied} literal, and each of
+which has a \emph{reason}. These two types of literals are
+distinguished by their \emph{reasons}. Decision literals $d$ have a
+null reason, $\reason{d} = \varnothing$. Implied literals $l$ have as
+a reason a clause of the formula $\formula$,
+$\reason{l} = C\in\formula$. ($\reason{l}$ can be a learnt clause that
+has been added to $\formula$).
+
+If literal $\ell$ is on the trail $\trail$ let
+$\trailIdxT{\trail}{\ell}$ denote its index on the trail, i.e,
+$\trail[\trailIdxT{\trail}{\ell}] = \ell$. For convenience, we also let
+$\trailIdxT{\trail}{\ell} = \trailIdxT{\trail}{\lnot \ell} =
+\trailIdxT{\trail}{\var(\ell)}$ even though neither $\lnot \ell$ nor
+$\var(\ell)$ are actually on $\trail$. If $x$ and $y$ are both on the
+trail and $\trailIdxT{\trail}{x} < \trailIdxT{\trail}{y}$ we say that
+\textit{$x$ appears before $y$ on the trail}. For convenience, when
+the trail being discussed in clear from context we simply write
+$\trailIdx{}$ instead of $\trailIdxT{\trail}{}$.
+
+Each literal $\ell\in\trail$ has a decision level $\dlevel{\ell}$
+which is equal to the number of decision literals appearing before
+$\ell$ on the trail plus one; hence, $\dlevel{d}=1$ for the first
+decision literal $d\in\trail$. The set of literals on $\trail$ that
+have the same decision level forms a \emph{contiguous}
+subsequence\footnote{Our approach uses standard trails in which the
+  decision levels are contiguous. Chronological backtracking
+  \cite{DBLP:conf/lpar/JiangZ13,DBLP:conf/sat/NadelR18,DBLP:conf/sat/MohleB19}
+  generates trails with non-contigous decision levels} that starts
+with a decision literal $d_i$ and ends just before the next decision
+literal $d_{i+1}$. We will often need to refer to different decision
+level subsequences of $\trail$. Hence, we let $\trail[[i]]$ denote the
+subsequence of literals at decision level $i$; and let
+$\trail[[\range{i}{j}]]$ denote the subsequence of literals at
+decision levels $k$ for $i\leq k\leq j$. Now we define the
+following properties that a trail $\trail$ can have.
+\begin{description}
+\item[non-contradictory:] A variable cannot appear in both polarities
+    in the trail: $l\in \trail\implies \lnot l \not \in \trail$.
+\item[non-redundant:] A literal can only appear once on $\trail$.
+\item[reason-sound:] For each implied literal $l\in\trail$ we have
+    that it reason clause has \emph{been made unit by $\trail$
+      implying $l$}. This means that the reason clause contains $l$
+    and the negation all of its other literals are on $\trail$ before
+    $l$:
+    $\forall l\in\trail.\, \reason{l}\neq \varnothing \implies
+    l\in\reason{l} \land \bigl(\forall x\in\reason{l}.\, x\neq l
+    \implies \lnot x \in \trail \land \trailIdx{\lnot x} <
+    \trailIdx{l}\bigr)$.
+\item[propagation-complete:] Unit propgation has been run to
+    completion at all decisions levels of $\trail$. This means that
+    literals appear on $\trail$ at the first decision level they were
+    unit implied. Formally, this can be captured by the condition:
+    $\forall i \in \{\dlevel{l}\,|\,l\in \trail\}. \bigl(\exists
+    C\in\formula. \mbox{$C$ is made unit by $\trail[[\range{0}{i}]]$
+      implying $l$}\bigr) \implies l\in \trail[[\range{0}{i}]]$. With
+    non-redundancy $l$ will appear only at the first decision level it
+    was implied, without non-redundancy, $l$ might appear more than
+    once. Note that this also implies that $\reason{l}$ must contain
+    at least one other literal $y\neq l$ with
+    $\dlevel{y} = \dlevel{l}$.
+\item[conflict free:] No clause of $F$ is falsified by
+    $\trail$. Clauses $C\in F$ falsified by $\trail$ are typically
+    called \emph{conflicts}.
+\end{description}
+
+In CDCL solvers using standard conflict directed backtracking all
+properties hold of the prefix of the solver's trail consisting of all
+decisions levels but the deepest. The full trail might, however,
+contain a conflict at its deepest level so is not necessarily conflict
+free. The full trail might also not be propagation-complete, as unit
+propagation at the deepest level is typically terminated early if a
+conflict is found. It can further be noted that the first four
+properties imply that if a clause $C$ is falsified at decision level
+$k$, then $C$ must contain at least two literals at level $k$
+(otherwise $C$ would have become unit at a prior level and then
+satisfied by making its last unfalsified literal \true).
+
+\paragraph{Standard Backtracking.}
+In CDCL \sat solving the solver extends its trail by adding new
+decision literals followed by finding and adding all unit implied
+literals arising from that new decision. This continues until it
+reaches a decision level $\deepestLvl$ where a conflict $C$ is found.
+
+In standard backtracking, the solver then constructs a new 1-UIP
+clause by resolving away all but one literal at level $\deepestLvl$
+from the conflict $C$ using the reason clauses of these literals. (As
+noted above $C$ must contain at least two literals at level
+$\deepestLvl$). Hence, the new clause $\uipcls$ will contain one
+literal $\deepestLit$ at level $\deepestLvl$ and have all of its other
+literals a levels less than $\deepestLvl$. The solver then backtracks
+to $\btL$ the second deepest level in $\uipcls$. This involves
+changing $\trail$ to its prefix $\trail[[\range{0}{\btL}]]$
+(unassigning all literals removed from $\trail$). The new clause
+$\uipcls$ is made unit by $\trail[[\range{0}{\btL}]]$ implying
+$\deepestLit$, so the solver then adds $\deepestLit$ to the trail and
+executes another round of unit propagation at level $\btL$, after
+which it continues by once again growing the trail with new decisions
+and unit implied literals until a new conflict or a satisfying
+assignment is found.
+
+In standard backtracking, the difference between the backtrack level,
+$\btL$ and the current deepest level $\deepestLvl$ can be very
+large. During its new descent from $\btL$ the solver can reproduce a
+large number of the same decisions and unit propagations, essentially
+wasting work. This potential inefficiency has been noted in prior work
+\cite{DBLP:journals/jsat/TakRH11,DBLP:conf/lpar/JiangZ13,DBLP:conf/sat/NadelR18,DBLP:conf/sat/MohleB19}.
+
+In \cite{DBLP:journals/jsat/TakRH11} a technique for reducing the
+length of the backtrack during restarts was presented. In restarts,
+the solver backtracks to level $0$, and this technique involves
+computing a new deeper backtrack level $M> 0$ for which it is known
+that on redescent the first $M+1$ levels of the trail will be
+unchanged (except perhaps for the ordering of the literals). This
+technique removes the redundant work of reproducing the first $M$
+trail levels. Unfortunately, with backtracking the trail will be
+changed at level $\btL$ ($\deepestLit$ will be newly inserted at this
+level). Thus there seems to be no choice but to backtrack to
+$\btL$. In this paper we will show that although we have to backtrack
+to $\btL$ we can make the subsequent redescent much more efficient.
+
+\paragraph{Chronological Backtracking.}
+Chronological bactracking (\cbt) is a technique originally propsed by
+Jiang and Zhang \cite{DBLP:conf/lpar/JiangZ13} under the name
+\textit{partial backtracking}. Its aim is to reduce the redundant work
+that might be done by the \sat solver on its redescent from the
+backtrack level $\btL$. The original paper proposed a technique for
+backtracking to any level $j$ in the range
+$\btL \leq j \leq \deepestLvl{-}1$ (where $\deepestLvl$ is the level
+the conflict was discovered. Nadel and Ryvchin
+\cite{DBLP:conf/sat/NadelR18} proposed to always backtrack to
+$\deepestLvl{-}1$ while M{\"{o}}hle and Biere
+\cite{DBLP:conf/sat/MohleB19} returned to the original proposal of
+flexibly backtracking any level in the allowed range. Note that the
+new 1-UIP clause $\uipcls$ is made unit at every level in this
+range. So after backtracking to level $j$ the newly implied literal
+$\deepestLit$ is added to the trail with
+$\reason{\deepestLit}=\uipcls$, and $\dlevel{\deepestLit}$ is set to
+$\btL$ (the second deepest level in $\uipcls$.
+
+This means that the decision levels on the trail are no longer
+contiguous, as $\deepestLit$ has a different level than the other
+literals at level $j$ (if $j\neq \btL$). This change has a number of
+consequences on the \sat solvers operation, all of which were
+described in the original paper \cite{DBLP:conf/lpar/JiangZ13}. In
+\cite{DBLP:conf/sat/NadelR18} some more practical implementation
+methods were developed to overcome this issues, and in
+\cite{DBLP:conf/sat/MohleB19} a formal framework for \cbt was
+developed under which the changes required to the \sat solver could be
+shown to be successful. We refer the reader to the cited papers for
+more details.
+
+\section{Potential Drawbacks of Chronological Backtracking}
+In this paper we present a new technique that allows the \sat solver
+to use standard backtracking, but also allows saving some redundant
+work on its redescent.  Our method has more overhead than \cbt so the
+first question that must be addressed is why not just use
+chronological backtracking.
+
+The reduction in redundant work achieved by \cbt does not come for
+free. In particular, some of the work avoided by \cbt is not actually
+redundant work; it is work that can actually help the solver. The
+obvious evidence for this statement is that in both
+\cite{DBLP:conf/sat/NadelR18} and \cite{DBLP:conf/sat/MohleB19} it was
+found that it was not optimal to always apply \cbt. In
+\cite{DBLP:conf/sat/NadelR18} \cbt was applied only when the length of
+the standard backtrack, $\btL-\deepestLvl$ was greater than a
+given threshold $T$. In their experiments they found that $T=100$ was
+the best value, i.e., \cbt is done only on longer backtracks. In
+practice, this meant that \cbt was \emph{hardly ever done}; in our
+measurements with their solver only about 3\% of the solver backtracks
+were \cbt backtracks. In \cite{DBLP:conf/sat/MohleB19} the value
+$T=100$ was also applied. However, they introduced an additional
+technique to add some applications of \cbt when the length of the
+backtrack is less than $T$. This allowed to utilize \cbt in \%15 of
+the backtracks. Nevertheless, the fact \cbt has to be used in a
+controlled fashion, indicates that can be benefits in the \sat solver
+employing standard backtracking; it cannot be the case that every
+redescent is only doing redundant work.
+
+What is this non-redundant work? With standard backtracking the
+literal $\deepestLit$ is placed on the trail at the end of $\btL$ and
+then unit propagated. This could impact the trail in at least the
+following ways. First, some literals might become unit at earlier
+levels. This could include decision literals becoming forced which
+might compress some decision levels together. Second, different
+decisions might be made due to changes in the variable scores arising
+from the newly learnt clause. And third, literals might be unit
+implied with different reasons. These changes could be sometimes be
+major, and they could have a large impact on future
+clauses. Unfortunately, it seems difficult to predict when these
+impacts might be large.
+
+The second impact, changing variable scores, is partially addressed in
+\cite{DBLP:conf/sat/MohleB19} who utilize the ideas of
+\cite{DBLP:journals/jsat/TakRH11} to backtrack to a level where the
+decisions would be unchanged. However, if the length of the backtrack
+is greater than $100$ there could still be a divergence between the
+variable decisions generated by standard backtracking and \cbt. An
+argument is also given in \cite{DBLP:journals/jsat/TakRH11} that the
+third impact, changing literal reasons, is not significant. However,
+the experiments in \cite{DBLP:journals/jsat/TakRH11} were run before
+good notions of clause quality were known
+\cite{DBLP:conf/ijcai/AudemardS09}. Our empirical results contradict
+this, indicating that the literal reasons can have a significant
+impact.
+
+The first impact is worth discussing since it was mentioned in the
+original \cbt paper \cite{DBLP:conf/lpar/JiangZ13} but not in the
+subsequent works. This is the issue of changing the decision levels of
+literals on the trail. \cbt computes the decision level of each
+implied literal based on the decision levels of its reasons, but it
+does not go backwards to change the decisions levels of literals
+earlier on the trail.
+
+\begin{example}
+    For example, say that $(x, \lnot y)\in \formula$, the literal $x$
+    is a decision literal on the trail with $\dlevel{x}=2$, and that
+    the solver is currently at level $150$ where it encounters a
+    conflict. If this conflict yields the unit clause $(y)$, standard
+    backtracking would backtrack to level $0$, where $x$ would be
+    implied. On redescent, $x$ would no longer form a new decision
+    level and it would not appear in any new clauses (as it is
+    entailed by $\formula$). \cbt, on the other hand, would backtrack
+    to level $149$. On its trail $x$ would still be at level
+    $2$. Until a backtrack past level $2$ occurs, learnt clauses might
+    contain $\lnot x$, and thus have level $2$ added to their set of
+    levels (potentially changing their LBD score). Only when backtrack
+    past level $2$ occurs would $x$ be restored to its correct level
+    $0$, and it would require inprocessing simplications to remove $x$
+    from the learnt clauses.
+\end{example}
+
+In sum, although these impacts of \cbt might or might not be harmful
+to the \sat solver, they exist. Hence, one unappealing aspect of \cbt
+is that introduces additional inter-dependencies between components of
+the solver. For example, the second impact means that heuristics for
+choosing decision variables might work differently in the presence of
+\cbt; and the first impact means that different ways of computing
+clause quality using decision levels in the clauses might also work
+differently in the presence of \cbt. So although there are bound to be
+cases where \cbt is superior, we would like to have a method that
+addresses the original motivation of reducing redundant work while not
+changing the operation of standard backtracking. We present our
+proposal for such a method next.
 
 \section{Trail Saving}
+Our approach is to save the trail, $\trail$, on backtrack, and to use
+the saved trail $\trailsave$ when the solver redesends to improve the
+effiency of propagations without affecting the decisions the solver
+wants to make. The saved trail $\trailsave$ also provides a secondary
+``lookahead mechanism'' that the \sat solver can exploit as it
+redescends.
+
+Say that the solver is at $\deepestLvl$ where it has encountered a
+conflict. From the 1-UIP clause it learns,, $\uipcls$, it now has to
+backtrack to $\btL$. This is accomplished by calling the following
+subroutine with argument $\btL$.
+\begin{algorithmic}[1]
+    \Procedure{backtrack}{$\btL$}
+    \State $\forall \ell\in\trail[[\range{\btL{+}1}{\deepestLvl}]]$ unassign($\ell$)
+    \State $\forall \ell\in\trail[[\range{\btL{+}1}{\deepestLvl{-}1}]]$ $\reasonsave{\ell} = \reason{\ell}$
+    \State $\trailsave = \trail[[\range{\btL{+}1}{\deepestLvl{-}1}]]$
+    \State $\trail = \trail[[\range{0}{\btL}]]$
+    \EndProcedure
+\end{algorithmic}
+
+Note that we do not save the deepest level of $\trail$. The full
+$\trail$ contains a conflict (at its deepest level). Hence the solver
+will never reproduce all the same levels, and it would be useless to
+save all of them. Note also that in addition to saving the literals in
+$\trailsave$ we also save the clause reason of the unit implied
+literals in a $\mathit{reason}_{save}$ vector. Finally, we see that
+the first literal on $\trailsave$ must be a decision literal: it is
+the first literal of $\trail$ at decision level $\btL+1$.
+
+\paragraph{Using $\trailsave$ for more efficient unit propagations.}
+After backtrack the solver will add $\deepestLit$ to the end of the
+updated $\trail$ with $\reason(\deepestLit)=\uipcls$ and then invoke
+unit propagation which will add more literals to
+$\trail$. $\trailsave$ is exploited during propagation by calling the
+following version of propagation with the argument
+$\trailIdx(\deepestLit)$ (i.e., the trail index of the newly added
+implicant.
+
+\begin{algorithm}[t]
+\begin{algorithmic}[1]
+    \Procedure{unitPropagate}{idx}
+    \While{idx $\mbox{} < \trail.\mathrm{size}$()}
+        \State $c\gets\mbox{}$ \textsc{useSavedTrail}()
+        \If{($c\neq \varnothing$)} \textbf{return} $c$ \Comment{Found conflict from $\trailsave$}
+        \EndIf
+        \State $\ell \gets \trail[idx]$
+        \For{each clause $c\in$ watchlist($\lnot \ell$)}
+            \If{($c$ is unit implying $x$)}
+                 \State $\trail.\mathrm{addToEnd}(x)$
+                 \State $\reason{x} \gets c$            
+            \ElsIf{($c$ is falsified)}\:
+                 \textbf{return} $c$ \Comment{Found conflict from unit prop.}
+            \Else\:
+                 Update $c$'s watches.
+            \EndIf
+        \EndFor
+        \State $idx$++
+    \EndWhile
+    \EndProcedure
+    \Statex
+    \Procedure{useSavedTrail}{\mbox{}}
+    \State $idx\gets 0$; $\litsave\gets \trailsave[idx]$
+    \State \qquad Stop at decision literal on $\trailsave$ not yet assigned \true in $\trail$
+    \While{($\reasonsave{\litsave} \neq \varnothing \lor \litsave \in \trail$)}
+          \If{($\lnot \litsave\in\trail$)}\: \textbf{return} $\reasonsave{\litsave}$ 
+                \Comment{Conflict: $\litsave$ already \false}
+          \ElsIf{($\litsave \not\in\trail$)} \Comment{$\litsave$ unassigned}
+              \State $\trail.\mathrm{addToEnd(\litsave)}$
+              \State $\reason{\litsave}\gets \reasonsave{\litsave}$
+          \EndIf
+          \State $idx$++
+    \EndWhile
+    \EndProcedure
+\end{algorithmic}
+
+\caption{Using $\trailsave$ to enhance unit propagation and conflict
+  detection}
+\end{algorithm}
+
 \note{
@@
 
-\begin{algorithm}
-  	\TitleOfAlgo{propagate()}
-	\begin{algorithmic}[1]
-        \WHILE{queue not empty}
-        \STATE { }
-        \STATE restoreTrailCache()
-        \STATE { }
-        \STATE $l := $ next literal on queue
-        \FOR{each clause $c \in $ watchlist($\neg l$)}
-		\FOR{each literal $x \in c$}
-        \IF{$val($x$) \neq$ false and x is not already a watcher of c}
-        \STATE let $x$ be a watcher of $c$ and add $c$ to watchlist($\neg x$)
-        \STATE break
-        \ENDIF
-		\ENDFOR
-		\IF{no new watcher was found}
-        \IF{other watcher $x_2$ of $c = $ false}
-        \STATE return c as conflict
-        \ELSIF{other watcher $x_2$ of $c$ is unassigned}
-        \STATE enqueue $x_2$
-        \ENDIF
-		\ENDIF
-        \ENDFOR
-        \ENDWHILE
-	\end{algorithmic}
-\end{algorithm}
+% \begin{algorithm}
+%   	\TitleOfAlgo{propagate()}
+% 	\begin{algorithmic}[1]
+%         \While{queue not empty}
+%         \Statex
+%         \State restoreTrailCache()
+%         \Statex
+%         \State $l := $ next literal on queue
+%         \For{each clause $c \in $ watchlist($\neg l$)}
+% 		\For{each literal $x \in c$}
+%         \If{$val($x$) \neq$ false and x is not already a watcher of c}
+%         \State let $x$ be a watcher of $c$ and add $c$ to watchlist($\neg x$)
+%         \State break
+%         \EndIf
+% 		\EndFor
+% 		\If{no new watcher was found}
+%         \If{other watcher $x_2$ of $c = $ false}
+%         \State return c as conflict
+%         \ElsIf{other watcher $x_2$ of $c$ is unassigned}
+%         \State enqueue $x_2$
+%         \EndIf
+% 		\EndIf
+%         \EndFor
+%         \EndWhile
+% 	\end{algorithmic}
+% \end{algorithm}
 
-\begin{algorithm}
-  	\TitleOfAlgo{restoreTrailCache()}
-	\begin{algorithmic}[1]
-        \IF{trail saving turned on}
-		\STATE{let tc $=$ top of trail cache}
-		\IF{val(top of trail cache) $=$ true}
-        \WHILE{trail cache not empty}
-        \STATE{move top of trail cache (tc) to next literal}
-        \IF{reason(tc) is null}
-        \STATE{break}
-        \ELSIF{val(tc) $=$ false}
-        \STATE{return conflict with reason(tc)}
-        \ELSIF{val(tc) $=$ unassigned}
-        \STATE{enqueue tc with reason reason(tc)}
-        \ENDIF
-        \ENDWHILE
-		\ENDIF
-        \ENDIF
-	\end{algorithmic}
-\end{algorithm}
+% \begin{algorithm}
+%   	\TitleOfAlgo{restoreTrailCache()}
+% 	\begin{algorithmic}[1]
+%         \If{trail saving turned on}
+% 		\State{let tc $=$ top of trail cache}
+% 		\If{val(top of trail cache) $=$ true}
+%         \While{trail cache not empty}
+%         \State{move top of trail cache (tc) to next literal}
+%         \If{reason(tc) is null}
+%         \State{break}
+%         \ElsIf{val(tc) $=$ false}
+%         \State{return conflict with reason(tc)}
+%         \ElsIf{val(tc) $=$ unassigned}
+%         \State{enqueue tc with reason reason(tc)}
+%         \EndIf
+%         \EndWhile
+% 		\EndIf
+%         \EndIf
+% 	\end{algorithmic}
+% \end{algorithm}
 

### 2020-02-22 paper.tex
typo

@@
     solver has to backtrack to level zero. When the length of the
-    backtrack is large the solver can end up reproducing many of the
-    same decisions and propagations when it redescent the search
+    backtrack is large, the solver can end up reproducing many of the
+    same decisions and propagations when it redescends the search
     tree. Motivated by this potential redundancy, partial backtracking
@@
     of the solver's subsequent redescent. Although our technique does
-    not save as much work as chronological backtracking it also does
+    not save as much work as chronological backtracking, it also does
     not have chronological backtracking's shortcomings. Furthermore,

### 2020-02-24 paper.tex
fb changes clean version (except for experiments)

--- 2020-02-21 paper.tex
@@
 \usepackage{listings}
-\usepackage{xcolor}
+\usepackage[table]{xcolor}
 \usepackage{theorem}
+
 \algrenewcommand\algorithmicprocedure{}
@@
 \algrenewcommand\algorithmicdo{}
+\definecolor{linen}{rgb}{0.98, 0.94, 0.9}
+\definecolor{lightcyan}{rgb}{0.88, 1.0, 1.0}
 
@@
 \newcommand{\trailsave}{\trail_{\mathit{save}}}
+\newcommand{\bt}{\textsc{backtrack}\xspace}
+\newcommand{\ust}{\textsc{useSavedTrail}\xspace}
 
@@
 \newtheorem{thm}{Theorem}
+\newtheorem{prop}{Proposition}
 \newtheorem{ex}{Example}
@@
 \author{Randy Hickey \and Fahiem Bacchus}
-\institute{Department of Computer Science, University of Toronto, Canada, 
+\institute{Department of Computer Science, University of Toronto, Canada\\
   \email{rgh000@gmail.com, fbacchus@cs.toronto.edu}}
@@
     not have chronological backtracking's shortcomings. Furthermore,
-    the saved trail also provides the solver the ability to look ahead
-    along the previous trail. This can be exploited to improve its
-    efficiency. Our empirical results show that our technique often
-    yields superior performance over chronological backtracking, and
-    that is is able to improve the performance of state of the art
+    the saved trail also provides the solver with the ability to look
+    ahead along the previous trail. This can be exploited to improve
+    its efficiency. Our empirical results show that our technique
+    often yields superior performance over chronological backtracking,
+    and that is is able to improve the performance of state of the art
     solvers.
@@
 \textbf{Overall message of the paper.}
-\begin{enumerate}
-\item Formalism for understanding a more general form of trail savings
-    which caches a copy of the trail that is backtracked over, and
-    verifying how it exploited soundly.
-
-\item Potential advantages over chronological backtracking.  Literals
-    remain at the correct level, supports the solver having more
-    freedom in its decisions.
-
-\item Potential advantages of techique (1) increase in efficiency of
-    propagation, (2) the new uses of the cached information.
-
-\item Potential disadvantage. Reasons are recycled, in some circumstances
-    new reasons might be better. We show in a simple experiment that
-    this effect can occur.
-
-\item Techique captures most of the positive effects of
-    non-chronological backtracking in a simpler to implement way. It
-    also preserves the simpler standard structure of CDCL
-    solvers. Already provides some advantages for current SAT
-    solvers. The ideas also have the potential for further
-    exploitation. Lookahead for conflicts seems to be beneficial, and
-    potentially if better discrimatory measures could be developed for
-    reason clauses we could also better determine whether or not to
-    use the cached information.. 
-\end{enumerate}
-
-\note{Reusing the Assignment Trail in CDCL Solvers: restart to level
-  $i$ where the next decision has a higher score than the decision at
-  level $i$ and a lower score than all decisions at level $j$ for
-  $j< i$.
-
-  They argum changes to reason clauses don't make much difference, but
-  that was with minisat, before metrics for clause quality were in place in sat solvers.
-
-  The savings for Luby-100 (used in minisat) were about 1\%}
+
+% \begin{enumerate}
+% \item Formalism for understanding a more general form of trail savings
+%     which caches a copy of the trail that is backtracked over, and
+%     verifying how it exploited soundly.
+
+% \item Potential advantages over chronological backtracking.  Literals
+%     remain at the correct level, supports the solver having more
+%     freedom in its decisions.
+
+% \item Potential advantages of technique (1) increase in efficiency of
+%     propagation, (2) the new uses of the cached information.
+
+% \item Potential disadvantage. Reasons are recycled, in some circumstances
+%     new reasons might be better. We show in a simple experiment that
+%     this effect can occur.
+
+% \item Technique captures most of the positive effects of
+%     non-chronological backtracking in a simpler to implement way. It
+%     also preserves the simpler standard structure of CDCL
+%     solvers. Already provides some advantages for current SAT
+%     solvers. The ideas also have the potential for further
+%     exploitation. Lookahead for conflicts seems to be beneficial, and
+%     potentially if better discrimatory measures could be developed for
+%     reason clauses we could also better determine whether or not to
+%     use the cached information.. 
+% \end{enumerate}
+
+% \note{Reusing the Assignment Trail in CDCL Solvers: restart to level
+%   $i$ where the next decision has a higher score than the decision at
+%   level $i$ and a lower score than all decisions at level $j$ for
+%   $j< i$.
+
+%   They argum changes to reason clauses don't make much difference, but
+%   that was with minisat, before metrics for clause quality were in place in sat solvers.
+
+%   The savings for Luby-100 (used in minisat) were about 1\%}
 
@@
 current decision level \cite{DBLP:conf/dac/MoskewiczMZZM01}. Recently,
-however, it has been shown that backtracking chronologically
-\cite{DBLP:conf/lpar/JiangZ13} (i.e. backtracking one level only after
-conflict analysis)
+however, it has been shown that partial backtracking
+\cite{DBLP:conf/lpar/JiangZ13} or chronological backtracking, \cbt,
+(i.e. backtracking one level only after conflict analysis)
 \cite{DBLP:conf/sat/NadelR18,DBLP:conf/sat/MohleB19} can be effective
@@
 
-There are at least two possible side-effects of chronological
-backtracking which may explain its effectiveness. It has been observed
-that when a solver backtracks non-chronologically after a conflict,
-many of the literals that are unassigned during the backtrack will be
-re-assigned again in roughly the same order when the solver
-continues. By using chronological backtracking, the solver keeps a lot
-more of its partial assignment intact and saves from having to
-reconstruct a lot of the trail via propagation that it otherwise would
-have. Saving from having to reconstruct portions of the trail after
-backtracking was first introduced in the context of restarts by van
-der Tak et al. \cite{DBLP:journals/jsat/TakRH112}. The second
-side-effect of chronological backtracking is that it forces the solver
-to stay in the same local neighbourhood during search, since it
-unassigns less of the partial assignment and one could argue that this
-might lead to finding additional conflicts more quickly or reaching a
-satisfying assignment (if one exists) more quickly. However, this
-second side-effect also leads to less search flexibility, since the
-solver does not jump out of local neighbourhoods as often and does not
-obey the variable selection heuristic. We will argue that it is the
-first side-effect that leads to the effectiveness of chronological
-backtracking and not the second by developing a technique which also
-saves from having to repeat propagations, but allows the search to
-remain flexible and obey the variable selection heuristic.
-
-Our alternative method is "trail saving", where we cache the part of
-the trail that is unassigned during a backtrack and then attempt to
-restore implied parts of it as each new decision is made. Trail saving
-preserves the traditional invariants of the SAT solver and its basic
-version is very simple to implement. It also allows the search to
-choose the order of decisions, but helps make propagation faster. We
-explore the theoretical speedup that trail saving provides, develop
-some enhancements to make the idea more effective, and demonstrate
-experimentally that it performs similarly well as chronological
-backtracking without suffering from some of the aforementioned
-drawbacks.
+The motivation for using \cbt is the observation that when a solver
+backtracks across many levels, many of the literals that are
+unassigned during the backtrack will be re-assigned again in roughly
+the same order when the solver redescends. This observation was first
+made in the context of restarts by van der Tak et
+al. \cite{DBLP:journals/jsat/TakRH11}. Their technique backtracks to
+the minimum change level, i.e., the first level at which the solver's
+trail can change on redescent. However, their technique cannot be used
+when backtracking from a conflict: the solver's trail is going to be
+changed at the backtrack level so the minimum change level is the same
+as the backtrack level.
+
+Chronological backtracking or partial backtracking instead allows a
+reduction in the length of the backtrack by placing literals on the
+trail out of decision level order. By reducing the length of the
+backtrack the solver can keep more of its assignment trail
+intact. This can save it from the work involved in reconstructing a
+lot of its trail. Using \cbt is not a panacea however. Its application
+must be limited for peak effectiveness. This indicates that it is
+sometimes beneficial for the solver backtrack fully and redo its
+trail, even if this takes more work. We will expand on why this might
+be case below.
+
+In this paper we present an alternative ``trail savings'' method where
+by we save the backtracked part of the solver's trail and attempt to
+use that information to make the solver's redescent more
+efficient. Trail saving preserves the traditional invariants of the
+SAT solver and its basic version is very simple to implement. It
+allows the search to retain complete control over the order of
+decisions, but helps make propagation faster. We develop some
+enhancements to make the idea more effective, and demonstrate
+experimentally that it performs as well as and sometimes better than
+chronological backtracking. We also show that with the enhancements we
+are able to improve the performance of state of the art solvers.
 
 \section{Background}
-A propositional formula $\formula$ expressed in Conjunctive Normal
-Form (CNF), contains a set of variables $V$. A literal is a variable
-$v\in V$ or its negation $\lnot v$, and for a literal $l$ we let
-$\var(l)$ denote its underlying variable. A CNF consists of a
-conjunction of clauses, each of which is a disjunction of literals. We
-often view a clause as being a set of literals and employ set
-notation, e.g., $\ell\in C$ and $C'\subset C$.
-
-We assume the reader is familiar with the basic operations of CDCL
-\sat solvers. A good source for this background is
-\cite{DBLP:series/faia/SilvaLM09}.
+\label{sec:background}
+\sat solvers determine the satisfiability of a propositional formula
+$\formula$ expressed in Conjunctive Normal Form (CNF). $\formula$
+contains a set of variables $V$. A literal is a variable $v\in V$ or
+its negation $\lnot v$, and for a literal $l$ we let $\var(l)$ denote
+its underlying variable. A CNF consists of a conjunction of clauses,
+each of which is a disjunction of literals. We often view a clause as
+being a set of literals and employ set notation, e.g., $\ell\in C$ and
+$C'\subset C$. We will assume that the reader is familiar with the
+basic operations of CDCL \sat solvers. A good source for this
+background is \cite{DBLP:series/faia/SilvaLM09}.
 
@@
 operation a \sat solver will add newly assigned literals to the end of
-the trail, and on backtrack remove literals from the end of the trail
-(unassigning them as they are removed).
-
-A \sat solver's trail satisfies a number of conditions that we will
-discuss below. However, in we will need some additional flexibility in
-our definitions, as we will be sometimes be working with trails that
-would never be constructed by a \sat solver. Hence, we define a
-\emph{trail} to be a sequence of literals each of which is either a
-\emph{decision} literal or an \emph{implied} literal, and each of
-which has a \emph{reason}. These two types of literals are
-distinguished by their \emph{reasons}. Decision literals $d$ have a
-null reason, $\reason{d} = \varnothing$. Implied literals $l$ have as
-a reason a clause of the formula $\formula$,
-$\reason{l} = C\in\formula$. ($\reason{l}$ can be a learnt clause that
-has been added to $\formula$).
+the trail, and on backtrack remove literals from the end of the
+trail. For convenience, we will regard \emph{literals as having been
+  assigned \true if and only if they are on the trail}. So
+removing/adding a literal to the trail is equivalent to
+unassigning/assigning the literal $\true$.
+
+A \sat solver's trail satisfies a number of conditions. However, in
+this work we will need some additional flexibility in our definitions,
+as we will be sometimes be working with trails that would never be
+constructed by a \sat solver. Hence, we define a \emph{trail} to be a
+sequence of literals each of which is either a \emph{decision} literal
+or an \emph{implied} literal, and each of which has a
+\emph{reason}. These two types of literals are distinguished by their
+\emph{reasons}. Decision literals $d$ have a null reason,
+$\reason{d} = \varnothing$. Implied literals $l$ have as a reason a
+clause of the formula $\formula$, $\reason{l} =
+C\in\formula$. (The clause $\reason{l}$ can be a learnt clause that has been
+added to $\formula$).
 
@@
 $\trailIdxT{\trail}{\ell}$ denote its index on the trail, i.e,
-$\trail[\trailIdxT{\trail}{\ell}] = \ell$. For convenience, we also let
-$\trailIdxT{\trail}{\ell} = \trailIdxT{\trail}{\lnot \ell} =
-\trailIdxT{\trail}{\var(\ell)}$ even though neither $\lnot \ell$ nor
-$\var(\ell)$ are actually on $\trail$. If $x$ and $y$ are both on the
-trail and $\trailIdxT{\trail}{x} < \trailIdxT{\trail}{y}$ we say that
-\textit{$x$ appears before $y$ on the trail}. For convenience, when
-the trail being discussed in clear from context we simply write
-$\trailIdx{}$ instead of $\trailIdxT{\trail}{}$.
+$\trail[\trailIdxT{\trail}{\ell}] = \ell$. If $x$ and $y$ are both on
+the trail and $\trailIdxT{\trail}{x} < \trailIdxT{\trail}{y}$ we say
+that \textit{$x$ appears before $y$ on the trail}. For convenience,
+when the trail being discussed is clear from context we simply write
+$\iota$ instead of $\iota_{\trail}$.
 
@@
   \cite{DBLP:conf/lpar/JiangZ13,DBLP:conf/sat/NadelR18,DBLP:conf/sat/MohleB19}
-  generates trails with non-contigous decision levels} that starts
+  generates trails with non-contiguous decision levels} that starts
 with a decision literal $d_i$ and ends just before the next decision
@@
 $\trail[[\range{i}{j}]]$ denote the subsequence of literals at
-decision levels $k$ for $i\leq k\leq j$. Now we define the
-following properties that a trail $\trail$ can have.
+decision levels $k$ for $i\leq k\leq j$. 
+\begin{defn}
+    A clause $C$ has \textbf{been made unit by $\trail$ implying $l$}
+    when
+    $l\in C \land \bigl(\forall x\in C. x\neq l \implies \lnot x \in
+    \trail\bigr)$. That is, all literals in $C$ except $l$ must have
+    been falsified by $\trail$
+\end{defn}
+
+Now we define the following properties that a trail $\trail$ can have.
 \begin{description}
@@
 \item[reason-sound:] For each implied literal $l\in\trail$ we have
-    that it reason clause has \emph{been made unit by $\trail$
-      implying $l$}. This means that the reason clause contains $l$
-    and the negation all of its other literals are on $\trail$ before
-    $l$:
+    that its reason clause $\reason{l}=C$ has been made unit by
+    $\trail$ implying $l$, and for each $x\in C$ with $x\neq l$ we
+    that $\lnot x$ appears before $l$ on $\trail$:
     $\forall l\in\trail.\, \reason{l}\neq \varnothing \implies
@@
     \trailIdx{l}\bigr)$.
-\item[propagation-complete:] Unit propgation has been run to
+\item[propagation-complete:] Unit propagation has been run to
     completion at all decisions levels of $\trail$. This means that
@@
     C\in\formula. \mbox{$C$ is made unit by $\trail[[\range{0}{i}]]$
-      implying $l$}\bigr) \implies l\in \trail[[\range{0}{i}]]$. With
-    non-redundancy $l$ will appear only at the first decision level it
-    was implied, without non-redundancy, $l$ might appear more than
-    once. Note that this also implies that $\reason{l}$ must contain
-    at least one other literal $y\neq l$ with
-    $\dlevel{y} = \dlevel{l}$.
+      implying $l$}\bigr) \implies l\in \trail[[\range{0}{i}]]$. Note
+    that propagation completeness implies that
+    $\reason{l}\neq \varnothing$ must contain at least one other
+    literal $y\neq l$ with $\dlevel{y} = \dlevel{l}$.
 \item[conflict free:] No clause of $F$ is falsified by
@@
 to $\btL$ the second deepest level in $\uipcls$. This involves
-changing $\trail$ to its prefix $\trail[[\range{0}{\btL}]]$
-(unassigning all literals removed from $\trail$). The new clause
-$\uipcls$ is made unit by $\trail[[\range{0}{\btL}]]$ implying
-$\deepestLit$, so the solver then adds $\deepestLit$ to the trail and
-executes another round of unit propagation at level $\btL$, after
-which it continues by once again growing the trail with new decisions
-and unit implied literals until a new conflict or a satisfying
-assignment is found.
+changing $\trail$ to its prefix $\trail[[\range{0}{\btL}]]$ (by our
+convention all literals removed from $\trail$ are now unassigned). The
+new clause $\uipcls$ is made unit by $\trail[[\range{0}{\btL}]]$
+implying $\deepestLit$, so the solver then adds $\deepestLit$ to the
+trail and executes another round of unit propagation at level $\btL$,
+after which it continues by once again growing the trail with new
+decisions and unit implied literals until a new conflict or a
+satisfying assignment is found.
 
@@
 technique removes the redundant work of reproducing the first $M$
-trail levels. Unfortunately, with backtracking the trail will be
-changed at level $\btL$ ($\deepestLit$ will be newly inserted at this
-level). Thus there seems to be no choice but to backtrack to
-$\btL$. In this paper we will show that although we have to backtrack
-to $\btL$ we can make the subsequent redescent much more efficient.
+trail levels. When backtracking from a conflict, however, the trail
+will be changed at level $\btL$ ($\deepestLit$ will be newly inserted
+at this level). Hence this technique cannot reduce the length of the
+backtrack. In this paper we will show that although we have to
+backtrack to $\btL$ we can make the subsequent redescent much more
+efficient.
 
 \paragraph{Chronological Backtracking.}
-Chronological bactracking (\cbt) is a technique originally propsed by
+Chronological bactracking (\cbt) is a technique originally proposed by
 Jiang and Zhang \cite{DBLP:conf/lpar/JiangZ13} under the name
@@
 $\btL \leq j \leq \deepestLvl{-}1$ (where $\deepestLvl$ is the level
-the conflict was discovered. Nadel and Ryvchin
+the conflict was discovered). Nadel and Ryvchin
 \cite{DBLP:conf/sat/NadelR18} proposed to always backtrack to
@@
 \cite{DBLP:conf/sat/MohleB19} returned to the original proposal of
-flexibly backtracking any level in the allowed range. Note that the
+flexibly backtracking to any level in the allowed range. Note that the
 new 1-UIP clause $\uipcls$ is made unit at every level in this
 range. So after backtracking to level $j$ the newly implied literal
-$\deepestLit$ is added to the trail with
+$\deepestLit$ can be added to the trail with
 $\reason{\deepestLit}=\uipcls$, and $\dlevel{\deepestLit}$ is set to
@@
 literals at level $j$ (if $j\neq \btL$). This change has a number of
-consequences on the \sat solvers operation, all of which were
+consequences for the \sat solver's operation, all of which were
 described in the original paper \cite{DBLP:conf/lpar/JiangZ13}. In
@@
 developed under which the changes required to the \sat solver could be
-shown to be successful. We refer the reader to the cited papers for
+shown to be sound. We refer the reader to the cited papers for
 more details.
@@
 \cite{DBLP:conf/sat/NadelR18} \cbt was applied only when the length of
-the standard backtrack, $\btL-\deepestLvl$ was greater than a
-given threshold $T$. In their experiments they found that $T=100$ was
-the best value, i.e., \cbt is done only on longer backtracks. In
-practice, this meant that \cbt was \emph{hardly ever done}; in our
-measurements with their solver only about 3\% of the solver backtracks
-were \cbt backtracks. In \cite{DBLP:conf/sat/MohleB19} the value
-$T=100$ was also applied. However, they introduced an additional
-technique to add some applications of \cbt when the length of the
-backtrack is less than $T$. This allowed to utilize \cbt in \%15 of
-the backtracks. Nevertheless, the fact \cbt has to be used in a
-controlled fashion, indicates that can be benefits in the \sat solver
-employing standard backtracking; it cannot be the case that every
-redescent is only doing redundant work.
+the standard backtrack, $\btL-\deepestLvl$ was greater than a given
+threshold $T$. In their experiments they found that $T=100$ was the
+best value, i.e., \cbt is done only on longer backtracks. In practice,
+this meant that \cbt was \emph{hardly ever done}; in our measurements
+with their solver only about 3\% of the solver backtracks were \cbt
+backtracks. In \cite{DBLP:conf/sat/MohleB19} the value $T=100$ was
+also applied. However, they introduced an additional technique to add
+some applications of \cbt when the length of the backtrack is less
+than $T$. This allowed \cite{DBLP:conf/sat/MohleB19} to utilize \cbt
+in about \%15 of the backtracks. Nevertheless, the fact \cbt has to be used
+in a controlled fashion, indicates that there can be benefits in the \sat
+solver employing standard backtracking; it cannot be the case that
+every redescent is only doing redundant work.
 
@@
 from the newly learnt clause. And third, literals might be unit
-implied with different reasons. These changes could be sometimes be
-major, and they could have a large impact on future
-clauses. Unfortunately, it seems difficult to predict when these
+implied with different reasons. These changes could have a major
+impact on the future learnt clauses, and thus on the solver's
+efficiency. Unfortunately, it seems difficult to predict when these
 impacts might be large.
@@
 good notions of clause quality were known
-\cite{DBLP:conf/ijcai/AudemardS09}. Our empirical results contradict
-this, indicating that the literal reasons can have a significant
-impact.
+\cite{DBLP:conf/ijcai/AudemardS09}. Our empirical results indicate
+that once clause quality is accounted for changing the literal reasons
+can have a significant impact.
 
@@
     past level $2$ occurs would $x$ be restored to its correct level
-    $0$, and it would require inprocessing simplications to remove $x$
+    $0$, and it would require inprocessing simplifications to remove $x$
     from the learnt clauses.
@@
 In sum, although these impacts of \cbt might or might not be harmful
-to the \sat solver, they exist. Hence, one unappealing aspect of \cbt
-is that introduces additional inter-dependencies between components of
-the solver. For example, the second impact means that heuristics for
-choosing decision variables might work differently in the presence of
-\cbt; and the first impact means that different ways of computing
-clause quality using decision levels in the clauses might also work
-differently in the presence of \cbt. So although there are bound to be
-cases where \cbt is superior, we would like to have a method that
-addresses the original motivation of reducing redundant work while not
-changing the operation of standard backtracking. We present our
-proposal for such a method next.
+to the \sat solver, they exist. Hence, perhaps the most unappealing
+aspect of \cbt is that introduces additional inter-dependencies
+between components of the solver. For example, the second impact means
+that heuristics for choosing decision variables might work differently
+in the presence of \cbt; and the first impact means that different
+ways of computing clause quality using the decision levels in the
+clauses might also work differently in the presence of \cbt. So
+although there are bound to be cases where \cbt is superior, we would
+like to have a method that addresses the original motivation of
+reducing redundant work while not changing the operation of standard
+backtracking. We present our proposal for such a method next.
 
 \section{Trail Saving}
-Our approach is to save the trail, $\trail$, on backtrack, and to use
-the saved trail $\trailsave$ when the solver redesends to improve the
-effiency of propagations without affecting the decisions the solver
+Our approach is to save the trail $\trail$ on backtrack, and to use
+the saved trail $\trailsave$ when the solver redescends to improve the
+efficiency of propagations without affecting the decisions the solver
 wants to make. The saved trail $\trailsave$ also provides a secondary
@@
 Say that the solver is at $\deepestLvl$ where it has encountered a
-conflict. From the 1-UIP clause it learns,, $\uipcls$, it now has to
-backtrack to $\btL$. This is accomplished by calling the following
-subroutine with argument $\btL$.
-\begin{algorithmic}[1]
-    \Procedure{backtrack}{$\btL$}
-    \State $\forall \ell\in\trail[[\range{\btL{+}1}{\deepestLvl}]]$ unassign($\ell$)
-    \State $\forall \ell\in\trail[[\range{\btL{+}1}{\deepestLvl{-}1}]]$ $\reasonsave{\ell} = \reason{\ell}$
-    \State $\trailsave = \trail[[\range{\btL{+}1}{\deepestLvl{-}1}]]$
-    \State $\trail = \trail[[\range{0}{\btL}]]$
-    \EndProcedure
-\end{algorithmic}
-
-Note that we do not save the deepest level of $\trail$. The full
+conflict. From the 1-UIP clause it learns, $\uipcls$, it now has to
+backtrack to $\btL$. This is accomplished by calling
+\textsc{backtrack($\btL$)}, shown in Figure~\ref{fig:trailUse}, which
+saves the backtracked portion of the trail.
+
+Note that \bt does not save the deepest level of $\trail$. The full
 $\trail$ contains a conflict (at its deepest level). Hence the solver
@@
 $\trailsave$ we also save the clause reason of the unit implied
-literals in a $\mathit{reason}_{save}$ vector. Finally, we see that
-the first literal on $\trailsave$ must be a decision literal: it is
-the first literal of $\trail$ at decision level $\btL+1$.
-
-\paragraph{Using $\trailsave$ for more efficient unit propagations.}
+literals in a separate $\mathit{reason_{save}}$ vector. Finally, we
+see that after backtrack the first literal on $\trailsave$ is a
+decision literal: it is the first literal of $\trail$ at decision
+level $\btL+1$. Literals will be removed from $\trailsave$ during its
+use, but always in units of complete decision levels. So
+$\trailsave[0]$ will always be a (previous) decision literal.
+
 After backtrack the solver will add $\deepestLit$ to the end of the
-updated $\trail$ with $\reason(\deepestLit)=\uipcls$ and then invoke
-unit propagation which will add more literals to
-$\trail$. $\trailsave$ is exploited during propagation by calling the
-following version of propagation with the argument
-$\trailIdx(\deepestLit)$ (i.e., the trail index of the newly added
-implicant.
-
-\begin{algorithm}[t]
+updated $\trail$ with $\reason{\deepestLit}$ $\mbox{}=\uipcls$ and
+then invoke unit propagation. $\trailsave$ is exploited during
+propagation by the version of \textsc{propagate} shown in
+Figure~\ref{fig:trailUse}, which will initially be invoked with the
+argument $\trailIdx{\deepestLit}$ (i.e., the trail index of the newly
+added implicant). The saved trail be continually consulted during the
+solver's descent whenever unit propagation is performed. When
+backtrack occurs $\trailsave$ will be overwritten to store the new
+backtracked portion of $\trail$.
+
+\begin{figure}[!t]
 \begin{algorithmic}[1]
-    \Procedure{unitPropagate}{idx}
+    \Procedure{backtrack}{$\btL$}
+    \State $\forall \ell\in\trail[[\range{\btL{+}1}{\deepestLvl{-}1}]]$ $\reasonsave{\ell} = \reason{\ell}$
+    \State $\trailsave = \trail[[\range{\btL{+}1}{\deepestLvl{-}1}]]$\label{ln:btsave}
+    \State $\trail = \trail[[\range{0}{\btL}]]$
+    \EndProcedure
+\end{algorithmic}
+
+\begin{algorithmic}[1]
+    \Procedure{propagate}{idx}
     \While{idx $\mbox{} < \trail.\mathrm{size}$()}
-        \State $c\gets\mbox{}$ \textsc{useSavedTrail}()
+        \State $c\gets\mbox{}$ \ust()
         \If{($c\neq \varnothing$)} \textbf{return} $c$ \Comment{Found conflict from $\trailsave$}
@@
             \If{($c$ is unit implying $x$)}
-                 \State $\trail.\mathrm{addToEnd}(x)$
-                 \State $\reason{x} \gets c$            
+                 \State $\trail.\mathrm{addToEnd}(x)$; \; $\reason{x} \gets c$            
             \ElsIf{($c$ is falsified)}\:
@@
     \Procedure{useSavedTrail}{\mbox{}}
-    \State $idx\gets 0$; $\litsave\gets \trailsave[idx]$
-    \State \qquad Stop at decision literal on $\trailsave$ not yet assigned \true in $\trail$
-    \While{($\reasonsave{\litsave} \neq \varnothing \lor \litsave \in \trail$)}
-          \If{($\lnot \litsave\in\trail$)}\: \textbf{return} $\reasonsave{\litsave}$ 
-                \Comment{Conflict: $\litsave$ already \false}
-          \ElsIf{($\litsave \not\in\trail$)} \Comment{$\litsave$ unassigned}
-              \State $\trail.\mathrm{addToEnd(\litsave)}$
-              \State $\reason{\litsave}\gets \reasonsave{\litsave}$
+    \State $idx\gets 0$; $c\gets \varnothing$
+    \For{( ; $idx < \trailsave.\mathrm{size}()$; $idx$++)}
+        \State $\litsave\gets \trailsave[idx]$
+        \If{($\reasonsave{\litsave} = \varnothing$)} \Comment{Decision on $\trailsave$}
+            \If{($\litsave\in\trail$)}\: \textbf{continue} \Comment{\true in solver, we can use its implied lits}
+            \Else{}\: \textbf{break} \Comment{Only solver can set decisions, so we stop here}\label{ln:decnottrue}
             \EndIf
-          \State $idx$++
-    \EndWhile
+        \Else{} \Comment{Implied Literal on $\trailsave$}
+             \If{($\litsave \in \trail$)}\: \textbf{continue}  \Comment{ignore redundant lits}\label{ln:skip}
+             \ElsIf{($\lnot \litsave \in \trail$)} 
+                \Comment{contradiction, return conflict and reset $idx$}\label{ln:conflict}
+                 \State $c\gets\reasonsave{\litsave}$; $idx\gets 0$; \textbf{break}
+             \Else{} \Comment{unset in solver, add to solver's trail}
+                  \State $\trail.\mathrm{addToEnd(\litsave)}$\label{ln:addToTrail}
+                  \State $\reason{\litsave}\gets \reasonsave{\litsave}$\label{ln:addToTrailNext}
+              \EndIf
+        \EndIf
+    \EndFor
+    \For{($i \gets 0$; $i < idx$; $i$++)} \Comment{$\trailsave$ is unchanged when $idx=0$}
+        \State $\trailsave.\mathrm{removeFront()}$\label{ln:popfront}
+    \EndFor
+    \State \textbf{return} $c$
     \EndProcedure
@@
 
-\caption{Using $\trailsave$ to enhance unit propagation and conflict
-  detection}
-\end{algorithm}
-
-\note{
+\caption{Using $\trailsave$ in unit propagation and conflict
+  detection} \label{fig:trailUse}
+\end{figure}
+
+$\trailsave$ is consulted in the procedure $\ust$
+(Figure~\ref{fig:trailUse}). This procedure tries to add saved implied
+literals and their reasons to the solver's trail, when these
+implications are valid. We will show below that those implications
+that are added are in fact valid. We do not interfere with the
+solver's variable decisions. Instead we opportunistically test to see
+if literals implied on $\trailsave$ are valid implications for the
+solver given the solver's current decisions.
+
+$\trailsave[0]$ is always a (previous) decision literal $d$ with
+$\reasonsave{d}=\varnothing$. Note that, since new literals (e.g.,
+$\deepestLit$) have been added to $\trail$, $d$ might now be an
+implied literal on $\trail$ (i.e., $\reason{d}\neq\varnothing$) even
+though before the backtrack it was previously a decision (i.e.,
+$\reasonsave{d}=\varnothing$). If $d$ has not been assigned $\true$ by
+the solver (i.e., $\lnot d \in \trail$), we cannot add any implied
+literals below it on $\trailsave$ to $\trail$ as these implied
+literals depend on $d$ being assigned $\true$. In this case we stop
+looking for more literals to add to $\trail$
+(line~\ref{ln:decnottrue}).
+
+On the other hand if $d$ has been made \true by the solver we can
+continue to add all of the implied literals below it (up to but not
+including the next decision literal on $\trailsave$) to $\trail$
+(line~\ref{ln:addToTrail}), reusing their saved reasons. Any literals
+that have already been made true by the solver can be skipped
+(line~\ref{ln:skip}). Finally, if we encounter a literal that has
+already been falsified by the solver, then its saved reason clause
+must be falsified by the solver and we can return it as a conflict
+(line~\ref{ln:conflict}). If a conflict is encountered we leave
+$\trailsave$ unchanged by resetting $idx$ to zero. Otherwise, $idx$
+will be the number of literals at the front of $\trailsave$ that have
+been moved to $\trail$ (or skipped over since they are already on
+$\trail$). We then remove the first $idx$ literals from $\trailsave$
+(line~\ref{ln:popfront}), and return the conflict (equal to
+$\varnothing$ if no conflict was found).
+
+\begin{figure}[!t]
+\begin{tabular}{l
+  >{\raggedleft}p{4.75ex}>{\raggedleft}p{4.75ex}>{\raggedleft}p{4.75ex}>{\raggedleft}p{4.75ex}
+  >{\raggedleft}p{4.75ex}>{\raggedleft}p{4.75ex}>{\raggedleft}p{4.75ex}>{\raggedleft}p{4.75ex}
+  >{\raggedleft}p{4.75ex}>{\raggedleft}p{4.75ex}>{\raggedleft}p{4.75ex}>{\raggedleft}p{4.75ex}
+  >{\raggedleft}p{4.75ex}>{\raggedleft}p{4.75ex}}
+      $\trail$ 
+      & $l_1^{1*}$ & $l_2^1$ & $l_3^{2*}$ & $l_4^2$
+      & $l_5^{3*}$ & $l_6^3$ & $l_7^{4*}$ & $l_8^4$
+      & $l_9^{5*}$ & $l_{10}^5$ & $l_{11}^5$ & $l_{12}^{6*}$
+      & $l_{13}^6$ & $l_{14}^6$  \tabularnewline
+     $\trailsave$
+      &  &  &  & 
+      &  &  &  &  
+      &  &  &  &  
+      &  & \tabularnewline
+      \rowcolor{linen}
+      \multicolumn{15}{l}{1-UIP ($\lnot l_1$, $\lnot l_3$, $\lnot l_{12}$) and backtrack 2 ($\btL$)}\tabularnewline
+      $\trail$ 
+      & $l_1^{1*}$ & $l_2^1$ & $l_3^{2*}$ & $l_4^2$
+      &         &       &         &      
+      &         &       &         &           
+      &         &         \tabularnewline
+      $\trailsave$
+      &         &       &         &                     
+      & $l_5^*$ & $l_6$ & $l_7^*$ & $l_8$
+      & $l_9^*$ & $l_{10}$ & $l_{11}$ & 
+      &         & 
+      \tabularnewline
+      \rowcolor{linen}
+      \multicolumn{15}{l}{Unit Prop $\lnot l_{12}$ ($\deepestLit$), $\trailsave$ not yet helpful}\tabularnewline
+      $\trail$ 
+      & $l_1^{1*}$ & $l_2^1$ & $l_3^{2*}$ & $l_4^2$
+      & $\smash{\lnot} l_{12}^2$  & $l_7^2$  & $l_9^2$    &      
+      &         &       &         &           
+      &         &         \tabularnewline
+      $\trailsave$
+      &         &       &         &                     
+      & $l_5^*$ & $l_6$ & $l_7^*$ & $l_8$
+      & $l_9^*$ & $l_{10}$ & $l_{11}$ &
+      &         & 
+      \tabularnewline
+      \rowcolor{linen}
+      \multicolumn{15}{l}{Make decision}\tabularnewline
+      $\trail$ 
+      & $l_1^{1*}$ & $l_2^1$ & $l_3^{2*}$ & $l_4^2$
+      & $\smash{\lnot} l_{12}^2$  & $l_7^2$  & $l_9^2$ 
+      & $l_5^{3*}$  &       &         &           
+      &         &         \tabularnewline
+      $\trailsave$
+      &         &       &         &                     
+      & $l_5^*$ & $l_6$ & $l_7^*$ & $l_8$
+      & $l_9^*$ & $l_{10}$ & $l_{11}$ & 
+      &         & 
+      \tabularnewline
+      \rowcolor{linen}
+      \multicolumn{15}{l}{Now $\trailsave$ can be used to augment trail}\tabularnewline
+      $\trail$ 
+      & $l_1^{1*}$ & $l_2^1$ & $l_3^{2*}$ & $l_4^2$
+      & $\lnot l_{12}^2$ \hfill & $l_7^2$  & $l_9^2$      
+      & $l_5^{3*}$  & $l_6^{3}$  & $l_8^3$  & $l_{10}^3$          
+      & $l_{11}^3$  &         \tabularnewline
+      $\trailsave$
+      &         &       &         &      
+      &         &       &                              
+      &         &       &         & 
+      &         &       & 
+    \end{tabular}
+    \caption{Use of $\trailsave$ from
+      Example~\ref{ex:trailsave}\label{fig:trailsave}. The literals
+      decision level is indicated in its superscript, and a $\mbox{}^*$
+      superscript indicates that the literal is a decision.}
+\end{figure}
+
+\begin{example}
+    \label{ex:trailsave} Figure~\ref{fig:trailsave} provides an
+    example of how $\trailsave$ is used. Initially the literals $l_1$
+    to $l_{14}$ are on the solvers $\trail$, and $\trailsave$ is
+    empty. This is shown in the first two lines of the figure. In the
+    figure the superscript on the literals indicates their decision
+    level, and a superscripted $*$ indicates that the literal is a
+    decision. Hence $l_1^{1*}$ indicates that $\dlevel{l_1}=1$ and
+    that $l_1$ is a decision.
+
+    Then a conflict is found at level $6$ and the 1-UIP clause
+    $(\lnot l_1, \lnot l_3, \lnot l_{12})$ is learnt. Thus the solver
+    will backtrack to level $2$, where it will add $\lnot l_{12}$ as a
+    unit implicant. The next two lines shows $\trail$ and $\trailsave$
+    right after the backtrack to level $2$: the backtracked levels
+    have been copied into $\trailsave$ omitting the conflict level $6$.
+
+    The new unit $\lnot l_{12}$ is now added to $\trail$ and unit
+    propagation performed adding $l_7$ and $l_9$ to level 2. Since the
+    first literal on $\trailsave$, $l_5$, has
+    $\reasonsave{l_5}=\varnothing$ ($l_5$ was a decision on $\trail$
+    at the time backtrack occurred) and is not yet $\true$,
+    $\trailsave$ is not helpful at this stage. The status of $\trail$
+    and $\trailsave$ at this point is shown in the figure.
+
+    After unit propagation is finished the solver makes a new
+    decision, which happens to be (but is not forced to be) $l_5$. Now
+    $\trailsave$ can be used: $l_5$ is \true so it is removed, $l_6$
+    is unassigned so it is added to $\trail$, $l_7$ is \true and so
+    removed, $l_8$ is unassigned so it is added to $\trail$, $l_9$ is
+    \true and removed, and finally $l_{10}$ and $l_{11}$ are
+    unassigned and so are added to $\trail$. In this example,
+    $\trailsave$ is emptied, and cannot contribute more to $\trail$.
+
+    All of these units are added to $\trail$ before the solver starts
+    to unit propagate $l_5$. Since, new literals have been added to
+    $\trail$ before $l_5$ the solver must propagate $l_5$ and all of
+    the literals that follow it before making its next
+    decision. 
+\end{example}
+
+As noted in the previous example unit propagation has to be rerun on
+all saved literals added to $\trail$ from $\trailsave$. Thus our
+technique, unlike \cbt, does not completely remove the overhead of
+reproducing the trail on the solver's redescent. Nevertheless, trail
+savings improves the efficiency of this redescent in three different
+ways. First, by adding more forced literals to the trail before
+continuing propagating the next literal, propagation can potentially
+gain a quadratic speedup
+\cite{DBLP:conf/sat/HickeyB19,DBLP:journals/jair/Gent13}. Second,
+propagation does not need to examine the reason clause of the added
+literals. If these literals were not added by \ust, propagation would
+have to traverse each of these reason clauses to determine that they
+are have in fact become unit. Third, when a conflict is returned by
+\ust all further propagations can be halted. The added literals and
+their reasons will be sufficient to perform clause learning from the
+conflict returned by \ust. Since trail saving can sometimes save
+hundreds or thousands of literals at a time these savings can in sum
+be significant.
+
+% \note{
+%   \begin{enumerate}
+%   \item Use simple backtrack and save to motivate and state the
+%       invariant. 
+%   \item Introduce notation for concatenating two trails.
+%   \item Name and state the invariant in text (e.g. a
+%       defn)---concenating the cached trail with the current trail
+%       yields a trail that is always reason valid, but not necessarily
+%       non-contradictory, non-conflicting, nor propagation complete).
+%   \item Proof (simple) that the invariant holds for caching on
+%       backtrack.
+%   \item On backtrack at least the asserted literal and any new
+%       propagants at the backtrack level are added.
+%   \item Prove that the invariant holds (???do we need an invariant or
+%       simply a property of the augmented trail) for trails with extras
+%       inbeween.
+%       $T_{\mathit{old}} + T_{\mathit{newlyAdded}} +
+%       T_{\mathit{saved}}$ also satisfies the invariant
+%   \item re-acheving propagation completeness for the augmented trail
+%       (a) only add one level of the cached trail at a time, and do
+%       unit prop after (b) only add non-contradicted decision from the
+%       old trail. Also note that conflicts are valid and can be learnt
+%       from even if the augmented trail is not propagation complete. 
+%   \item Finally figure out a way to formalize the extra technique of
+%       storing on top after repeated backtracking. 
+%   \item Explain how freedom for the search is maintained by makeing
+%       decisions then consulting the trail.
+%   \item 
+%   \end{enumerate}
+% }
+
+\subsection{Correctness}
+Now we will prove that our use of $\trailsave$ preserves the \sat
+solver's soundness. In particular, $\trailsave$ is only used in the
+procedure \ust, in which it either adds new literals
+to the solver's trail, or returns conflict clauses to the solver.
+Hence, we only need to show that these new literals are in fact unit
+implied and the conflicts are in fact falsified by the solver's trail.
+\
+Since both $\trail$ and $\trailsave$ are sequences of literals (with
+associated reasons) we can consider their concatenation denoted as
+$\trail + \trailsave$.
+\begin{theorem}
+    \label{thm:sound}
+    If $\trail + \trailsave$ is \textbf{reason sound}
+    (Section~\ref{sec:background}) then the following holds.  If the
+    first $i$ literals on $\trailsave$ are all in $\trail$
+    ($\forall j. 0\leq j< i. \trailsave[i]\in \trail$) and
+    $\trailsave[i]=l$ is an implied literal with $\reasonsave{l} = C$,
+    then $C$ has been made unit by $\trail$ implying $l$.
+\end{theorem}
+\noindent\emph{Proof:}
+Since $\trail + \trailsave$ is reason sound, every literal in $C$
+other than $l$ appears negated before $l$ in the sequence
+$\trail + \trailsave$. Thus for $x\in C$ we have $\lnot x \in \trail$
+or $\lnot x \in \trailsave[0]\ldots\trailsave[i-1]$. But in the later
+case we also $\lnot x \in \trail$. 
+\whitebox
+
+This theorem shows that \ust's processing is
+sound. In this procedure, an implied literal from $\trailsave$ is added
+to $\trail$ (line~\ref{ln:addToTrail}) only when all prior literals on
+$\trailsave$ are already on $\trail$ (i.e., previously on $\trail$ or
+already added to $\trail$). Thus each new addition is sound given the
+inductive soundness of the previous additions, with the base case
+covered by Theorem~\ref{thm:sound}. If $l$ is to be added, the theorem
+shows that every other literal in $\reasonsave{l}$ has been falsified
+by $\trail$. Hence if $l$ is also falsified by $\trail$ then
+$\reasonsave{l}$ is a clause that is falsified by $\trail$, thus it is
+a sound conflict for the solver.
+
+Now we only have to show that $\trail+\trailsave$ is always
+\textbf{reason sound} during the operation of the solver.
+\begin{prop}
+    \label{prop:sound}
+    If $\trail+\trailsave$ is reason sound then $\trail'+\trailsave'$
+    is reason sound in all of the following cases.
     \begin{enumerate}
-  \item Use simple backtrack and save to motivate and state the
-      invariant. 
-  \item Introduce notation for concatenating two trails.
-  \item Name and state the invariant in text (e.g. a
-      defn)---concenating the cached trail with the current trail
-      yields a trail that is always reason valid, but not necessarily
-      non-contradictory, non-conflicting, nor propagation complete).
-  \item Proof (simple) that the invariant holds for caching on
-      backtrack.
-  \item On backtrack at least the asserted literal and any new
-      propagants at the backtrack level are added.
-  \item Prove that the invariant holds (???do we need an invariant or
-      simply a property of the augmented trail) for trails with extras
-      inbeween.
-      $T_{\mathit{old}} + T_{\mathit{newlyAdded}} +
-      T_{\mathit{saved}}$ also satisfies the invariant
-  \item re-acheving propagation completeness for the augmented trail
-      (a) only add one level of the cached trail at a time, and do
-      unit prop after (b) only add non-contradicted decision from the
-      old trail. Also note that conflicts are valid and can be learnt
-      from even if the augmented trail is not propagation complete. 
-  \item Finally figure out a way to formalize the extra technique of
-      storing on top after repeated backtracking. 
-  \item Explain how freedom for the search is maintained by makeing
-      decisions then consulting the trail.
-  \item 
+    \item $\trailsave[0]\in \trail$, $\trail'=\trail$, and
+        $\trailsave' = \trailsave.\mathrm{removeFront()}$.
+    \item $\trail' = \trail+\trailsave[0]$ and
+        $\trailsave' = \trailsave.\mathrm{removeFront()}$.
+    \item $\trail' = \trail + \trail_{\mathit{new}}$ and
+        $\trailsave' = \trailsave'$ and $\trail'$ is reason sound.
     \end{enumerate}
-}
-
-It can be observed that a SAT solver often repeats the same
-propagations over and over again during execution. Although no two
-descents down the trail will ever be identical because of clause
-learning, the solver will often propagate many of the same literals
-with the same reasons in a similar order to what it had done
-previously, especially since the order of the decisions usually
-remains similar. We plan to cache the part of the trail that we are
-backtracking over and use it to restore some of those literals without
-propagating them from scratch.
-
-\subsection{Storing the trail cache}
-
-The first step that is needed to use "trail saving" is to store all of
-the literals which are being unassigned during a backtrack, along with
-their respective reason clauses, in a "trail cache". Decisions should
-be marked with a null reason. The literals should appear in the same
-order on the trail cache as they had appeared on the trail. We will
-keep a pointer, which we will refer to as the "top" of the trail
-cache, and initially set it to be the literal which had appeared the
-earliest on the trail. Note that initially the top of the trail cache
-always has a null reason (i.e., it was the start of a new decision
-level on the previous trail). As the literals from the trail cache get
-restored (i.e., assigned), the top of the trail cache will be updated
-to the next literal not yet restored.
-
-\begin{figure}\center\includegraphics[scale=0.7]{figures/trail_diagram.pdf}\caption{Storing into a trail cache after backtracking.}\end{figure}
-
-\subsection{Restoring the trail cache}
-We will base trail saving on the following critical invariant. If the
-top of the trail cache becomes true at any point, even after an
-arbitrary number of decisions and propagations have been made since
-the last backtrack, the set of literals on the trail cache after the
-top and up to but excluding the next literal with a null reason is
-still implied by the current trail. The top of the trail cache can
-become true either by decision or unit propagation, and this invariant
-still holds.\newline
-
-\textit{Proof 1} We will refer to the set of literals that become
-unassigned during the backtrack after conflict analysis has taken
-place as $L_u$, and will refer to the rest of the trail that remains
-assigned as $L_0$. We will refer to all of the literals that get
-placed on the trail after $L_0$ during subsequent propagations and
-decisions as $L_1$. We will refer to the current trail $L_0 \cup L_1$
-as $L_T$. The trail cache will contain the literals in $L_u$ ordered
-by their appearance on the old trail. Let $L_c[i]$ denote the set of
-the first i literals on the trail cache. Let $L_c[top]$ be the top of
-the old trail, which gets updated as parts of the trail cache get
-restored.
-
-The $i$th literal on the trail cache is implied by
-$L_0 \cup L_c[i-1]$, as long as the $i$th literal has a non-null
-reason, since $L_0 \cup L_c[i-1]$ had forced $L_c[i]$ on the previous
-trail. $L_0 \subseteq L_T$, since $L_T = L_0 \cup L_1$. Therefore, the
-$i$th literal on the trail cache is also implied by
-$L_T \cup L_c[i-1]$.
-
-Assume $L_c[top]$ is true, which means that
-$L_c[0], ..., L_C[top] \subseteq L_1 \subseteq L_T$. Then, from above,
-every literal on the trail cache up to but excluding the next literal
-with a nun-null reason would be implied. $\blacksquare$\newline
-
-Once the backtrack has been completed and the solver resumes with
-decisions and propagations, we will attempt to restore parts of the
-trail cache automatically during propagation to make it faster by
-using the invariant stated above. Each time the propagation routine is
-about to propagate a new literal, we first check the top of the trail
-cache to determine its truth value. If that literal's truth value is
-already true, we assign all of the literals that come after it on the
-trail cache to be true up until (and excluding) the next literal with
-a null reason.
-
-A potentially quadratic speedup can be realized during the propagation
-of these literals since they are being propagated as a set, rather
-than one at a time, as argued in \cite{DBLP:conf/sat/HickeyB19,DBLP:journals/jair/Gent13}. Restoring literals from the trail cache
-automatically not only leads to quicker propagation of future
-literals, but also means that the propagation routine has less work to
-do as it searches for new watchers in clauses on watchlists. All of
-the reasons attached to these literals that are automatically restored
-will now be satisfied, thus the propagation routine can skip over them
-rather than having to access each of them potentially multiple times
-to find they are unit.
-
-If at any time the top of the trail cache is a literal $l$ with a
-non-null reason whose truth value is already false, we determine a
-conflict has occurred with the conflict clause being $l$'s reason from
-the trail cache. This is because we know from Proof 1 that $l$ is
-implied to be true at this point by the reason on the trail cache
-(i.e., the reason on the trail cache is unit), but $\lnot l$ was also
-previously assigned to be true, therefore we have a conflict. This
-leads to a third potential speed up, since a conflict has been
-detected and none of the pending propagations have to be
-completed. Since trail saving can sometimes save hundreds or thousands
-of literals at a time, the number of propagations required to detect a
-conflict could be very costly.
-
-% \begin{algorithm}
-%   	\TitleOfAlgo{propagate()}
-% 	\begin{algorithmic}[1]
-%         \While{queue not empty}
-%         \Statex
-%         \State restoreTrailCache()
-%         \Statex
-%         \State $l := $ next literal on queue
-%         \For{each clause $c \in $ watchlist($\neg l$)}
-% 		\For{each literal $x \in c$}
-%         \If{$val($x$) \neq$ false and x is not already a watcher of c}
-%         \State let $x$ be a watcher of $c$ and add $c$ to watchlist($\neg x$)
-%         \State break
-%         \EndIf
-% 		\EndFor
-% 		\If{no new watcher was found}
-%         \If{other watcher $x_2$ of $c = $ false}
-%         \State return c as conflict
-%         \ElsIf{other watcher $x_2$ of $c$ is unassigned}
-%         \State enqueue $x_2$
-%         \EndIf
-% 		\EndIf
-%         \EndFor
-%         \EndWhile
-% 	\end{algorithmic}
-% \end{algorithm}
-
-% \begin{algorithm}
-%   	\TitleOfAlgo{restoreTrailCache()}
-% 	\begin{algorithmic}[1]
-%         \If{trail saving turned on}
-% 		\State{let tc $=$ top of trail cache}
-% 		\If{val(top of trail cache) $=$ true}
-%         \While{trail cache not empty}
-%         \State{move top of trail cache (tc) to next literal}
-%         \If{reason(tc) is null}
-%         \State{break}
-%         \ElsIf{val(tc) $=$ false}
-%         \State{return conflict with reason(tc)}
-%         \ElsIf{val(tc) $=$ unassigned}
-%         \State{enqueue tc with reason reason(tc)}
-%         \EndIf
-%         \EndWhile
-% 		\EndIf
-%         \EndIf
-% 	\end{algorithmic}
-% \end{algorithm}
+    4. We also have that $\trail$ is reason sound if $\trail$ was
+    generated by the solver.
+\end{prop}
+\noindent \emph{Proof:} (1) $\trailsave[0]$ already appears earlier in
+the $\trail$ so it can be removed without affecting the soundness of
+any reason following it. (2) is obvious as the sequence is
+unchanged. (3) the reasons in $\trail+\trail_{\mathit{new}}$ are sound
+by assumption. Those in $\trailsave$ remain sound as they depend only
+on the literals in $\trail$ and prior literals on $\trailsave$, both
+of which are unchanged. (4) is obvious from the operation of unit
+propagation in the solver.  \whitebox
+
+\begin{theorem}
+    \label{thm:rsound}
+    $\trail+\trailsave$ is always reason sound during the operation of
+    the solver.
+\end{theorem}
+\noindent \emph{Proof:} $\trailsave$ starts off being empty, so
+$\trail+\trailsave = \trail$ is reason sound as it was generated by
+the solver (4). In procedure \bt $\trail+\trailsave$ is
+set to a trail that was previously generated by solver (4). The solver
+can add to $\trail$ by decisions and propagations with out using
+$\trailsave$. In this case $\trail' = \trail+\trail_{\mathit{new}}$,
+and $\trail'$ is reason sound by (4), thus the new
+$\trail'+\trailsave$ is reason sound by (3). Finally, in procedure
+\ust either (a) literals at the front of
+$\trailsave$ are discarded since they already appear on $\trail$, or
+(b) literals are moved from $\trailsave$ to $\trail$. Under both of
+these changes $\trail+\trailsave$ remains reason sound by (1) and (2).
+\whitebox
 
 \subsection{Enhancements}
-
-In the standard version of trail saving outlined above, the trail
-cache needs to be deleted after each backtrack in order to make room
-for the new backtracked literals to be cached. The first enhancement
-to trail saving is to keep the trail cache intact instead of clearing
-it before each backtrack, and to add the literals that will be
-backtracked over to the beginning of the already existing trail
-cache. In order to preserve soundness, we must add the literals from
-the current backtrack to the beginning of the trail cache, such that
-the trail cache has as its prefix the literals which are being
-backtracked over, and as its suffix the previous trail cache. \newline
-
-\textit{Proof 2} We will use all of the same set names as Proof 1 and
-also refer to the set of literals on our previous trail cache as
-$L_{pc}$. We concatenate the literals in $L_{pc}$ to the end of $L_c$
-to get $L_{sc}$. We know that $L_{pc}$ is a valid trail cache for
-$L_0 \cup L_u$. If we reach the end of $L_c$ without any conflicts, we
-know that $L_u \subseteq L_T$. Since $L_0 \subseteq L_T$, then
-$L_u \cup L_0 \subseteq L_T$ and therefore $L_{pc}$ is a valid trail
-cache for $L_T$. $\blacksquare$\newline
-
-In order for the previous trail cache to be of any use, one must
-actually never store the last level of the current trail to the trail
-cache at all, since a learnt clause prevents the last level from being
-duplicated in the future. If any part of the previous trail cache was
-used to construct this last level, it must also be discarded from the
-previous trail cache before the previous trail cache is concatenated
-to the end.
-
-When concatenating trail caches together as above, it is important to
-note that the running trail cache can grow indefinitely. The trail
-cache can be trimmed periodically in the following way. Traverse down
-the trail cache and whenever a literal appears more than once, delete
-its entry in the trail cache. Whenever a literal $x$ appears whose
-complement $\lnot x$ has already appeared on the trail cache, $x$
-should be deleted and everything remaining until the end of the trail
-cache should be deleted (in order to keep the trail cache free of
-conflicts).  We decided to trim the trail cache whenever its size
-exceeded twice the total number of variables in the SAT instance, as
-that is what worked best in preliminary experiments.\newline
-
-The next enhancement to trail saving is to scan the trail cache to
-check if there is a literal that is already assigned false within some
-limit. The limit we found worked best is to scan the trail until the
-third null reason is found (i.e. "looking ahead 2 decisions"). If a
-literal that is already assigned false is found, we know that we are
-guaranteed to have a conflict in the near future, and we force the SAT
-solver's subsequent decisions to come from the trail cache instead of
-using the regular variable selection heuristic.
-
-Sometimes using trail saving to automatically restore literals with
-their previous reasons will lead to a literal getting a reason that is
-"worse" (in terms of number of literals, but one could use a different
-metric) than the reason it would have ended up with under normal
-propagation (i.e. without trail saving). The last enhancement we made
-to trail saving was to stop restoring literals from the old trail as
-soon as we hit a reason that we suspected might be "bad". For this, we
-kept a running mean and standard deviation of the length of all
-reasons used in the solver so far, and decided to stop restoring from
-the trail whenever we hit a reason that was greater in length than two
-standard deviations above the mean length. This way, when normal
-propagation resumes, we give it a chance to find a better reason than
-the one that trail saving had stored from the last descent down the
-trail.
-
-Another potential enhancement one could use is to use the trail cache
-to try to learn a second clause automatically when the solver reaches
-a conflict. However, we did not find a way to implement this to make
-it beneficial and the number of levels we have to scan for a second
-conflict would have to be more than the number of levels we already
-look ahead on each decision (when the solver is not under conflict).
-
-\section{Related Work}
-\subsection{Trail Saving for Assumption-Based SAT}
-Trail saving was briefly introduced in the context of assumption-based
-SAT \cite{DBLP:conf/sat/HickeyB19} in order to speed up the
-propagation of the assumptions each time the solver backtracked past
-the assumption levels. Note that the process is the same, but it only
-works to restore the propagants and implicants of the assumptions and
-only after a learnt unit clause. Becuase the number of propagants and
-implicants of the assumptions tends to become dominated by the number
-of assumptions in the applications it was tested on, the technique did
-not yield a significant improvement. It would be interesting to
-combine the extended version of trail saving outlined in this paper
-with trail saving over the assumptions to see if that could produce
-better results.
-
-\subsection{Comparison to Chronological Backtracking}
-As mentioned earlier, chronological backtracking also achieves some of
-the theoretical speedups that trail saving achieves. One advantage of
-chronological backtracking is that it does not require any caching of
-the trail or accessing such a cache during propagation; it simply
-leaves a vast portion of the trail intact rather than backtracking all
-the way to the conflict level. Chronological backtracking will also
-speed up propagations slightly more than trail saving, because it
-leaves more literals assigned at once which leads to clauses being
-detected as unit more quickly.
-
-One disadvantage that chronological backtracking has compared to trail
-saving is that it doesn't allow the search as much flexibility to jump
-out of local neighbourhoods, as the common variable selection
-heuristics tend to depend on
-\cite{DBLP:conf/dac/MoskewiczMZZM01,DBLP:conf/sat/2015,DBLP:conf/sat/LiangGPC16}. Because
-chronological backtracking leaves a large portion of the trail
-assigned, it will behave similarly to how the solver would behave if
-it was to force the decisions that do occur within this large portion
-of literals, whereas trail saving does not force any decisions. As
-variable selection heuristics become more refined, it is likely that
-adhering to them will become more crucial to performance.
-
-Another disadvantage is that, because propagation is not performed at
-every decision level, literals that remain assigned during
-chronological backtracking (but otherwise would have been unassigned)
-do not have a chance to move their decision levels up. An example of
-this is when the solver was supposed to backtrack over a literal that
-would end up being unit propagated at level 0, but stays assigned at
-its current decision level when chronological backtracking is
-used. This could potentially effect the lbd score or length of the
-learnt clause that gets generated at the next conflict. Additionally,
-chronological backtracking breaks the invarant that all literals
-appear on the trail in monotonically increasing order by decision
-level, which makes its implementation slightly more complicated.
+We developed three enhancements of the base trail saving method
+described above. In this section we present these enhancements.
+
+\paragraph{Saving the trail over multiple backtracks.}
+It can often be the case that when the solver finds a conflict and
+backtracks to $\btL$ it might immediately find a another conflict at
+$\btL$ causing a further backtrack. In the procedure \bt every
+backtrack causes $\trailsave$ to be overwritten. Hence, in these cases
+most of the trail will not be saved---only the portion from the last
+backtrack. Our first extension addresses this potential issue and also
+provides more general trail savings in other contexts as well.
+
+This extension is simply to add the latest backtrack to the front of
+$\trailsave$ leaving all of the previous contents of $\trailsave$
+intact. Specifically, we replace line~\ref{ln:btsave} of \bt by the new line:\\
+\hspace*{3em}3. \qquad $\trailsave = \trail[[\range{\btL{+}1}{\deepestLvl{-}1}]] + \trailsave$\\
+It is not difficult to show that this change preserves soundness. Only
+Theorem~\ref{thm:rsound} is potentially affected. However, we know
+that $\trailsave$ is unchanged at the level at which a conflict
+occurs: either the conflict is detected without consulting
+$\trailsave$ or if the conflict comes from $\trailsave$ then \ust
+leaves $\trailsave$ unchanged (line~\ref{ln:conflict}). Hence, at the
+level before the conflict occurred we have inductively that
+$\trail[[\range{0}{\deepestLvl{-}1}]]+\trailsave$ was reason sound,
+and hence so is $\trail'+\trailsave'$ with
+$\trail' = \trail[[\range{0}{\btL}]]$ and
+$\trailsave' = \trail[[\range{\btL{+}1}{\deepestLvl{-}1}]] +
+\trailsave$.
+
+When adding to the front of $\trailsave$ in this manner $\trailsave$
+can grow indefinitely. So we prune $\trailsave$ when it gets too large
+by (a) removing $\trailsave[i]$ if $\trailsave[i] = \trailsave[j]$ for
+some $j<i$ ($\trailsave[i]$ is redundant), and (b) removing the suffix
+of $\trailsave$ starting at $\trailsave[i]$ when
+$\trailsave[j] = \lnot \trailsave[i]$ for some $j<i$ ($\trailsave[i]$
+will never be useful as its negation, $\trailsave[j]$, would have to be
+added to $\trail$ first). In this way $\trailsave$ need never become
+larger than the number of variables in $\formula$.
+
+\paragraph{Lookahead for Conflicts.}
+In \ust we stop adding literals from $\trailsave$ to $\trail$ once we
+reach a decision literal $d$ on $\trailsave$ that is not yet on
+$\trail$ (line~\ref{ln:decnottrue} of \ust). This is done so that the
+solver has full control over variable decisions without interference
+from the trail saving mechanism (unlike the case with \cbt). However,
+another option would be to force the solver to use $d$ as its next
+decision literal, which would then allow us to further add all of
+$d$'s implied literals on $\trailsave$ onto $\trail$. This can be done
+for the first $k$ decisions on $\trailsave$ for any $k$. But in
+general, we do not want to remove the solver's autonomy by forcing it
+to made potentially different decisions than it might have wanted to.
+
+However, if there is a literal $l\in\trailsave$ for which
+$\lnot l\in \trail$, we can observe that forcing the solver to make
+all of the decisions of $\trailsave$ that lie above $l$ will
+immediately generate a conflict in the solver: $\reasonsave{l}$ will
+be falsified. In fact, in this situation we would not even need to
+perform unit propagation over the literals added from $\trailsave$;
+the literals and their reasons obtained from $\trailsave$ would be
+sufficient to perform 1-UIP learning from $\reasonsave{l}$.
+
+We experimented with this ``lookahead for conflicts'' idea using
+various values of $k$. We found that $k=2$, i.e., forcing up to two
+decisions from $\trailsave$ to be made by the solver if this yields a
+conflict, often enhanced the solver's performance. Limiting the lookahead to
+only one decision level of $\trailsave$ was not as good, and looking
+ahead more than 2 decisions of $\trailsave$ also degraded
+performance. This is provides some evidence that taking too much
+control away from the solver and forcing it to make too many decisions
+from $\trailsave$ can lead to conflicts that are not as useful to the
+solver.
+
+\paragraph{Reason quality.}
+The saved trail can be thought of as remembering the solver's recent
+trajectory. Sometimes we want to follow the past trajectory, but
+perhaps sometimes we don't. In particular, when adding literals from
+$\trailsave{}$ to $\trail$ we can examine the quality of the saved
+reasons to see if they are worth using. Once we encounter a literal
+with a low quality saved reason we stop adding literals from
+$\trailsave$ to the solver's trail. In particular, we can change
+lines~\ref{ln:addToTrail} and \ref{ln:addToTrailNext} of \ust
+to the following:\\
+\begin{tabular}[t]{l}
+23.5 \qquad \textbf{if} (lowQuality($\reasonsave{\litsave}$)) \textbf{break}\\
+24.  \qquad $\trail.\mathrm{addToEnd(\litsave)}$\\
+25.  \qquad $\reason{\litsave}\gets \reasonsave{\litsave}$
+\end{tabular}
+
+Note that the solver will still set the un-added literals as they are
+unit implied by $\trail$, but it might be able to find better reasons
+for these implicants. There is of course no guarantee that better
+reasons will be found, but our empirical results show that sometimes
+this does happen. We experimented with two quality metrics, clause
+size and clause LBD, obtaining positive results with both. These
+results also provides evidence against the argument given in
+\cite{DBLP:journals/jsat/TakRH11} that changing literal reasons is not
+impactful. With an appropriate clause quality metric the changing literal
+reasons can have an impact.
+
+% \section{Related Work}
+% \subsection{Trail Saving for Assumption-Based SAT}
+% Trail saving was briefly introduced in the context of assumption-based
+% SAT \cite{DBLP:conf/sat/HickeyB19} in order to speed up the
+% propagation of the assumptions each time the solver backtracked past
+% the assumption levels. Note that the process is the same, but it only
+% works to restore the propagants and implicants of the assumptions and
+% only after a learnt unit clause. Becuase the number of propagants and
+% implicants of the assumptions tends to become dominated by the number
+% of assumptions in the applications it was tested on, the technique did
+% not yield a significant improvement. It would be interesting to
+% combine the extended version of trail saving outlined in this paper
+% with trail saving over the assumptions to see if that could produce
+% better results.
+
+% \subsection{Comparison to Chronological Backtracking}
+% As mentioned earlier, chronological backtracking also achieves some of
+% the theoretical speedups that trail saving achieves. One advantage of
+% chronological backtracking is that it does not require any caching of
+% the trail or accessing such a cache during propagation; it simply
+% leaves a vast portion of the trail intact rather than backtracking all
+% the way to the conflict level. Chronological backtracking will also
+% speed up propagations slightly more than trail saving, because it
+% leaves more literals assigned at once which leads to clauses being
+% detected as unit more quickly.
+
+% One disadvantage that chronological backtracking has compared to trail
+% saving is that it doesn't allow the search as much flexibility to jump
+% out of local neighbourhoods, as the common variable selection
+% heuristics tend to depend on
+% \cite{DBLP:conf/dac/MoskewiczMZZM01,DBLP:conf/sat/2015,DBLP:conf/sat/LiangGPC16}. Because
+% chronological backtracking leaves a large portion of the trail
+% assigned, it will behave similarly to how the solver would behave if
+% it was to force the decisions that do occur within this large portion
+% of literals, whereas trail saving does not force any decisions. As
+% variable selection heuristics become more refined, it is likely that
+% adhering to them will become more crucial to performance.
+
+% Another disadvantage is that, because propagation is not performed at
+% every decision level, literals that remain assigned during
+% chronological backtracking (but otherwise would have been unassigned)
+% do not have a chance to move their decision levels up. An example of
+% this is when the solver was supposed to backtrack over a literal that
+% would end up being unit propagated at level 0, but stays assigned at
+% its current decision level when chronological backtracking is
+% used. This could potentially effect the lbd score or length of the
+% learnt clause that gets generated at the next conflict. Additionally,
+% chronological backtracking breaks the invarant that all literals
+% appear on the trail in monotonically increasing order by decision
+% level, which makes its implementation slightly more complicated.
 
@@
 watcher is searched for to replace $\lnot x_2$, n literals are
-saerched over without finding any, thus c is detected as unit and $l$
+searched over without finding any, thus c is detected as unit and $l$
 becomes forced to be true. This propagation took $O(n)$ searches

### 2020-02-25 figures/scatter.tex
experiments

--- 2020-02-16 figures/test2.tex
@@
 \begin{tikzpicture}
-\begin{axis}[title=Trail Saving Running Time Per Instance,axis equal,ymin=0,xmin=0,
-width=11cm,height=11cm,xlabel=original time (s),ylabel=trail saving time (s),axis x line=bottom,axis y line=left,xlabel near ticks,ylabel near ticks]
+\begin{axis}[title=Propagations Per Second in Cadical,axis equal,ymin=0,xmin=0,
+width=11cm,height=11cm,xlabel=Baseline props/sec (in M),ylabel=Trail Saving props/sec (in M),axis x line=bottom,axis y line=left,xlabel near ticks,ylabel near ticks]
 \addplot[only marks,mark size=1pt,mark=*,mark options={fill=black}]
 	table[x=original,y=trail]
-{test2.txt};
+{scatter.txt};
 \addplot [red,samples at={0,1}] {x};

### 2020-02-25 paper.tex
experiments

@@
 must be limited for peak effectiveness. This indicates that it is
-sometimes beneficial for the solver backtrack fully and redo its
+sometimes beneficial for the solver to backtrack fully and redo its
 trail, even if this takes more work. We will expand on why this might
-be case below.
-
-In this paper we present an alternative ``trail savings'' method where
-by we save the backtracked part of the solver's trail and attempt to
+be the case below.
+
+In this paper we present an alternative ``trail savings'' method whereby
+we save the backtracked part of the solver's trail and attempt to
 use that information to make the solver's redescent more
@@
 chronological backtracking. We also show that with the enhancements we
-are able to improve the performance of state of the art solvers.
+are able to improve the performance of state-of-the-art solvers.
 
@@
 this work we will need some additional flexibility in our definitions,
-as we will be sometimes be working with trails that would never be
+as we will sometimes be working with trails that would never be
 constructed by a \sat solver. Hence, we define a \emph{trail} to be a
@@
 decisions levels but the deepest. The full trail might, however,
-contain a conflict at its deepest level so is not necessarily conflict
+contain a conflict at its deepest level so is not necessarily conflict-
 free. The full trail might also not be propagation-complete, as unit
@@
 \section{Experiments and Results}
-For the version of cadical published in
-\cite{DBLP:conf/sat/MohleB19}chronological backtracking gives solves 6
-more problems and has a lower Par-2 score by a decent margin. When
-adding trail saving and all enhancements to the default
-non-chronological version of cadical, we end up solving 4 more problems
-and the Par-2 score goes decreases by a margin not as good as that
-produced by the non-chronological backtracking version. The trail saving
-versions improve on the standard non-chronological backtracking but do
-slightly worse than the chronological backtracking versions.
-
-For the latest version of cadical that we downloaded as of January 1,
-2020, chronological backtracking (set to default parameters)
-interestingly makes the solver perform worse than standard
-non-chronological backtracking does. This shows that not all solvers
-will benefit from chronological backtracking; it depends on the solver's
-mix of heuristics and may require fine-tuning the parameters associated
-with chronological backtracking. However, after implementing trail
-saving and enhancements on top of this version of cadical with
-non-chronological backtracking, the solver performs roughly the same as
-it does with standard non-chronological backtracking, and does not
-suffer a slow down as chronological backtracking appears to.
-
-For MapleSat, chronological backtracking makes a significant improvement
-on MapleLCMDist both in terms of total number of problems solved and
-Par-2 score. Similarly, trail saving implemented on top of MapleLCMDist
-with non-chronological backtracking receives a similarly significant
-improvement in terms of total instances solved and Par-2 score, and
-actually outperforms chronological backtracking on both measures.
+We implemented our techniques in two different SAT solvers, MapleSAT and Cadical,
+both of which have finished at or near the top of SAT competitions for the past several
+years \cite{Heule2019ProceedingsOS}. We then ran each solver on the 800 benchmark instances used
+in the main tracks of the 2018 and 2019 SAT competitions. The
+experiments were executed on a cluster of 2.7 GHz Intel cores with 5000 seconds CPU time and 7 GB memory limits for each instance.
+We chose not to output or verify the proofs generated by any of the solvers.
+The Par-2 scores obtained and total instances solved by each solver are reported in Figures 3, 4, and 5.\newline
+
+In Figure 3, we use the newest version of cadical (downloaded as of January 1, 2020) as the baseline solver.
+In our table, cadical refers to the baseline solver with non-chronological backtracking, cadical-chrono refers to the solver with \cbt enabled (using default settings), cadical-trail is the baseline with plain trail saving added (as described in Figure~\ref{fig:trailUse}), cadical-trail-append is the baseline with trail saving plus the first enhancement of saving over multiple backtracks, cadical-trail-append-lookahead adds the enhancement of lookahead for conflicts by 2 decision levels, and lastly cadical-trail-append-lookahead-reason also adds the final enhancement to cease trail saving once a reason of size greater than 10 literals is reached.
+For more details on the enhancments, please see section 4.2.
+
+Interestingly, \cbt makes the solver perform worse than the baseline.
+However, after implementing trail saving and all enhancements, the solver outperforms the baseline by a decent margin.
+The key enhancement for this solver seems to be the last one where we stop using the saved trail once we detect a reason of "poor" quality.
+We tried both clause size and lbd to act as a measure of clause quality, and both yielded a positive gain, with clause size being slightly more effective.
+\newline
+
+In Figure 4, we use the version of cadical published in
+\cite{DBLP:conf/sat/MohleB19} as a baseline, since we know it is a version of cadical that \cbt works particularly well on. When
+adding trail saving and all enhancements to the baseline, it solves the same number of instances as
+\cbt.
+Both \cbt and trail saving boost this solver's performance significantly.
+MapleLCMDist is another solver we know \cbt works well on, and in Figure 5 we use it as the baseline solver.
+\cbt and trail saving both make a significant improvement on this solver, but trail saving performs slightly better than \cbt in terms of total instances solved and Par-2 score.\newline
+
+In Figure 6 we plot the propagations per second for each instance with and without trail saving.
+With trail saving about 1.87 million literals per second are propagated on average, and without trail saving about 1.80 million literals per second are propagated on average.
+The graph further illustrates that the majority of the time trail saving will propagate literals faster than a solver without trail saving, as we argued earlier in the paper.
 
 \begin{figure}
+\centering
     \begin{tabular}{|l|l|l|l|l|l|}
       \hline
-      & Total Solved & SAT & UNSAT & Par-2 (x 10\textasciicircum{}6 s) \\ \hline
-      cadical                  & 532          & 314 &  218  & 3.220                             \\ \hline
-      cadical\_chrono          & 525          & 308 &  217  & 3.269                             \\ \hline
-      cadical\_trail           & 529          & 310 &  219  & 3.248                             \\ \hline
-      cadical\_trail\_enhanced & 531          & 313 &  218  & 3.216                             \\ \hline
+      & Total & SAT & UNSAT & Avg. Par-2 \\ \hline
+      cadical                  & 532          & 314 &  218  & 4025                             \\ \hline
+      cadical-chrono          & 525          & 308 &  217  & 4086                             \\ \hline
+      cadical-trail           & 529          & 310 &  \textbf{219}  & 4060                   \\ \hline
+      cadical-trail-append & 530 & 313 & 217 & 3930 \\ \hline
+      cadical-trail-append-lookahead & 531          & 313 &  218  & 4020                      \\ \hline
+      cadical-trail-append-lookahead-reason & \textbf{538} & \textbf{319} & \textbf{219} & \textbf{3854} \\ \hline
     \end{tabular}
@@
 \begin{figure}
+\centering
     \begin{tabular}{|l|l|l|l|l|l|}
       \hline
-      & Total Solved & SAT & UNSAT & Par-2 (x 10\textasciicircum{}6 s) \\ \hline
-      cadical                  & 492          & 289 &  203  & 3.5020                            \\ \hline
-      cadical\_chrono          & 498          & 292 &  206  & 3.4400                            \\ \hline
-      cadical\_trail           & 493          & 290 &  203  & 3.5228                            \\ \hline
-      cadical\_trail\_enhanced & 496          & 292 &  204  & 3.4847                            \\ \hline
+      & Total & SAT & UNSAT & Avg. Par-2 \\ \hline
+      cadical                  & 492          & 289 &  203  & 4378                            \\ \hline
+      cadical-chrono   & \textbf{498}  & \textbf{292} &  \textbf{206}  & \textbf{4300}      \\ \hline
+      cadical-trail           & 493          & 290 &  203  & 4403                            \\ \hline
+      cadical-trail-append-lookahead & 496  & \textbf{292} &  204  & 4356                    \\ \hline
+      cadical-trail-append-lookahead-reason & \textbf{498} & \textbf{292} & \textbf{206} & 4351 \\ \hline
     \end{tabular}
@@
 \begin{figure}
+\centering
     \begin{tabular}{|l|l|l|l|l|l|}
       \hline
-      & Total Solved & SAT & UNSAT & Par-2 (x 10\textasciicircum{}6 s) \\ \hline
-      maple           & 458          & 259 &  199  & 3.797                             \\ \hline
-      maple\_chrono   & 470          & 271 &  199  & 3.690                             \\ \hline
-      maple\_trail    & 472          & 271 &  201  & 3.694                             \\ \hline
-      maple\_enhanced & 473          & 277 &  195  & 3.667                             \\ \hline
+      & Total & SAT & UNSAT & Avg. Par-2 \\ \hline
+      maple           & 458          & 259 &  199  & 4746                             \\ \hline
+      maple-chrono   & 470          & 271 &  199  & 4613                             \\ \hline
+      maple-trail    & 472          & 271 &  \textbf{201}  & 4618                \\ \hline
+      maple-trail-append-lookahead & \textbf{473} & \textbf{277} &  195  & \textbf{4584} \\ \hline
     \end{tabular}
@@
 \end{figure}
-\clearpage
+
+\begin{figure} \centering \includegraphics[scale=0.7]{figures/scatter.pdf}\caption{Comparing propagations per second with and without trail saving in Cadical.}\end{figure}
 

### 2020-02-25 paper.tex
experiments

@@
 In Figure 4, we use the version of cadical published in
-\cite{DBLP:conf/sat/MohleB19} as a baseline, since we know it is a version of cadical that \cbt works particularly well on. When
-adding trail saving and all enhancements to the baseline, it solves the same number of instances as
+\cite{DBLP:conf/sat/MohleB19} as a baseline, since we know it is a version of cadical that \cbt works particularly well on.
+When adding trail saving and all enhancements to the baseline, it solves the same number of instances as
 \cbt.
 Both \cbt and trail saving boost this solver's performance significantly.
-MapleLCMDist is another solver we know \cbt works well on, and in Figure 5 we use it as the baseline solver.
-\cbt and trail saving both make a significant improvement on this solver, but trail saving performs slightly better than \cbt in terms of total instances solved and Par-2 score.\newline
-
-In Figure 6 we plot the propagations per second for each instance with and without trail saving.
-With trail saving about 1.87 million literals per second are propagated on average, and without trail saving about 1.80 million literals per second are propagated on average.
-The graph further illustrates that the majority of the time trail saving will propagate literals faster than a solver without trail saving, as we argued earlier in the paper.
+MapleLCMDist is another solver we know \cbt works well on, and in Figure 5 we show the result of implementing trail saving into that solver.
+\cbt and trail saving both make a significant improvement to MapleLCMDist, but trail saving performs slightly better than \cbt in terms of total instances solved and Par-2 score.\newline
+
+In Figure 6 we plot the propagations per second for each instance with and without trail saving enabled in cadical.
+With trail saving enabled about 1.87 million propagations per second occur, and without trail saving about 1.80 million propagations per second occur.
+The graph further illustrates that the majority of the time cadical with trail saving enabled will propagate literals faster than cadical without trail saving enabled, as we predicted would be the case earlier in the paper.
 

### 2020-02-27 paper.tex
conclusion

@@
 both of which have finished at or near the top of SAT competitions for the past several
-years \cite{Heule2019ProceedingsOS}. We then ran each solver on the 800 benchmark instances used
-in the main tracks of the 2018 and 2019 SAT competitions. The
+years \cite{Heule2018ProceedingsOS,Heule2019ProceedingsOS}. We then ran each solver on the 800 total benchmark instances used in the main tracks of the 2018 and 2019 SAT competitions. The
 experiments were executed on a cluster of 2.7 GHz Intel cores with 5000 seconds CPU time and 7 GB memory limits for each instance.
 We chose not to output or verify the proofs generated by any of the solvers.
-The Par-2 scores obtained and total instances solved by each solver are reported in Figures 3, 4, and 5.\newline
-
-In Figure 3, we use the newest version of cadical (downloaded as of January 1, 2020) as the baseline solver.
-In our table, cadical refers to the baseline solver with non-chronological backtracking, cadical-chrono refers to the solver with \cbt enabled (using default settings), cadical-trail is the baseline with plain trail saving added (as described in Figure~\ref{fig:trailUse}), cadical-trail-append is the baseline with trail saving plus the first enhancement of saving over multiple backtracks, cadical-trail-append-lookahead adds the enhancement of lookahead for conflicts by 2 decision levels, and lastly cadical-trail-append-lookahead-reason also adds the final enhancement to cease trail saving once a reason of size greater than 10 literals is reached.
+The Par-2 scores obtained and total instances solved by each solver are reported in Figures ~\ref{fig:cadicalNew}, ~\ref{fig:cadicalOld}, and ~\ref{fig:maple}.
+
+In Figure~\ref{fig:cadicalNew} we used the newest version of cadical (downloaded as of January 1, 2020) as the baseline solver, in Figure~\ref{fig:cadicalOld} we used the version of cadical published in
+\cite{DBLP:conf/sat/MohleB19} as the baseline, and in Figure~\ref{fig:maple} we used MapleLCMDist \cite{xiao2017maplelrb} as the baseline.
+Each of the baselines were run with standard non-chronological backtracking.
+We then refer to versions of each solver with additional features implemented on top by adding suffixes.
+"-chrono" refers to the solver with \cbt enabled (using the solver's default settings), "-trail" refers to the baseline with plain trail saving added (as described in Figure~\ref{fig:trailUse}), "-trail-multipleBT" refers to the baseline with trail saving plus the first enhancement of saving over multiple backtracks, "-trail-multipleBT-lookahead" also adds the enhancement of lookahead for conflicts by 2 decision levels, and "-trail-multipleBT-lookahead-reason" also adds the final enhancement to cease trail saving once a reason of "low quality" is reached.
 For more details on the enhancments, please see section 4.2.
 
-Interestingly, \cbt makes the solver perform worse than the baseline.
-However, after implementing trail saving and all enhancements, the solver outperforms the baseline by a decent margin.
-The key enhancement for this solver seems to be the last one where we stop using the saved trail once we detect a reason of "poor" quality.
-We tried both clause size and lbd to act as a measure of clause quality, and both yielded a positive gain, with clause size being slightly more effective.
-\newline
-
-In Figure 4, we use the version of cadical published in
-\cite{DBLP:conf/sat/MohleB19} as a baseline, since we know it is a version of cadical that \cbt works particularly well on.
-When adding trail saving and all enhancements to the baseline, it solves the same number of instances as
-\cbt.
-Both \cbt and trail saving boost this solver's performance significantly.
-MapleLCMDist is another solver we know \cbt works well on, and in Figure 5 we show the result of implementing trail saving into that solver.
-\cbt and trail saving both make a significant improvement to MapleLCMDist, but trail saving performs slightly better than \cbt in terms of total instances solved and Par-2 score.\newline
-
-In Figure 6 we plot the propagations per second for each instance with and without trail saving enabled in cadical.
+Interestingly, \cbt made the newest version of cadical perform worse than the baseline (Figure~\ref{fig:cadicalNew}).
+This demonstrates that \cbt is not always beneficial.
+Trail saving alone did not impact the performance of this solver significantly, but adding all of the enhancements on top of trail saving resulted in solving six more instances and yielding a better Par-2 score than the baseline.
+The key enhancement for this solver seemed to be the last one where we stop using the saved trail once we detect a reason of "low quality".
+We tried both clause size and lbd as the clause quality metric, and both yielded a positive gain, with clause size being slightly more effective.
+
+The version of cadical used in Figure~\ref{fig:cadicalOld} benefited from \cbt, which agrees with previously published results \cite{DBLP:conf/sat/MohleB19}.
+Trail saving alone did not significantly impact this solver, but adding all of the enhancements on top of trail saving resulted in solving the same number of instances as the solver with \cbt did.
+
+MapleLCMDist (in Figure~\ref{fig:maple}) is another solver that benefited from \cbt, and trail saving alone solved more instances than the solver with \cbt did.
+Adding the first two enhancements on top of trail saving resulted in solving three more instances and yielding a better Par-2 score than the solver with \cbt did.
+Adding the last enhancement of ceasing trail saving on a "low quality" reason made the performance worse, whether clause size or lbd was used as the clause quality metric.
+This suggests that the enhancements to trail saving have different impacts on different solvers.
+
+In Figure~\ref{fig:props} we plotted the propagations per second for each instance with and without trail saving enabled in cadical.
 With trail saving enabled about 1.87 million propagations per second occur, and without trail saving about 1.80 million propagations per second occur.
-The graph further illustrates that the majority of the time cadical with trail saving enabled will propagate literals faster than cadical without trail saving enabled, as we predicted would be the case earlier in the paper.
+The graph further illustrates that the majority of the time cadical with trail saving enabled will propagate literals faster than cadical without trail saving enabled, as we predicted would be the case earlier in this paper.
 
@@
       cadical-trail           & 529          & 310 &  \textbf{219}  & 4060                   \\ \hline
-      cadical-trail-append & 530 & 313 & 217 & 3930 \\ \hline
-      cadical-trail-append-lookahead & 531          & 313 &  218  & 4020                      \\ \hline
-      cadical-trail-append-lookahead-reason & \textbf{538} & \textbf{319} & \textbf{219} & \textbf{3854} \\ \hline
+      cadical-trail-multipleBT & 530 & 313 & 217 & 3930 \\ \hline
+      cadical-trail-multipleBT-lookahead & 531          & 313 &  218  & 4020                      \\ \hline
+      cadical-trail-multipleBT-lookahead-reason & \textbf{538} & \textbf{319} & \textbf{219} & \textbf{3854} \\ \hline
     \end{tabular}
     \caption{Table of results for cadical, version pulled from github as of January 1, 2020.}
+    \label{fig:cadicalNew}
 \end{figure}
@@
       cadical-trail           & 493          & 290 &  203  & 4403                            \\ \hline
-      cadical-trail-append-lookahead & 496  & \textbf{292} &  204  & 4356                    \\ \hline
-      cadical-trail-append-lookahead-reason & \textbf{498} & \textbf{292} & \textbf{206} & 4351 \\ \hline
+      cadical-trail-multipleBT-lookahead & 496  & \textbf{292} &  204  & 4356                    \\ \hline
+      cadical-trail-multipleBT-lookahead-reason & \textbf{498} & \textbf{292} & \textbf{206} & 4351 \\ \hline
     \end{tabular}
     \caption{Table of results for cadical or "chrono", version published in \cite{DBLP:conf/sat/MohleB19}.}
+    \label{fig:cadicalOld}
 \end{figure}
@@
       maple-trail    & 472          & 271 &  \textbf{201}  & 4618                \\ \hline
-      maple-trail-append-lookahead & \textbf{473} & \textbf{277} &  195  & \textbf{4584} \\ \hline
+      maple-trail-multipleBT-lookahead & \textbf{473} & \textbf{277} &  195  & \textbf{4584} \\ \hline
+      maple-trail-multipleBT-lookahead-reason & 999 & 999 & 999 & 9999 \\ \hline
     \end{tabular}
     \caption{Table of results for MapleLCMDist.}
+    \label{fig:maple}
 \end{figure}
 
-\begin{figure} \centering \includegraphics[scale=0.7]{figures/scatter.pdf}\caption{Comparing propagations per second with and without trail saving in Cadical.}\end{figure}
-
-\section{Future Work}
-
-With more clever data structures the efficiency of restoring the trail
-could be improved. There are a number of parameters which can be
-tuned, including how often to refresh the trail, how large the cutoff
-point for rejecting "long" reasons should be, how long to wait to
-prune the trail cache if one is appending multiple trail caches
-together, how far to look ahead on the trail cache for conflicts (in
-terms of number of decisions or total number of literals). We only
-experimented with a few different settings of these parameters that
-made sense, but more careful fine-tuning of these parameters could
-further enhance the effectiveness of the technique. There are
-potentially other techniques one could use a trail cache for,
-including inprocessing steps that involve probing. A look down the
-trail cache is an incomplete but very fast way to do a probing step,
-as opposed to propagation. It is also possible that trail saving and
-chronological backtracking could be combined in some way to produce
-even better results.
+\begin{figure} \centering \includegraphics[scale=0.7]{figures/scatter.pdf}\caption{Comparing propagations per second with and without trail saving in Cadical.}\label{fig:props}\end{figure}
+
+\section{Conclusion}
+We have shown that our method "trail savings" can speed up two
+state-of-the-art \sat solvers, cadical and MapleSAT, about as well as
+chronological backtracking can without breaking the conventional invariants
+of CDCL \sat solvers. We also introduced three enhancements one can
+implement if they are using a saved trail and demonstrated experimentally
+that these enhancements can sometimes improve the performance by a significant
+amount. Future work could include tuning the parameters of the enhancements 
+to work optimally or introducing further enhancements, such as using the saved
+trail to help make inprocessing techniques faster or
+using the saved trail to learn a second learnt clause from a single conflict.
+Lastly, it might be possible to combine trail saving with chronological
+backtracking and figure out when each is most beneficial.
+
+%\section{Future Work}
+
+%With more clever data structures the efficiency of restoring the trail
+%could be improved. There are a number of parameters which can be
+%tuned, including how often to refresh the trail, how large the cutoff
+%point for rejecting "long" reasons should be, how long to wait to
+%prune the trail cache if one is appending multiple trail caches
+%together, how far to look ahead on the trail cache for conflicts (in
+%terms of number of decisions or total number of literals). We only
+%experimented with a few different settings of these parameters that
+%made sense, but more careful fine-tuning of these parameters could
+%further enhance the effectiveness of the technique. There are
+%potentially other techniques one could use a trail cache for,
+%including inprocessing steps that involve probing. A look down the
+%trail cache is an incomplete but very fast way to do a probing step,
+%as opposed to propagation. It is also possible that trail saving and
+%chronological backtracking could be combined in some way to produce
+%even better results.
 

### 2020-02-27 paper.tex
typos

@@
 
-In this paper we present an alternative ``trail savings'' method whereby
+In this paper we present an alternative ``trail saving'' method whereby
 we save the backtracked part of the solver's trail and attempt to
@@
 Each literal $\ell\in\trail$ has a decision level $\dlevel{\ell}$
-which is equal to the number of decision literals appearing before
-$\ell$ on the trail plus one; hence, $\dlevel{d}=1$ for the first
+which is equal to the number of decision literals appearing on the trail
+up to and including $\ell$; hence, $\dlevel{d}=1$ for the first
 decision literal $d\in\trail$. The set of literals on $\trail$ that
@@
     $\trail$ implying $l$, and for each $x\in C$ with $x\neq l$ we
-    that $\lnot x$ appears before $l$ on $\trail$:
+    have that $\lnot x$ appears before $l$ on $\trail$:
     $\forall l\in\trail.\, \reason{l}\neq \varnothing \implies
@@
     literal $y\neq l$ with $\dlevel{y} = \dlevel{l}$.
-\item[conflict free:] No clause of $F$ is falsified by
+\item[conflict-free:] No clause of $F$ is falsified by
     $\trail$. Clauses $C\in F$ falsified by $\trail$ are typically
@@
 \begin{example}
-    For example, say that $(x, \lnot y)\in \formula$, the literal $x$
+    For example, suppose that $(x, \lnot y)\in \formula$, the literal $x$
     is a decision literal on the trail with $\dlevel{x}=2$, and that
@@
 to the \sat solver, they exist. Hence, perhaps the most unappealing
-aspect of \cbt is that introduces additional inter-dependencies
+aspect of \cbt is that it introduces additional interdependencies
 between components of the solver. For example, the second impact means
@@
 
-Say that the solver is at $\deepestLvl$ where it has encountered a
+Suppose that the solver is at $\deepestLvl$ where it has encountered a
 conflict. From the 1-UIP clause it learns, $\uipcls$, it now has to
@@
 argument $\trailIdx{\deepestLit}$ (i.e., the trail index of the newly
-added implicant). The saved trail be continually consulted during the
+added implicant). The saved trail will be continually consulted during the
 solver's descent whenever unit propagation is performed. When
@@
     \caption{Use of $\trailsave$ from
-      Example~\ref{ex:trailsave}\label{fig:trailsave}. The literals
+      Example~\ref{ex:trailsave}\label{fig:trailsave}. The literal's
       decision level is indicated in its superscript, and a $\mbox{}^*$
@@
     example of how $\trailsave$ is used. Initially the literals $l_1$
-    to $l_{14}$ are on the solvers $\trail$, and $\trailsave$ is
+    to $l_{14}$ are on the solver's $\trail$, and $\trailsave$ is
     empty. This is shown in the first two lines of the figure. In the
@@
     will backtrack to level $2$, where it will add $\lnot l_{12}$ as a
-    unit implicant. The next two lines shows $\trail$ and $\trailsave$
+    unit implicant. The next two lines show $\trail$ and $\trailsave$
     right after the backtrack to level $2$: the backtracked levels
@@
 reproducing the trail on the solver's redescent. Nevertheless, trail
-savings improves the efficiency of this redescent in three different
+saving improves the efficiency of this redescent in three different
 ways. First, by adding more forced literals to the trail before
@@
 have to traverse each of these reason clauses to determine that they
-are have in fact become unit. Third, when a conflict is returned by
+have in fact become unit. Third, when a conflict is returned by
 \ust all further propagations can be halted. The added literals and
@@
     first $i$ literals on $\trailsave$ are all in $\trail$
-    ($\forall j. 0\leq j< i. \trailsave[i]\in \trail$) and
+    ($\forall j. 0\leq j< i. \trailsave[j]\in \trail$) and
     $\trailsave[i]=l$ is an implied literal with $\reasonsave{l} = C$,
@@
 or $\lnot x \in \trailsave[0]\ldots\trailsave[i-1]$. But in the later
-case we also $\lnot x \in \trail$. 
+case we also have $\lnot x \in \trail$. 
 \whitebox
@@
 the solver (4). In procedure \bt $\trail+\trailsave$ is
-set to a trail that was previously generated by solver (4). The solver
-can add to $\trail$ by decisions and propagations with out using
+set to a trail that was previously generated by the solver (4). The solver
+can add to $\trail$ by decisions and propagations without using
 $\trailsave$. In this case $\trail' = \trail+\trail_{\mathit{new}}$,
@@
 backtrack. Our first extension addresses this potential issue and also
-provides more general trail savings in other contexts as well.
+provides more general trail saving in other contexts as well.
 
@@
 general, we do not want to remove the solver's autonomy by forcing it
-to made potentially different decisions than it might have wanted to.
+to make potentially different decisions than it might have wanted to.
 
@@
 ahead more than 2 decisions of $\trailsave$ also degraded
-performance. This is provides some evidence that taking too much
+performance. This provides some evidence that taking too much
 control away from the solver and forcing it to make too many decisions
@@
 trajectory. Sometimes we want to follow the past trajectory, but
-perhaps sometimes we don't. In particular, when adding literals from
+perhaps sometimes we do not. In particular, when adding literals from
 $\trailsave{}$ to $\trail$ we can examine the quality of the saved
@@
 \cite{DBLP:journals/jsat/TakRH11} that changing literal reasons is not
-impactful. With an appropriate clause quality metric the changing literal
+impactful. With an appropriate clause quality metric the changing of literal
 reasons can have an impact.
@@
       maple-trail    & 472          & 271 &  \textbf{201}  & 4618                \\ \hline
-      maple-trail-multipleBT-lookahead & \textbf{473} & \textbf{277} &  195  & \textbf{4584} \\ \hline
-      maple-trail-multipleBT-lookahead-reason & 999 & 999 & 999 & 9999 \\ \hline
+      maple-trail-multipleBT-lookahead & \textbf{473} & \textbf{278} &  195  & \textbf{4584} \\ \hline
+      maple-trail-multipleBT-lookahead-reason & 467 & 273 & 194 & 4668 \\ \hline
     \end{tabular}
@@
 \section{Conclusion}
-We have shown that our method "trail savings" can speed up two
+We have shown that our method trail saving can speed up two
 state-of-the-art \sat solvers, cadical and MapleSAT, about as well as

### 2020-02-28 figures/cactus_cadical_new.tex
plot

\documentclass{standalone}
\title{}
\date{}
\author{}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{algorithm}
\usepackage[noend]{algorithmic}
\usepackage{tikz,pgfplotstable,pgfplots}
\usetikzlibrary{arrows}
\begin{document}

\begin{tikzpicture}
\begin{axis}[
	xlabel={Instances Solved},
	ylabel={Run Time (sec)},width=18cm,height=12cm,legend pos=north west,xmin=400,ymin=1000,xmax=540,ymax=5100
]

\addplot[red,mark=x] table [x=index, y=original_time, col sep=comma] {results_cadical_nochrono_data.csv};
\addplot[green,mark=o] table [x=index, y=original_time, col sep=comma] {results_cadical_chrono_data.csv};
\addplot[blue,mark=+] table [x=index, y=original_time, col sep=comma] {results_cadical_nochrono_trail_data.csv};
\addplot[cyan,mark=square] table [x=index, y=original_time, col sep=comma] {results_cadical_nochrono_trail_1lvl_data.csv};
\addplot[gray,mark=triangle] table [x=index, y=original_time, col sep=comma] {results_cadical_nochrono_trail_1lvl_10reason_data.csv};

\legend{cadical,cadical-chrono,cadical-trail,cadical-trail-multipleBT-lookahead, cadical-trail-multipleBT-lookahead-reason}
\end{axis}
\end{tikzpicture}

\end{document}

### 2020-02-28 paper.tex
plot

@@
     often yields superior performance over chronological backtracking,
-    and that is is able to improve the performance of state of the art
+    and that is is able to improve the performance of state-of-the-art
     solvers.
@@
 
-\begin{figure} \centering \includegraphics[scale=0.7]{figures/scatter.pdf}\caption{Comparing propagations per second with and without trail saving in Cadical.}\label{fig:props}\end{figure}
+\begin{figure}\centering\includegraphics[scale=0.65]{figures/cactus_cadical_new.pdf}\caption{\small{Cactus plot for the newest version of cadical comparing standard non-chronological backtracking to \cbt and various configurations of trail saving.
+The first 400 problems were solved in less than 1200 seconds, so that part of the plot is truncated.}}\label{fig:cactus_cadical}\end{figure}
+
+\begin{figure}\centering\includegraphics[scale=0.7]{figures/scatter.pdf}\caption{Comparing propagations per second with and without trail saving in Cadical.}\label{fig:props}\end{figure}
 
@@
 chronological backtracking can without breaking the conventional invariants
-of CDCL \sat solvers. We also introduced three enhancements one can
-implement if they are using a saved trail and demonstrated experimentally
-that these enhancements can sometimes improve the performance by a significant
-amount. Future work could include tuning the parameters of the enhancements 
-to work optimally or introducing further enhancements, such as using the saved
-trail to help make inprocessing techniques faster or
+of CDCL. We also introduced three enhancements one can
+implement when using a saved trail and demonstrated experimentally
+that these enhancements can sometimes improve a solver's performance by a significant
+amount. We have shown that trail saving and all enhancements we proposed are sound.
+A line of future work might include introducing further enhancements,
+such as using the saved trail to help make inprocessing techniques faster or
 using the saved trail to learn a second learnt clause from a single conflict.
-Lastly, it might be possible to combine trail saving with chronological
-backtracking and figure out when each is most beneficial.
+It is also possible to combine trail saving with chronological
+backtracking if one is willing to break the traditional invariants of CDCL,
+but we leave it to future work to determine whether or not this would be
+useful and how to best approach it.
 

### 2020-02-28 paper.tex
minor changes

@@
 $\trailsave$ to the solver's trail. In particular, we can change
-lines~\ref{ln:addToTrail} and \ref{ln:addToTrailNext} of \ust
-to the following:\\
+lines~\ref{ln:addToTrail}--\ref{ln:addToTrailNext} of \ust
+to the following:\\[5pt]
 \begin{tabular}[t]{l}
@@
 25.  \qquad $\reason{\litsave}\gets \reasonsave{\litsave}$
-\end{tabular}
+\end{tabular}\\[5pt]
 
@@
 \section{Experiments and Results}
-We implemented our techniques in two different SAT solvers, MapleSAT and Cadical,
-both of which have finished at or near the top of SAT competitions for the past several
-years \cite{Heule2018ProceedingsOS,Heule2019ProceedingsOS}. We then ran each solver on the 800 total benchmark instances used in the main tracks of the 2018 and 2019 SAT competitions. The
-experiments were executed on a cluster of 2.7 GHz Intel cores with 5000 seconds CPU time and 7 GB memory limits for each instance.
-We chose not to output or verify the proofs generated by any of the solvers.
-The Par-2 scores obtained and total instances solved by each solver are reported in Figures ~\ref{fig:cadicalNew}, ~\ref{fig:cadicalOld}, and ~\ref{fig:maple}.
-
-In Figure~\ref{fig:cadicalNew} we used the newest version of cadical (downloaded as of January 1, 2020) as the baseline solver, in Figure~\ref{fig:cadicalOld} we used the version of cadical published in
-\cite{DBLP:conf/sat/MohleB19} as the baseline, and in Figure~\ref{fig:maple} we used MapleLCMDist \cite{xiao2017maplelrb} as the baseline.
-Each of the baselines were run with standard non-chronological backtracking.
-We then refer to versions of each solver with additional features implemented on top by adding suffixes.
-"-chrono" refers to the solver with \cbt enabled (using the solver's default settings), "-trail" refers to the baseline with plain trail saving added (as described in Figure~\ref{fig:trailUse}), "-trail-multipleBT" refers to the baseline with trail saving plus the first enhancement of saving over multiple backtracks, "-trail-multipleBT-lookahead" also adds the enhancement of lookahead for conflicts by 2 decision levels, and "-trail-multipleBT-lookahead-reason" also adds the final enhancement to cease trail saving once a reason of "low quality" is reached.
-For more details on the enhancments, please see section 4.2.
-
-Interestingly, \cbt made the newest version of cadical perform worse than the baseline (Figure~\ref{fig:cadicalNew}).
-This demonstrates that \cbt is not always beneficial.
-Trail saving alone did not impact the performance of this solver significantly, but adding all of the enhancements on top of trail saving resulted in solving six more instances and yielding a better Par-2 score than the baseline.
-The key enhancement for this solver seemed to be the last one where we stop using the saved trail once we detect a reason of "low quality".
-We tried both clause size and lbd as the clause quality metric, and both yielded a positive gain, with clause size being slightly more effective.
-
-The version of cadical used in Figure~\ref{fig:cadicalOld} benefited from \cbt, which agrees with previously published results \cite{DBLP:conf/sat/MohleB19}.
-Trail saving alone did not significantly impact this solver, but adding all of the enhancements on top of trail saving resulted in solving the same number of instances as the solver with \cbt did.
-
-MapleLCMDist (in Figure~\ref{fig:maple}) is another solver that benefited from \cbt, and trail saving alone solved more instances than the solver with \cbt did.
-Adding the first two enhancements on top of trail saving resulted in solving three more instances and yielding a better Par-2 score than the solver with \cbt did.
-Adding the last enhancement of ceasing trail saving on a "low quality" reason made the performance worse, whether clause size or lbd was used as the clause quality metric.
-This suggests that the enhancements to trail saving have different impacts on different solvers.
-
-In Figure~\ref{fig:props} we plotted the propagations per second for each instance with and without trail saving enabled in cadical.
-With trail saving enabled about 1.87 million propagations per second occur, and without trail saving about 1.80 million propagations per second occur.
-The graph further illustrates that the majority of the time cadical with trail saving enabled will propagate literals faster than cadical without trail saving enabled, as we predicted would be the case earlier in this paper.
-
-\begin{figure}
+We implemented our techniques in two different SAT solvers, MapleSAT
+and Cadical, both of which have finished at or near the top of SAT
+competitions for the past several years
+\cite{Heule2018ProceedingsOS,Heule2019ProceedingsOS}. We then ran each
+solver on the 800 total benchmark instances used in the main tracks of
+the 2018 and 2019 SAT competitions. The experiments were executed on a
+cluster of 2.7 GHz Intel cores with 5000 seconds CPU time and 7 GB
+memory limits for each instance. We chose not to output or verify the
+proofs generated by any of the solvers. The Par-2 scores obtained and
+total instances solved by each solver are reported in
+Figures~\ref{fig:cadicalNew}, \ref{fig:cadicalOld}, and
+\ref{fig:maple}.
+
+In Figure~\ref{fig:cadicalNew} we used the newest version of cadical
+(downloaded as of January 1, 2020) as the baseline solver, in
+Figure~\ref{fig:cadicalOld} we used the version of cadical published
+in \cite{DBLP:conf/sat/MohleB19} as the baseline, and in
+Figure~\ref{fig:maple} we used MapleLCMDist \cite{xiao2017maplelrb} as
+the baseline. Each of the baselines were run with standard
+non-chronological backtracking. We then refer to versions of each
+solver with additional features implemented on top by adding
+suffixes. "-chrono" refers to the solver with \cbt enabled (using the
+solver's default settings), "-trail" refers to the baseline with plain
+trail saving added (as described in Figure~\ref{fig:trailUse}),
+"-trail-multipleBT" refers to the baseline with trail saving plus the
+first enhancement of saving over multiple backtracks,
+"-trail-multipleBT-lookahead" also adds the enhancement of lookahead
+for conflicts by 2 decision levels, and
+"-trail-multipleBT-lookahead-reason" also adds the final enhancement
+to cease trail saving once a reason of "low quality" is reached. For
+more details on the enhancments, please see section 4.2.
+
+Interestingly, \cbt made the newest version of cadical perform worse
+than the baseline (Figure~\ref{fig:cadicalNew}). This demonstrates
+that \cbt is not always beneficial. Trail saving alone did not impact
+the performance of this solver significantly, but adding all of the
+enhancements on top of trail saving resulted in solving six more
+instances and yielding a better Par-2 score than the baseline. The key
+enhancement for this solver seemed to be the last one where we stop
+using the saved trail once we detect a reason of "low quality". We
+tried both clause size and lbd as the clause quality metric, and both
+yielded a positive gain, with clause size being slightly more
+effective.
+
+\begin{figure}[t]
 \centering
@@
 
+The version of cadical used in Figure~\ref{fig:cadicalOld} did show
+benefits from \cbt in agreement with previously published results
+\cite{DBLP:conf/sat/MohleB19}. Trail saving alone did not
+significantly impact this solver, but adding all of the enhancements
+on top of trail saving resulted in solving the same number of
+instances as the solver with \cbt did, albeit with a slight increase
+in the Par-2 score.
+
 \begin{figure}
@@
 
+MapleLCMDist (in Figure~\ref{fig:maple}) is another solver that
+benefited from \cbt. In this solver trail saving alone solved more
+instances than the solver with \cbt did. Adding the first two
+enhancements on top of trail saving resulted in solving three more
+instances and yielding a better Par-2 score than the solver with
+\cbt. Adding the last enhancement of ceasing trail saving on a "low
+quality" reason made the performance worse, whether clause size or lbd
+was used as the clause quality metric. This suggests that the
+enhancements to trail saving have different impacts on different
+solvers.
+
 \begin{figure}
@@
 
+In Figure~\ref{fig:props} we plotted the propagations per second for
+each instance with and without trail saving enabled in cadical. With
+trail saving enabled about 1.87 million propagations per second occur,
+and without trail saving about 1.80 million propagations per second
+occur. The graph further illustrates that the majority of the time
+cadical with trail saving enabled will propagate literals faster than
+cadical without trail saving enabled, as we predicted would be the
+case earlier in this paper.
+
 \begin{figure}\centering\includegraphics[scale=0.65]{figures/cactus_cadical_new.pdf}\caption{\small{Cactus plot for the newest version of cadical comparing standard non-chronological backtracking to \cbt and various configurations of trail saving.
@@
 \section{Conclusion}
-We have shown that our method trail saving can speed up two
-state-of-the-art \sat solvers, cadical and MapleSAT, about as well as
-chronological backtracking can without breaking the conventional invariants
-of CDCL. We also introduced three enhancements one can
-implement when using a saved trail and demonstrated experimentally
-that these enhancements can sometimes improve a solver's performance by a significant
-amount. We have shown that trail saving and all enhancements we proposed are sound.
-A line of future work might include introducing further enhancements,
-such as using the saved trail to help make inprocessing techniques faster or
-using the saved trail to learn a second learnt clause from a single conflict.
-It is also possible to combine trail saving with chronological
-backtracking if one is willing to break the traditional invariants of CDCL,
-but we leave it to future work to determine whether or not this would be
-useful and how to best approach it.
+We have shown that our trail saving technique can speed up two
+state-of-the-art \sat solvers, cadical and MapleSAT, as or more
+effectively than chronological backtracking can without breaking the
+conventional invariants of CDCL. We also introduced three enhancements
+one can implement when using a saved trail and demonstrated
+experimentally that these enhancements can sometimes improve a
+solver's performance by a significant amount. We have shown that trail
+saving and all enhancements we proposed are sound.
+
+There are many avenues that can be pursued in future work, such as
+using the saved trail to help make inprocessing techniques faster or
+using the saved trail to learn a multiple clauses from a single
+conflict. It is also possible to combine trail saving with
+chronological backtracking if one is willing to break the traditional
+invariants of CDCL, but it would require further work to determine
+whether or not this would be useful and how to best approach it.
 

### 2020-02-29 figures/cactus_maple_new.tex
final

\documentclass{standalone}
\title{}
\date{}
\author{}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{algorithm}
\usepackage[noend]{algorithmic}
\usepackage{tikz,pgfplotstable,pgfplots}
\usetikzlibrary{arrows}
\begin{document}

\begin{tikzpicture}
\begin{axis}[
	xlabel={Instances Solved},
	ylabel={Run Time (sec)},width=18cm,height=12cm,legend pos=north west,xmin=300,ymin=600,xmax=480,ymax=5100
]

\addplot[red,mark=x] table [x=index, y=original_time, col sep=comma] {results_maplelcmdist_data.csv};
\addplot[green,mark=o] table [x=index, y=original_time, col sep=comma] {results_maplelcmdist_chrono_data.csv};
\addplot[blue,mark=+] table [x=index, y=original_time, col sep=comma] {results_maplelcmdist_trail_noforce_data.csv};
\addplot[cyan,mark=square] table [x=index, y=original_time, col sep=comma] {results_maplelcmdist_trail_noforce_1lvl_data.csv};
\addplot[gray,mark=triangle] table [x=index, y=original_time, col sep=comma] {results_maplelcmdist_trail_1lvl_5lbdreason_data.csv};

\legend{maple,maple-chrono,maple-trail,maple-trail-multipleBT-lookahead,maple-trail-multipleBT-lookahead-reason}
\end{axis}
\end{tikzpicture}

\end{document}

### 2020-02-29 paper.tex
final

@@
 Figures~\ref{fig:cadicalNew}, \ref{fig:cadicalOld}, and
-\ref{fig:maple}.
+\ref{fig:maple}. We also show the cactus plots of the new version of
+cadical in Figure \ref{fig:cactus_cadical} and MapleLCMDist in Figure
+\ref{fig:cactus_maple}.
 
@@
 In Figure~\ref{fig:props} we plotted the propagations per second for
-each instance with and without trail saving enabled in cadical. With
-trail saving enabled about 1.87 million propagations per second occur,
-and without trail saving about 1.80 million propagations per second
-occur. The graph further illustrates that the majority of the time
-cadical with trail saving enabled will propagate literals faster than
+each instance with and without trail saving enabled in cadical (rounded
+to the nearest 0.1 million). Only
+the 506 instances solved by both solvers are plotted. With trail saving
+enabled, 213 instances had more propagations per second, 149 instances
+had less propagations per second, and 144 instances had about the same
+number of propagations per second. On the graph, points above the diagonal
+represent instances where trail saving had faster propagation. The
+graph illustrates that the majority of the time
+cadical with trail saving enabled propagated literals faster than
 cadical without trail saving enabled, as we predicted would be the
@@
 The first 400 problems were solved in less than 1200 seconds, so that part of the plot is truncated.}}\label{fig:cactus_cadical}\end{figure}
+
+\begin{figure}\centering\includegraphics[scale=0.65]{figures/cactus_maple_new.pdf}\caption{\small{Cactus plot for MapleLCMDist comparing standard non-chronological backtracking to \cbt and various configurations of trail saving.
+The first 300 problems were solved in less than 600 seconds, so that part of the plot is truncated.}}\label{fig:cactus_maple}\end{figure}
 
@@
 using the saved trail to help make inprocessing techniques faster or
-using the saved trail to learn a multiple clauses from a single
+using the saved trail to learn multiple clauses from a single
 conflict. It is also possible to combine trail saving with

### 2020-03-02 paper.tex
used maple data more consistent with definitions

@@
 \usepackage{theorem}
+\usepackage{float}
 
@@
 to cease trail saving once a reason of "low quality" is reached. For
-more details on the enhancments, please see section 4.2.
+more details on the enhancements, please see section 4.2.
 
@@
 
-\begin{figure}[t]
+\begin{figure}[H]
 \centering
@@
 
+\begin{figure}\centering\includegraphics[scale=0.65]{figures/cactus_cadical_new.pdf}\caption{\small{Cactus plot for the newest version of cadical comparing standard non-chronological backtracking to \cbt and various configurations of trail saving.
+The first 400 problems were solved in less than 1200 seconds, so that part of the plot is truncated.}}\label{fig:cactus_cadical}\end{figure}
+
 The version of cadical used in Figure~\ref{fig:cadicalOld} did show
@@
 MapleLCMDist (in Figure~\ref{fig:maple}) is another solver that
-benefited from \cbt. In this solver trail saving alone solved more
+benefited from \cbt. In this solver trail saving alone solved two more
 instances than the solver with \cbt did. Adding the first two
-enhancements on top of trail saving resulted in solving three more
-instances and yielding a better Par-2 score than the solver with
+enhancements on top of trail saving resulted in solving only one more
+instance but yielded a better Par-2 score than the solver with
 \cbt. Adding the last enhancement of ceasing trail saving on a "low
@@
       maple-chrono   & 470          & 271 &  199  & 4613                             \\ \hline
-      maple-trail    & 472          & 271 &  \textbf{201}  & 4618                \\ \hline
-      maple-trail-multipleBT-lookahead & \textbf{473} & \textbf{278} &  195  & \textbf{4584} \\ \hline
-      maple-trail-multipleBT-lookahead-reason & 467 & 273 & 194 & 4668 \\ \hline
+      maple-trail    & \textbf{472}          & 271 &  \textbf{201}  & 4618                \\ \hline
+      maple-trail-multipleBT-lookahead & 471 & 272 &  199  & \textbf{4597}        \\ \hline
+      maple-trail-multipleBT-lookahead-reason & 467 & \textbf{273} & 194 & 4668 \\ \hline
     \end{tabular}
@@
 \end{figure}
+
+\begin{figure}\centering\includegraphics[scale=0.65]{figures/cactus_maple_new.pdf}\caption{\small{Cactus plot for MapleLCMDist comparing standard non-chronological backtracking to \cbt and various configurations of trail saving.
+The first 300 problems were solved in less than 600 seconds, so that part of the plot is truncated.}}\label{fig:cactus_maple}\end{figure}
 
@@
 case earlier in this paper.
-
-\begin{figure}\centering\includegraphics[scale=0.65]{figures/cactus_cadical_new.pdf}\caption{\small{Cactus plot for the newest version of cadical comparing standard non-chronological backtracking to \cbt and various configurations of trail saving.
-The first 400 problems were solved in less than 1200 seconds, so that part of the plot is truncated.}}\label{fig:cactus_cadical}\end{figure}
-
-\begin{figure}\centering\includegraphics[scale=0.65]{figures/cactus_maple_new.pdf}\caption{\small{Cactus plot for MapleLCMDist comparing standard non-chronological backtracking to \cbt and various configurations of trail saving.
-The first 300 problems were solved in less than 600 seconds, so that part of the plot is truncated.}}\label{fig:cactus_maple}\end{figure}
 

### 2020-03-02 figures/cactus_maple_new.tex
used maple data more consistent with definitions
a little cleaner

@@
 \addplot[blue,mark=+] table [x=index, y=original_time, col sep=comma] {results_maplelcmdist_trail_noforce_data.csv};
-\addplot[cyan,mark=square] table [x=index, y=original_time, col sep=comma] {results_maplelcmdist_trail_noforce_1lvl_data.csv};
-\addplot[gray,mark=triangle] table [x=index, y=original_time, col sep=comma] {results_maplelcmdist_trail_1lvl_5lbdreason_data.csv};
+\addplot[cyan,mark=square] table [x=index, y=original_time, col sep=comma] {results_maplelcmdist_trail_noforce_1lvl_again__data.csv};
+%\addplot[gray,mark=triangle] table [x=index, y=original_time, col sep=comma] {results_maplelcmdist_trail_1lvl_5lbdreason_data.csv};
+\addplot[gray,mark=triangle] table [x=index, y=original_time, col sep=comma] {results_maplelcmdist_trail_noforce_1lvl_again__5lbdreasons_data.csv};
 

### 2020-03-02 paper.tex
a little cleaner

@@
       maple-trail    & \textbf{472}          & 271 &  \textbf{201}  & 4618                \\ \hline
-      maple-trail-multipleBT-lookahead & 471 & 272 &  199  & \textbf{4597}        \\ \hline
-      maple-trail-multipleBT-lookahead-reason & 467 & \textbf{273} & 194 & 4668 \\ \hline
+      maple-trail-multipleBT-lookahead & 471 & \textbf{272} &  199  & \textbf{4597}        \\ \hline
+      maple-trail-multipleBT-lookahead-reason & 469 & 271 & 198 & 4633 \\ \hline
     \end{tabular}

### 2020-05-05 paper.tex
minor mods

@@
     often yields superior performance over chronological backtracking,
-    and that is is able to improve the performance of state-of-the-art
+    and that is able to improve the performance of state-of-the-art
     solvers.
@@
 which is equal to the number of decision literals appearing on the trail
-up to and including $\ell$; hence, $\dlevel{d}=1$ for the first
+up to and including $\ell$ ; hence, $\dlevel{d}=1$ for the first
 decision literal $d\in\trail$. The set of literals on $\trail$ that
@@
 \cite{DBLP:conf/sat/NadelR18} some more practical implementation
-methods were developed to overcome this issues, and in
+methods were developed to overcome these issues, and in
 \cite{DBLP:conf/sat/MohleB19} a formal framework for \cbt was
@@
 solver on the 800 total benchmark instances used in the main tracks of
-the 2018 and 2019 SAT competitions. The experiments were executed on a
+the 2018 SAT Competition and 2019 SAT Race. The experiments were executed on a
 cluster of 2.7 GHz Intel cores with 5000 seconds CPU time and 7 GB

### 2020-05-09 paper.tex
reviewer 1 changes

@@
     same decisions and propagations when it redescends the search
-    tree. Motivated by this potential redundancy, partial backtracking
-    (chronological backtracking) has been proposed. This technique
-    allows the solver to reduce the length of its backtrack, e.g.,
-    allowing the solver to backtrack to the previous level after a
-    conflict is found. However, this technique has its shortcomings
-    and only tends to work when limits are placed on its
-    application. In this paper we present a new trail saving technique
-    that makes a copy of the part of the trail that was backtracked
-    over. This saved copy can then be used to improve the efficiency
-    of the solver's subsequent redescent. Although our technique does
-    not save as much work as chronological backtracking, it also does
-    not have chronological backtracking's shortcomings. Furthermore,
-    the saved trail also provides the solver with the ability to look
-    ahead along the previous trail. This can be exploited to improve
-    its efficiency. Our empirical results show that our technique
-    often yields superior performance over chronological backtracking,
-    and that is able to improve the performance of state-of-the-art
-    solvers.
-
+    tree. Different techniques have been proposed to reduce this
+    potential redundancy, e.g., partial/chronological backtracking and
+    trail saving on restarts. In this paper we present a new trail
+    saving technique that is not restricted to restarts, unlike prior
+    trail saving methods. Our technique makes a copy of the part of
+    the trail that is backtracked over. This saved copy can then be
+    used to improve the efficiency of the solver's subsequent
+    redescent. Furthermore, the saved trail also provides the solver
+    with the ability to look ahead along the previous trail which can
+    be exploited to improve its efficiency. Our new trail saving
+    technique offers different tradeoffs in comparison with
+    chronological backtracking and often yields superior performance.
+    We also show that our technique is able to improve the performance
+    of state-of-the-art solvers.
 \end{abstract}
@@
 \section{Introduction}
-\textbf{Overall message of the paper.}
-
-% \begin{enumerate}
-% \item Formalism for understanding a more general form of trail savings
-%     which caches a copy of the trail that is backtracked over, and
-%     verifying how it exploited soundly.
-
-% \item Potential advantages over chronological backtracking.  Literals
-%     remain at the correct level, supports the solver having more
-%     freedom in its decisions.
-
-% \item Potential advantages of technique (1) increase in efficiency of
-%     propagation, (2) the new uses of the cached information.
-
-% \item Potential disadvantage. Reasons are recycled, in some circumstances
-%     new reasons might be better. We show in a simple experiment that
-%     this effect can occur.
-
-% \item Technique captures most of the positive effects of
-%     non-chronological backtracking in a simpler to implement way. It
-%     also preserves the simpler standard structure of CDCL
-%     solvers. Already provides some advantages for current SAT
-%     solvers. The ideas also have the potential for further
-%     exploitation. Lookahead for conflicts seems to be beneficial, and
-%     potentially if better discrimatory measures could be developed for
-%     reason clauses we could also better determine whether or not to
-%     use the cached information.. 
-% \end{enumerate}
-
-% \note{Reusing the Assignment Trail in CDCL Solvers: restart to level
-%   $i$ where the next decision has a higher score than the decision at
-%   level $i$ and a lower score than all decisions at level $j$ for
-%   $j< i$.
-
-%   They argum changes to reason clauses don't make much difference, but
-%   that was with minisat, before metrics for clause quality were in place in sat solvers.
-
-%   The savings for Luby-100 (used in minisat) were about 1\%}
-
 The vast majority of modern SAT solvers that are used to solve
@@
 every conflict, where all literals from one or more decision levels
-become unassigned, before the solver resumes making decisions and
+become unassigned before the solver resumes making decisions and
 performing unit propagations. Traditionally, CDCL solvers would
-backtrack non-chronologically to the conflict level, which is the
-second highest decision level remaining in the conflict clause after
-conflict analysis has resolved away all but one literal from the
-current decision level \cite{DBLP:conf/dac/MoskewiczMZZM01}. Recently,
-however, it has been shown that partial backtracking
-\cite{DBLP:conf/lpar/JiangZ13} or chronological backtracking, \cbt,
-(i.e. backtracking one level only after conflict analysis)
+backtrack to the conflict level, which is the second highest decision
+level remaining in the conflict clause after conflict analysis has
+resolved away all but one literal from the current decision level
+\cite{DBLP:conf/dac/MoskewiczMZZM01}. Recently, however, it has been
+shown that partial backtracking \cite{DBLP:conf/lpar/JiangZ13} or
+chronological backtracking, \cbt, (i.e., backtracking only to the
+previous level after conflict analysis)
 \cite{DBLP:conf/sat/NadelR18,DBLP:conf/sat/MohleB19} can be effective
-on many instances and was implemented in solvers that won the last two
-SAT competitions. Although chronological backtracking breaks some of
-the conventional invariants of CDCL solvers, it has been formalized
-and proven correct \cite{DBLP:conf/sat/MohleB19}.
+on many instances. Partial backtracking has been used in the solvers
+that won the last two SAT competitions. Although chronological
+backtracking breaks some of the conventional invariants of CDCL
+solvers, it has been formalized and proven correct
+\cite{DBLP:conf/sat/MohleB19} (also see related formalizations 
+\cite{DBLP:journals/jacm/NieuwenhuisOT06,nadelPhD}).
 
@@
 backtracks across many levels, many of the literals that are
-unassigned during the backtrack will be re-assigned again in roughly
+unassigned during the backtrack might be re-assigned again in roughly
 the same order when the solver redescends. This observation was first
@@
 
-In this paper we present an alternative ``trail saving'' method whereby
-we save the backtracked part of the solver's trail and attempt to
-use that information to make the solver's redescent more
-efficient. Trail saving preserves the traditional invariants of the
-SAT solver and its basic version is very simple to implement. It
+In this paper we present an new trail saving method whereby we save
+the backtracked part of the solver's trail and attempt to use that
+information to make the solver's redescent more efficient. Unlike
+\cbt, our trail saving method preserves the traditional invariants of
+the SAT solver and its basic version is very simple to implement. It
 allows the search to retain complete control over the order of
@@
 enhancements to make the idea more effective, and demonstrate
-experimentally that it performs as well as and sometimes better than
-chronological backtracking. We also show that with the enhancements we
+experimentally that it performs as well as and often better than
+chronological backtracking. We also show that with our enhancements we
 are able to improve the performance of state-of-the-art solvers.
@@
 \paragraph{Chronological Backtracking.}
-Chronological bactracking (\cbt) is a technique originally proposed by
-Jiang and Zhang \cite{DBLP:conf/lpar/JiangZ13} under the name
-\textit{partial backtracking}. Its aim is to reduce the redundant work
-that might be done by the \sat solver on its redescent from the
-backtrack level $\btL$. The original paper proposed a technique for
-backtracking to any level $j$ in the range
-$\btL \leq j \leq \deepestLvl{-}1$ (where $\deepestLvl$ is the level
-the conflict was discovered). Nadel and Ryvchin
-\cite{DBLP:conf/sat/NadelR18} proposed to always backtrack to
-$\deepestLvl{-}1$ while M{\"{o}}hle and Biere
-\cite{DBLP:conf/sat/MohleB19} returned to the original proposal of
-flexibly backtracking to any level in the allowed range. Note that the
-new 1-UIP clause $\uipcls$ is made unit at every level in this
-range. So after backtracking to level $j$ the newly implied literal
-$\deepestLit$ can be added to the trail with
-$\reason{\deepestLit}=\uipcls$, and $\dlevel{\deepestLit}$ is set to
-$\btL$ (the second deepest level in $\uipcls$.
+Chronological backtracking (\cbt) and partial backtracking in the
+context of clause learning solvers are alternatives to standard
+backtracking which allow the solver to execute a shorter
+backtrack. That is, with these techniques the solver can avoid having
+to go all the way back to the second deepest level in the learnt
+clause, as in standard backtracking.
+
+Formalisms for partial backtracking in clause learning solvers have
+been presented in
+\cite{DBLP:journals/jacm/NieuwenhuisOT06,nadelPhD}. In
+\cite{DBLP:conf/lpar/JiangZ13} practical issues of implementation were
+addressed, and experiments shown with a CDCL solver using partial
+backtracking. In \cite{DBLP:conf/sat/NadelR18} improved and more
+efficient implementation techniques were developed which allowed \cbt
+to make improvements to state-of-the-art \sat solvers, and
+\cite{DBLP:conf/sat/MohleB19} presented additional implementation
+ideas and details along with correctness results for these methods.
+
+The aim of partial backtracking is to reduce the redundant work that
+might be done by the \sat solver on its redescent from the backtrack
+level $\btL$. The technique allows the solver to backtrack to any
+level $j$ in the range $\btL \leq j \leq \deepestLvl{-}1$ (where
+$\deepestLvl$ is the level the conflict was discovered). Nadel and
+Ryvchin \cite{DBLP:conf/sat/NadelR18} proposed to always backtrack
+chronologically to $\deepestLvl{-}1$ while M{\"{o}}hle and Biere
+\cite{DBLP:conf/sat/MohleB19} returned to the proposal of
+\cite{DBLP:conf/lpar/JiangZ13} of flexibly backtracking to any level
+in the allowed range. Note that the new learnt 1-UIP clause $\uipcls$
+is made unit at every level in this range. So after backtracking to
+level $j$ the newly implied literal $\deepestLit$ is added to the
+trail with $\reason{\deepestLit}=\uipcls$, and $\dlevel{\deepestLit}$
+is set to $\btL$ (the second deepest level in $\uipcls$).
 
@@
 consequences for the \sat solver's operation, all of which were
-described in the original paper \cite{DBLP:conf/lpar/JiangZ13}. In
-\cite{DBLP:conf/sat/NadelR18} some more practical implementation
-methods were developed to overcome these issues, and in
-\cite{DBLP:conf/sat/MohleB19} a formal framework for \cbt was
-developed under which the changes required to the \sat solver could be
-shown to be sound. We refer the reader to the cited papers for
-more details.
-
-\section{Potential Drawbacks of Chronological Backtracking}
+described in \cite{DBLP:conf/lpar/JiangZ13}. M{\"{o}}hle and Biere
+\cite{DBLP:conf/satt/MohleB19} showed that despite these consequences
+partial backtracking can be made to preserve the soundness of a CDCL
+solver.
+
+\section{Chronological Backtracking Effects on Search}
 In this paper we present a new technique that allows the \sat solver
@@
 
-The reduction in redundant work achieved by \cbt does not come for
-free. In particular, some of the work avoided by \cbt is not actually
-redundant work; it is work that can actually help the solver. The
-obvious evidence for this statement is that in both
-\cite{DBLP:conf/sat/NadelR18} and \cite{DBLP:conf/sat/MohleB19} it was
-found that it was not optimal to always apply \cbt. In
-\cite{DBLP:conf/sat/NadelR18} \cbt was applied only when the length of
-the standard backtrack, $\btL-\deepestLvl$ was greater than a given
-threshold $T$. In their experiments they found that $T=100$ was the
-best value, i.e., \cbt is done only on longer backtracks. In practice,
-this meant that \cbt was \emph{hardly ever done}; in our measurements
-with their solver only about 3\% of the solver backtracks were \cbt
-backtracks. In \cite{DBLP:conf/sat/MohleB19} the value $T=100$ was
-also applied. However, they introduced an additional technique to add
-some applications of \cbt when the length of the backtrack is less
-than $T$. This allowed \cite{DBLP:conf/sat/MohleB19} to utilize \cbt
-in about \%15 of the backtracks. Nevertheless, the fact \cbt has to be used
-in a controlled fashion, indicates that there can be benefits in the \sat
-solver employing standard backtracking; it cannot be the case that
-every redescent is only doing redundant work.
-
-What is this non-redundant work? With standard backtracking the
+Although \cbt is able to avoid a lot of redundant work it also has
+other effects on the \sat solver search. These effects are sometimes
+detrimental to the solver's performance and so it is not always
+beneficial to use \cbt. In fact, in both \cite{DBLP:conf/sat/NadelR18}
+and \cite{DBLP:conf/sat/MohleB19} it was found that fairly limited
+application of \cbt performed best. In \cite{DBLP:conf/sat/NadelR18}
+\cbt was applied only when the length of the standard backtrack,
+$\btL-\deepestLvl$ was greater than a given threshold $T$. In their
+experiments they found that $T=100$ was the best value, i.e., \cbt is
+done only on longer backtracks. In practice, this meant that \cbt was
+relatively infrequent; in our measurements with their solver only
+about 3\% of the solver backtracks were \cbt backtracks. In
+\cite{DBLP:conf/sat/MohleB19} the value $T=100$ was also
+applied. However, they introduced an additional technique to add some
+applications of \cbt when the length of the backtrack is less than
+$T$. This allowed \cite{DBLP:conf/sat/MohleB19} to utilize \cbt in
+about 15\% of the backtracks.
+
+Although it is difficult to know precisely why \cbt is not always
+beneficial, we can identify some different ways in which \cbt can
+affect the \sat solver's search. With standard backtracking the
 literal $\deepestLit$ is placed on the trail at the end of $\btL$ and
@@
 from the newly learnt clause. And third, literals might be unit
-implied with different reasons. These changes could have a major
-impact on the future learnt clauses, and thus on the solver's
-efficiency. Unfortunately, it seems difficult to predict when these
-impacts might be large.
+implied with different reasons. \cbt can change all of these things,
+each of which could have an impact on the future learnt clauses, and
+thus on the solver's overall efficiency.
 
@@
 is greater than $100$ there could still be a divergence between the
-variable decisions generated by standard backtracking and \cbt. An
+variable decisions generated in standard backtracking and \cbt. An
 argument is also given in \cite{DBLP:journals/jsat/TakRH11} that the
@@
 \cite{DBLP:conf/ijcai/AudemardS09}. Our empirical results indicate
-that once clause quality is accounted for changing the literal reasons
-can have a significant impact.
-
-The first impact is worth discussing since it was mentioned in the
-original \cbt paper \cite{DBLP:conf/lpar/JiangZ13} but not in the
-subsequent works. This is the issue of changing the decision levels of
-literals on the trail. \cbt computes the decision level of each
-implied literal based on the decision levels of its reasons, but it
-does not go backwards to change the decisions levels of literals
-earlier on the trail.
+that once clause quality is accounted, for changing the literal
+reasons can have a significant impact.
+
+The first impact is worth discussing since it was mentioned in
+\cite{DBLP:conf/lpar/JiangZ13} but not in the subsequent works. This
+is the issue of changing the decision levels of literals on the
+trail. \cbt computes the decision level of each implied literal based
+on the decision levels of the literals in its reason, but it does not
+go backwards to change the decisions levels of literals earlier on the
+trail.
 
@@
 
-In sum, although these impacts of \cbt might or might not be harmful
-to the \sat solver, they exist. Hence, perhaps the most unappealing
-aspect of \cbt is that it introduces additional interdependencies
-between components of the solver. For example, the second impact means
-that heuristics for choosing decision variables might work differently
-in the presence of \cbt; and the first impact means that different
-ways of computing clause quality using the decision levels in the
-clauses might also work differently in the presence of \cbt. So
-although there are bound to be cases where \cbt is superior, we would
-like to have a method that addresses the original motivation of
-reducing redundant work while not changing the operation of standard
-backtracking. We present our proposal for such a method next.
+In sum, although these impacts of \cbt on the \sat solver's search
+might or might not be harmful to the \sat solver, they do exist. In
+fact, there are two pieces of evidence that these impacts can
+sometimes be harmful. First, as mentioned above, previous work found
+that it is best to only apply \cbt on large backtracks where it has
+the potential to save the most work. If there were no harmful effects
+it would always be effective to apply \cbt. And second, in our
+empirical results below we show that our new trail saving technique,
+which always uses standard backtracking, can often outperform
+\cbt. Although our technique reduces the solver's work on redescent it
+does not completely eliminate it like \cbt does. Hence its superior
+performance can only occur if \cbt is sometimes harmful.
+
+It is possible to combine \cbt with our trail saving technique to
+reduce the amount of work required whenever the solver performs
+non-chronological backtracking. However, \cbt greatly reduces the
+potential savings that could be achieved by our method since most of
+its non-chronological backtracks are relatively short (less than
+threshold $T$ levels). In our preliminary experiments this combination
+did not seem promising.
+
+Nevertheless, there is good evidence that \cbt can improve \sat solver
+performance.\footnote{\cbt can also be extremely useful in contexts
+  where each descent can be very expensive, e.g., when doing theory
+  propagation in SMT solving, or component analysis in \#\sat
+  solving. In these cases, \cbt, by avoiding backtracking and
+  subsequent redescent, has considerable potential for improving
+  solver performance.} Hence, it should be that it is better to
+perform \cbt in some branches. Hence, an interesting direction for
+future work would be to develop better heuristics about when to use
+\cbt in a branch and when to use standard backtracking augmented by
+our trail saving method.
 
@@
 in \cite{DBLP:conf/sat/MohleB19} as the baseline, and in
-Figure~\ref{fig:maple} we used MapleLCMDist \cite{xiao2017maplelrb} as
+Figure~\ref{fig:maple} we used MapleLCMDist \cite{DBLP:conf/ijcai/LuoLXML17,xiao2017maplelrb} as
 the baseline. Each of the baselines were run with standard

### 2020-05-11 figures/scatter2.tex
final version

--- 2020-02-16 figures/test2.tex
@@
 \begin{tikzpicture}
-\begin{axis}[title=Trail Saving Running Time Per Instance,axis equal,ymin=0,xmin=0,
-width=11cm,height=11cm,xlabel=original time (s),ylabel=trail saving time (s),axis x line=bottom,axis y line=left,xlabel near ticks,ylabel near ticks]
+\begin{axis}[title=Baseline vs Trail Saving Running Times in Cadical,axis equal,ymin=0,xmin=0,
+width=11cm,height=11cm,xlabel=Baseline running time (sec),ylabel=Trail Saving running time (sec),axis x line=bottom,axis y line=left,xlabel near ticks,ylabel near ticks]
 \addplot[only marks,mark size=1pt,mark=*,mark options={fill=black}]
 	table[x=original,y=trail]
-{test2.txt};
+{scatter2.txt};
 \addplot [red,samples at={0,1}] {x};
 \draw [blue,dashed] (rel axis cs:0,0) -- (rel axis cs:1,1);
+
 \end{axis}

### 2020-05-11 paper.tex
final version

@@
 \institute{Department of Computer Science, University of Toronto, Canada\\
-  \email{rgh000@gmail.com, fbacchus@cs.toronto.edu}}
+  \email{rhickey@cs.toronto.edu, fbacchus@cs.toronto.edu}}
 
@@
 
-In this paper we present an new trail saving method whereby we save
+In this paper we present a new trail saving method whereby we save
 the backtracked part of the solver's trail and attempt to use that
@@
 described in \cite{DBLP:conf/lpar/JiangZ13}. M{\"{o}}hle and Biere
-\cite{DBLP:conf/satt/MohleB19} showed that despite these consequences
+\cite{DBLP:conf/sat/MohleB19} showed that despite these consequences
 partial backtracking can be made to preserve the soundness of a CDCL
@@
 \cite{DBLP:conf/ijcai/AudemardS09}. Our empirical results indicate
-that once clause quality is accounted, for changing the literal
+that once clause quality is accounted for, changing the literal
 reasons can have a significant impact.
@@
 
-% \note{
-%   \begin{enumerate}
-%   \item Use simple backtrack and save to motivate and state the
-%       invariant. 
-%   \item Introduce notation for concatenating two trails.
-%   \item Name and state the invariant in text (e.g. a
-%       defn)---concenating the cached trail with the current trail
-%       yields a trail that is always reason valid, but not necessarily
-%       non-contradictory, non-conflicting, nor propagation complete).
-%   \item Proof (simple) that the invariant holds for caching on
-%       backtrack.
-%   \item On backtrack at least the asserted literal and any new
-%       propagants at the backtrack level are added.
-%   \item Prove that the invariant holds (???do we need an invariant or
-%       simply a property of the augmented trail) for trails with extras
-%       inbeween.
-%       $T_{\mathit{old}} + T_{\mathit{newlyAdded}} +
-%       T_{\mathit{saved}}$ also satisfies the invariant
-%   \item re-acheving propagation completeness for the augmented trail
-%       (a) only add one level of the cached trail at a time, and do
-%       unit prop after (b) only add non-contradicted decision from the
-%       old trail. Also note that conflicts are valid and can be learnt
-%       from even if the augmented trail is not propagation complete. 
-%   \item Finally figure out a way to formalize the extra technique of
-%       storing on top after repeated backtracking. 
-%   \item Explain how freedom for the search is maintained by makeing
-%       decisions then consulting the trail.
-%   \item 
-%   \end{enumerate}
-% }
-
 \subsection{Correctness}
@@
 
-% \section{Related Work}
-% \subsection{Trail Saving for Assumption-Based SAT}
-% Trail saving was briefly introduced in the context of assumption-based
-% SAT \cite{DBLP:conf/sat/HickeyB19} in order to speed up the
-% propagation of the assumptions each time the solver backtracked past
-% the assumption levels. Note that the process is the same, but it only
-% works to restore the propagants and implicants of the assumptions and
-% only after a learnt unit clause. Becuase the number of propagants and
-% implicants of the assumptions tends to become dominated by the number
-% of assumptions in the applications it was tested on, the technique did
-% not yield a significant improvement. It would be interesting to
-% combine the extended version of trail saving outlined in this paper
-% with trail saving over the assumptions to see if that could produce
-% better results.
-
-% \subsection{Comparison to Chronological Backtracking}
-% As mentioned earlier, chronological backtracking also achieves some of
-% the theoretical speedups that trail saving achieves. One advantage of
-% chronological backtracking is that it does not require any caching of
-% the trail or accessing such a cache during propagation; it simply
-% leaves a vast portion of the trail intact rather than backtracking all
-% the way to the conflict level. Chronological backtracking will also
-% speed up propagations slightly more than trail saving, because it
-% leaves more literals assigned at once which leads to clauses being
-% detected as unit more quickly.
-
-% One disadvantage that chronological backtracking has compared to trail
-% saving is that it doesn't allow the search as much flexibility to jump
-% out of local neighbourhoods, as the common variable selection
-% heuristics tend to depend on
-% \cite{DBLP:conf/dac/MoskewiczMZZM01,DBLP:conf/sat/2015,DBLP:conf/sat/LiangGPC16}. Because
-% chronological backtracking leaves a large portion of the trail
-% assigned, it will behave similarly to how the solver would behave if
-% it was to force the decisions that do occur within this large portion
-% of literals, whereas trail saving does not force any decisions. As
-% variable selection heuristics become more refined, it is likely that
-% adhering to them will become more crucial to performance.
-
-% Another disadvantage is that, because propagation is not performed at
-% every decision level, literals that remain assigned during
-% chronological backtracking (but otherwise would have been unassigned)
-% do not have a chance to move their decision levels up. An example of
-% this is when the solver was supposed to backtrack over a literal that
-% would end up being unit propagated at level 0, but stays assigned at
-% its current decision level when chronological backtracking is
-% used. This could potentially effect the lbd score or length of the
-% learnt clause that gets generated at the next conflict. Additionally,
-% chronological backtracking breaks the invarant that all literals
-% appear on the trail in monotonically increasing order by decision
-% level, which makes its implementation slightly more complicated.
-
 \section{Experiments and Results}
@@
 to cease trail saving once a reason of "low quality" is reached. For
-more details on the enhancements, please see section 4.2.
+more details on the enhancements, please see section 4.2. The source
+code for our implementation of trail saving can be obtained from the
+web\footnote{https://github.com/rgh000/cadical-trail} or by downloading our submissions to the 2020 SAT Competition.
 
@@
 
-\begin{figure}[H]
+\begin{figure}
 \centering
@@
 
-In Figure~\ref{fig:props} we plotted the propagations per second for
-each instance with and without trail saving enabled in cadical (rounded
-to the nearest 0.1 million). Only
-the 506 instances solved by both solvers are plotted. With trail saving
-enabled, 213 instances had more propagations per second, 149 instances
-had less propagations per second, and 144 instances had about the same
-number of propagations per second. On the graph, points above the diagonal
-represent instances where trail saving had faster propagation. The
-graph illustrates that the majority of the time
-cadical with trail saving enabled propagated literals faster than
-cadical without trail saving enabled, as we predicted would be the
-case earlier in this paper.
-
-\begin{figure}\centering\includegraphics[scale=0.7]{figures/scatter.pdf}\caption{Comparing propagations per second with and without trail saving in Cadical.}\label{fig:props}\end{figure}
-
 \section{Conclusion}
@@
 
-%\section{Future Work}
-
-%With more clever data structures the efficiency of restoring the trail
-%could be improved. There are a number of parameters which can be
-%tuned, including how often to refresh the trail, how large the cutoff
-%point for rejecting "long" reasons should be, how long to wait to
-%prune the trail cache if one is appending multiple trail caches
-%together, how far to look ahead on the trail cache for conflicts (in
-%terms of number of decisions or total number of literals). We only
-%experimented with a few different settings of these parameters that
-%made sense, but more careful fine-tuning of these parameters could
-%further enhance the effectiveness of the technique. There are
-%potentially other techniques one could use a trail cache for,
-%including inprocessing steps that involve probing. A look down the
-%trail cache is an incomplete but very fast way to do a probing step,
-%as opposed to propagation. It is also possible that trail saving and
-%chronological backtracking could be combined in some way to produce
-%even better results.
-
 \bibliography{sat}{}
@@
 \end{document}
-
-\iffalse \textit{Example 1} Suppose the literal $y$ can be restored
-from the trail cache, and its reason clause $cr$ contains the literals
-${y \lnot x_1, \lnot x_2, ..., \lnot x_n}$. This means that
-$x1, ..., x_n$ have become true and are currently on the
-trail. Suppose $x_n$ is the most recent literal on the trail and
-Without trail saving, the propagation routine will have to access $cr$
-from the watchlist of one of the literals $\blacksquare$\newline
-
-\textit{Example 1} Suppose the literals $x_1, ... x_n$ are on the
-trail cache, with none of them having a null reason, and $x_1$ becomes
-true (e.g., the solver makes a decision for $x_1$ to be true). Then,
-$x_2, ... x_n$ are implied by the current trail. Now we will examine
-the propagation of a clause c which consists of the literals
-$l, \lnot x_1, \lnot x_2, ..., \lnot x_n$ with and without trail
-saving. Assume $l$ and $\lnot x_1$ are the current watched literals.
-
-Without trail saving, $\lnot x_1$ becomes false and a new watcher is
-searched for and found 1 literal later, $\lnot x_2$, and then
-$\lnot x_1$ and $\lnot x_2$ are swapped. Next $\lnot x_2$ becomes
-false, a new watcher is searched for and found 2 literals later,
-$\lnot x_3$, and then $\lnot x_2$ and $\lnot x_3$ are swapped. This
-process will continue until all $x_n$ literals are false, c becomes
-unit, and $l$ becomes forced to be true. This propagation took
-$1 + 2 + ... + n$ searches through clause c, for a total of $O(n^2)$
-searches.
-
-With trail saving, $x2, ... x_n$ are all assigned to be true as soon
-as $x_1$ is made true, so immediately the values of literals
-$\lnot x_1, ..., \lnot x_n$ in c are false. During propagation, a new
-watcher is searched for to replace $\lnot x_2$, n literals are
-searched over without finding any, thus c is detected as unit and $l$
-becomes forced to be true. This propagation took $O(n)$ searches
-through clause c. $\blacksquare$\newline \fi
-
-\iffalse
-\begin{figure}\includegraphics[scale=0.8]{figures/cactus_cadical.pdf}\caption{\small{Comparison
-        of run times for versions of cadical. cadical is without
-        chronological backtracking, cadical-chrono is with
-        chronological backtracking, cadical-trail is with the first
-        version of trail saving, cadical-trail-2levels-ahead is the
-        trail saving with all enhancements added.}}\end{figure}
-\begin{figure}\includegraphics[scale=0.8]{figures/cactus_maple2.pdf}\caption{\small{Comparison
-        of run times for versions of MapleSAT. maplelcmdist is from
-        the 2017 SAT Competition, maplelecmdist-chrono is with
-        chronological backtracking (2018 SAT Competition),
-        maplelcmdist-trail is with the first version of trail saving,
-        maplelcmdist-trail-enhanced is with trail saving and all
-        enhancements added.}}\end{figure}
-\begin{figure}\includegraphics[scale=0.8]{figures/test.pdf}\caption{}\end{figure}
-\begin{figure}\includegraphics[scale=0.8]{figures/test2.pdf}\caption{}\end{figure}
-\fi
-\iffalse
-\begin{figure}\includegraphics[scale=0.8]{figures/cactus_maple.pdf}\caption{\small{Comparison
-        of run times for versions of MapleSAT. maplelcmdist is from
-        the 2017 SAT Competition, maplelecmdist-chrono is with
-        chronological backtracking (2018 SAT Competition),
-        maplelcmdist-trail is with the first version of trail saving,
-        maplelcmdist-chrono-dl is with chronological backtracking and
-        duplicate learnts (2019 SAT Competition),
-        maplelcmdist-trail-dl is with trail saving and duplicate
-        learnts}}\end{figure}
-\fi
-
-\iffalse
-\begin{figure}[t]
-    \begin{center}
-        \fss{9pt}{10pt}
-        \setlength\tabcolsep{1pt}
-        \begin{tabular}{| p{4.2cm} | c | c | c | c | c | c | c | c |}
-          \hline
-          & \multicolumn{4}{|c|}{Total} & \multicolumn{2}{|c|}{Unweighted} & \multicolumn{2}{|c|}{Weighted} \\ \hline
-          & MaxHS & +/- & \hspace{1pt} RC2 \hspace{1pt} & +/- & MaxHS & \hspace{1pt} RC2 \hspace{1pt} & MaxHS & \hspace{1pt} RC2 \hspace{1pt} \\ \hline
-          original & 6052 & 0/0 & 6030 & 0/0 & 3940 & 3993 & 2112 & 2037 \\ \hline
-          enqueue assumptions as set & 6131 & 114/35 & 6081 & \hspace{0.5pt} 99/48 \hspace{0.5pt} & 3995 & 4032 & 2136 & 2049 \\ \hline
-          enqueue assumptions as set + \newline save literals after learnt units & 6136 & 120/36 & 6079 & \hspace{0.5pt} 95/46 \hspace{0.5pt} & 3991 & 4030 & 2145 & 2049 \\ \hline
-          enqueue assumptions as set + \newline save literals after learnt units + \newline save literals from last invocation & 6138 & 115/29 & 6080 & \hspace{0.5pt} 92/42 \hspace{0.5pt} & 3989 & 4030 & 2149 & 2050 \\ \hline
-        \end{tabular}
-    \end{center}
-    \caption{Number of \maxsat instances solved by MaxHS and RC2 using
-      different extensions of the underlying \sat solver. The +/- column shows
-      the number of instances gained/lost vs the original.}
-    \label{fig:nsolved}
-\end{figure}
-\fi

### 2020-05-15 paper.tex
removed times font and cactus plot

@@
 \documentclass[runningheads]{llncs}
-\usepackage{times}
 \usepackage{amsmath}
@@
 We implemented our techniques in two different SAT solvers, MapleSAT
-and Cadical, both of which have finished at or near the top of SAT
+and Cadical, \footnote{Our implementation is available at https://github.com/rgh000/cadical-trail.}
+both of which have finished at or near the top of SAT
 competitions for the past several years
@@
 \ref{fig:maple}. We also show the cactus plots of the new version of
-cadical in Figure \ref{fig:cactus_cadical} and MapleLCMDist in Figure
-\ref{fig:cactus_maple}.
+cadical in Figure \ref{fig:cactus_cadical}.
+%and MapleLCMDist in Figure \ref{fig:cactus_maple}.
 
@@
 to cease trail saving once a reason of "low quality" is reached. For
-more details on the enhancements, please see section 4.2. The source
-code for our implementation of trail saving can be obtained from the
-web\footnote{https://github.com/rgh000/cadical-trail} or by downloading our submissions to the 2020 SAT Competition.
+more details on the enhancements, please see section 4.2.
 
@@
 
-\begin{figure}\centering\includegraphics[scale=0.65]{figures/cactus_maple_new.pdf}\caption{\small{Cactus plot for MapleLCMDist comparing standard non-chronological backtracking to \cbt and various configurations of trail saving.
-The first 300 problems were solved in less than 600 seconds, so that part of the plot is truncated.}}\label{fig:cactus_maple}\end{figure}
+%\begin{figure}\centering\includegraphics[scale=0.65]{figures/cactus_maple_new.pdf}\caption{\small{Cactus plot for MapleLCMDist %comparing standard non-chronological backtracking to \cbt and various configurations of trail saving.
+The first 300 problems were %solved in less than 600 seconds, so that part of the plot is truncated.}}\label{fig:cactus_maple}\end{figure}
 
@@
 state-of-the-art \sat solvers, cadical and MapleSAT, as or more
-effectively than chronological backtracking can without breaking the
-conventional invariants of CDCL. We also introduced three enhancements
+effectively than chronological backtracking can. We also introduced three enhancements
 one can implement when using a saved trail and demonstrated
@@
 conflict. It is also possible to combine trail saving with
-chronological backtracking if one is willing to break the traditional
-invariants of CDCL, but it would require further work to determine
+chronological backtracking, but it would require further work to determine
 whether or not this would be useful and how to best approach it.
